---
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
# Load packages and project local libraries
options(
  box.path = "code/",           # Path to project local libraries
  box.warn.legacy = FALSE       # No warnings for source inside a module
)
box::use(
  projlib/msdial,     # Handle MS-Dial files and data
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[rowData, `rowData<-`, colData, `colData<-`, assays, `assays<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble 
)
# Read `params.yml` if `params` not provided
if (! exists("params")) params <- yaml::read_yaml("params.yml")
overall_sumexp <- msdial$read_parsed_msdial_data(params)
set.seed(7579)

# Extract standard chemicals
is_targeted <- rowData(overall_sumexp)$std_type != ""
targeted_chems_se <- overall_sumexp[is_targeted, ]
```

# Given data

{{< include chapters/_read-msdial.qmd >}}

```{r}
#| purl: true
# Output file
stopifnot(    # Required parameters
  !is.null(params$input_file), 
  !is.null(params$table_dir)
)
FILE <- list(
  i = params$input_file,
  o = local({
    b <- basename(params$input_file) |> 
      tools::file_path_sans_ext()    # Without extension
    rlang::list2(
      conc = file.path(params$table_dir, paste0(b, ".conc.tsv")),
      norm = file.path(params$table_dir, paste0(b, ".norm.tsv")),
    )
  })
)
box::use(
  projlib/proc,         # Preprocessing functions
  patchwork,            # Combine ggplot objects. `+`
)
```

--------------------------------------------

# Normalization using internal standards

## Internal standard (IS) chemicals

```{r}
#| purl: true
internal_std_se0 <- internal_std_se <- msdial$get_internal_std_se(overall_sumexp) |> 
  proc$add_num_zeros() |> 
  proc$add_rsd()

# Number of outlying internal standard chemicals per sample 
n_out_istd <- proc$count_outliers_per_sample(internal_std_se, assay = "raw", times = 3)
is_outlier <- n_out_istd > nrow(internal_std_se) * 0.2
if (any(is_outlier)) {
  # Exclude the outlying samples
  stopifnot(names(is_outlier) == colnames(internal_std_se))
  internal_std_se <- internal_std_se[, !is_outlier] |> 
    proc$add_num_zeros() |> 
    proc$add_rsd()
  stopifnot(names(is_outlier) == colnames(overall_sumexp))
  overall_sumexp <- overall_sumexp[, !is_outlier]
}

# Failed internal standards
is_failed_istd <- rowData(internal_std_se)$num_zeros > 0
failed_istd <- internal_std_se[is_failed_istd, ]
# Remove failed internal standard chemicals
internal_std_se <- internal_std_se[!is_failed_istd, ]
```

{{< include chapters/_preprocess-internal-stds.qmd >}}

## Normalization

The data is normalized using the internal standard chemicals.

```{r}
#| purl: true
# Find the range of RT values
overall_rt_range <- range(rowData(overall_sumexp)$rt)
targetted_rt_range <- range(rowData(targeted_chems_se)$rt)
rt <- rowData(internal_std_se)$rt
```

The range of RT values of overall chemicals is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted chemicals is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and 
the range of internal standard chemicals is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard chemicals

```{r}
#| label: 'normalize-rt-closest-istd'
#| purl: true

closest_istd_raw <- proc$get_raw_of_closest_istd(overall_sumexp, internal_std_se)
# Normalize the data using closest internal standard chemicals
assays(overall_sumexp)$closest_norm <- assays(overall_sumexp)$raw / closest_istd_raw
```

```{r}
# Plot the RSD% of various data sets of chemicals
ggplot_rsdp_metab <- function(rsd_df, assay_ids) {
  nms <- names(assay_ids)
  ids <- unname(assay_ids)    # Duplicated by `dplyr::all_of`
  rsd_df <- rsd_df |> 
    dplyr::mutate(
      # NA was generated when all values were zero
      dplyr::across(dplyr::all_of(ids), \(.x) ifelse(is.na(.x), 0, .x)),
    ) |> 
    tidyr::pivot_longer(dplyr::all_of(ids), names_to = "Data", values_to = "RSD%") |> 
    dplyr::mutate(Data = factor(Data, levels = ids, labels = nms))
  ggplot(rsd_df) +
    aes(x = Data, y = `RSD%`) +
    geom_line(aes(group = chem_id), alpha = 0.5)
}
```

{{< include chapters/_preprocess-norm-closest.qmd >}}

### LOESS fit along RT

```{r}
#| label: 'normalize-rt'
#| purl: true

excluded_category_for_normalization <- c("Blank", "CalCurve")

loess_fit <- proc$get_loess_fit(
  istd_se = internal_std_se,
  excl_cat = excluded_category_for_normalization,
  overall_rt_range = overall_rt_range,
  span = 1
)
# Normalize the data by LOESS fit along RT
rt <- rowData(overall_sumexp)$rt
raw <- assays(overall_sumexp)$raw
assays(overall_sumexp)$loess_norm <- sapply(
  colnames(overall_sumexp), function(sample_id) {
    norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
    exp(log(raw[, sample_id]) - norm_factor)
  }
)
```

{{< include chapters/_preprocess-norm-loess.qmd >}}

```{r}
#| label: 'write-loess-normalized-data'
#| purl: true

msdial$write_normalized_data_tsv(overall_sumexp, "loess_norm", FILE$i, FILE$o$norm)
```

# Calibration

## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The samples with a concentration of `0` are separated as the zero samples.
The data of the zeros is used to calculate lower-limits of quantification (LLOQ).

```{r}
#| label: 'calcurve-samples-extraction'
#| purl: true
# Extract quantification standard data
quant_se <- overall_sumexp[rowData(overall_sumexp)$std_type == "Quant", ]
# Calibration curve samples only
calcurve_se <- quant_se[, quant_se$proc_cat == "CalCurve"]
stopifnot("Calibration curve samples are required." = nrow(calcurve_se) > 0)
# Calibration zero samples
cal_0_se <- calcurve_se[, calcurve_se$c_conc == 0] 
stopifnot("Multiple `Cal_0` samples are required." = ncol(cal_0_se) > 1)
# Remove the 0 concentration samples
calcurve_se <- calcurve_se[, calcurve_se$c_conc != 0]
```

The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
table(calcurve_se$c_conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Concentration ranges

Quite a few concentrations of calibration curve samples of a compound may be too far from the range of the measured samples. 
Hence the concentrations were excluded for which the peak areas after normalization were more than 10 times higher than the **maximum** peak area of the samples excluding `CalCurve` and `QC` categories.

```{r}
#| label: 'calcurve-upper-limit-concentration-range'
#| purl: true

rowData(calcurve_se)$max_c_conc <- local({
  stopifnot(rownames(quant_se) == rownames(calcurve_se))
  se1 <- quant_se
  se1 <- se1[, se1$proc_cat %in% c("CalCurve", "QC") == FALSE]
  # Find maximum peak area of the samples of the classes for measurement
  max_loess_norm <- apply(assays(se1)$loess_norm, 1, max, na.rm = TRUE)
  
  cc <- calcurve_se
  # Find the maximum concentration of the samples that are not too far (10x) from the measured
  apply(assays(cc)$loess_norm <= 10 * max_loess_norm, 1, \(ea_row) {
    if (all(!ea_row)) return(NA_real_)
    max(cc$c_conc[ea_row])
  })
})
```

When the obtained peak area of low concentration of the calibration samples is too low, the concentrations are excluded.
The value is considered **too low** when the average of the values of one concentration is less than 10 times of the standard deviation from the mean of zero concentration samples. 

```{r}
#| label: 'calcurve-lower-limit-concentration-range'
#| purl: true

# Function: find the mean + `times` of standard deviation 
get_mean_plus_sd <- function(mat, times) {
  m <- apply(mat, 1, mean, na.rm = TRUE)
  s <- apply(mat, 1, sd, na.rm = TRUE)
  m + times * s
}
# Find the lower-limit of quantification using values of the zero concentration samples
rowData(calcurve_se)$lloq <- get_mean_plus_sd(assays(cal_0_se)$loess_norm, times = 10)
rowData(calcurve_se)$min_c_conc <- local({
  cc <- calcurve_se
  # Find the minimum concentration of the calibration samples
  apply(assays(cc)$loess_norm > rowData(cc)$lloq, 1, \(ea_row) {
    if (all(!ea_row)) return(NA_real_)
    min(cc$c_conc[ea_row])
  })
})

# Mark the chemicals that have the minimum concentration equal to or greater than the maximum concentration
rowData(calcurve_se)$to_exclude <- as.data.frame(rowData(calcurve_se)) |> 
  with(is.na(min_c_conc) | is.na(max_c_conc) | min_c_conc >= max_c_conc)
```

When the minimum concentration is equal to or greater than maximum concentration, the chemical is excluded from the calibration curve.
`r english::Words(sum(rowData(calcurve_se)$to_exclude))` chemicals are excluded from the calibration curve.

Remaining concentration range of each chemical is presented below.

```{r}
#| label: 'calcurve-conc-range-table'

rowData(calcurve_se) |> 
  tibble::as_tibble() |> 
  dplyr::mutate(
    min_c_conc = dplyr::case_when(
      is.na(min_c_conc) ~ paste0("All < LLOQ (", scales::scientific(lloq, 2), ")"), 
      is.na(max_c_conc) ~ "",
      TRUE ~ scales::number(min_c_conc)
    ),
    max_c_conc = ifelse(is.na(max_c_conc), "All 0s, no valid", scales::number(max_c_conc)),
    to_exclude = ifelse(to_exclude, "Excluded", "")
  ) |>
  dplyr::select(chem_name, min_c_conc, max_c_conc, to_exclude) |>
  knitr::kable(
    caption = "Concentration range of the calibration curve samples",
    col.names = c("Chemical",
                  "Minimum (ng/ml)",
                  "Maximum concentration (ng/ml)",
                  "Excluded?"), 
    align = c("l", "r", "r", "r"),
  ) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px")
```

```{r}
#| label: 'calcurve-exclude-data-by-concentration-range'
#| purl: true

# Exclude the marked chemicals
is_marked <- rowData(calcurve_se)$to_exclude
stopifnot(rownames(calcurve_se) == rownames(quant_se))
calcurve_se <- calcurve_se[is_marked == FALSE, ]
quant_se <- quant_se[is_marked == FALSE, ]
rowData(quant_se)$lloq <- rowData(calcurve_se)$lloq    # Copy the LLOQ values
# Exclude the values outside the concentration range
calcurve_se <- local({
  cc <- calcurve_se
  # Concentration values in a matrix, rows are chemicals and columns are concentrations
  conc_mat <- matrix(rep(cc$c_conc, each = nrow(cc)), nrow = nrow(cc))
  to_na <- conc_mat < rowData(cc)$min_c_conc | conc_mat > rowData(cc)$max_c_conc
  assays(cc)$loess_norm[to_na] <- NA_real_
  cc
})
```

## Calibration curve per chemical

The calibration curve of each of randomly selected 9 chemicals is presented below.

```{r}
#| purl: true

set.seed(7579)
# Randomly selected 9 chemicals to present in calibration curve plots
i_9chem_to_show <- local({
  ii <- c(1:nrow(calcurve_se))
  ii <- ii[rowData(calcurve_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve_to_show_df0 <- tibble::as_tibble(calcurve_se[i_9chem_to_show, ])
calcurve_to_show_df <- calcurve_to_show_df0 |> 
  tidyr::drop_na(loess_norm)        # Drop concentrations outside the range
```

The peak areas were normalized by the internal standards applying the LOESS normalization.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

### Linear scale

```{r}
#| label: 'calcurve-per-chemical-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
p <- ggplot(calcurve_to_show_df, aes(x = c_conc, y = loess_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
print(p)
```

### Log scale

```{r}
#| label: 'calcurve-per-chemical-log'
#| fig-width: 12
#| fig-height: 11
p + scale_x_log10() + scale_y_log10()
```

## Calibration of the chemicals

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.

```{r}
#| label: 'calibration-data'
#| purl: true

quadratic_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc + I(conc^2), weights = weights)
  beta <- coef(lmfit)
  a <- beta[["I(conc^2)"]]
  b <- beta[["conc"]]
  cc <- beta[["(Intercept)"]]
  out <- function(x) {
    det <- b ^ 2 - 4 * a * (cc - x)
    conc <- suppressWarnings((-b + sqrt(det)) / (2 * a))      # det < 0
    conc[conc < 0 | conc < 0] <- 0
    conc
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}
linear_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc, weights = weights)
  beta <- coef(lmfit)
  b1 <- beta[["conc"]]
  b0 <- beta[["(Intercept)"]]
  out <- function(x) {
    (x - b0) / b1
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}

build_calcurve_model <- function(conc, value) {
  if (all(is.na(value) | value == 0)) {
    return(list(
      "best_model" = function(x) x,
      "R2s" = NA,
      "best_model_name" = NA_character_
    ))
  }
  # Weight alternatives
  weights_alt <- rlang::list2(
    "1" = rep(1, length(conc)),
    "1/x" = 1 / conc,
    "1/x2" = 1 / (conc ^ 2),
  )
  # Linear and quadratic models
  lmodels <- lapply(weights_alt, \(w) linear_calcurve_model(conc, value, weights = w))
  names(lmodels) <- paste("linear", names(lmodels), sep = "-")
  qmodels <- lapply(weights_alt, \(w) quadratic_calcurve_model(conc, value, weights = w))
  names(qmodels) <- paste("quadratic", names(qmodels), sep = "-")
  models <- c(lmodels, qmodels)
  R2s <- sapply(models, \(f) attr(f, "R2"))
  list(
    "best_model" = models[[which.max(R2s)]],        # Best model by R2
    "R2s" = R2s,
    "best_model_name" = names(models)[which.max(R2s)]
  )
}

# Fit the calibration curve
to_excl <- rownames(calcurve_se)[rowData(calcurve_se)$to_exclude]
c_conc <- calcurve_se$c_conc
rt_norm_mat <- assays(calcurve_se)$loess_norm
calcurve_models <- lapply(setNames(nm = rownames(calcurve_se)), function(ii) {
  build_calcurve_model(c_conc, rt_norm_mat[ii, ])
})
```

The distribution of $R^2$ values by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'

R2s <- lapply(calcurve_models, \(m) m$R2s)
R2s <- R2s[!is.na(R2s)] |> 
  dplyr::bind_rows(.id = "chem_id") |> 
  tidyr::pivot_longer(cols = -chem_id, names_to = "Model", values_to = "R2")

ggplot(R2s) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(
    x = "Model",
    y = expression(R^2)
  )
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chem'
#| results: 'markup'
table(sapply(calcurve_models, \(m) m$best_model_name))
```

```{r}
#| label: 'concentration-computation'
#| purl: true
# Compute the concentration of the samples skipping the calibration curve samples
measurements_se <- quant_se[, ! quant_se$proc_cat %in% "CalCurve"]
colData(measurements_se)$c_conc <- NULL
assays(measurements_se)$conc <- rowData(measurements_se)$chem_id |> 
  sapply(function(chem_id) {
    v <- assays(measurements_se)$loess_norm[chem_id, ]
    r <- calcurve_models[[chem_id]]$best_model(v)
    # Replace the values below LLOQ with half of the minimum concentration
    lloq <- rowData(measurements_se)[chem_id, "lloq"]
    min_c_conc <- rowData(calcurve_se)[chem_id, "min_c_conc"]
    r[v < lloq] <- min_c_conc / 2
    return(r)
  }) |>
    t()          # Chemicals to rows
# Negative concentrations are adjusted to zero
assays(measurements_se)$conc[assays(measurements_se)$conc < 0] <- 0
# Maximum/minimum concentration after trimming out of each chemical
maxmin_conc <- rowData(calcurve_se)[c("max_c_conc", "min_c_conc")]
stopifnot(rownames(maxmin_conc) == rownames(calcurve_se))
rowData(measurements_se) <- cbind(rowData(measurements_se), maxmin_conc)
```

The concentration of the sample classes, `r knitr::combine_words(unique(measurements_se$Class), before = "\x60")`,
are computed using the calibration curve.
Any negative concentration is adjusted to zero.

```{r}
#| purl: true

# Table of the samples to show in the calibration curve
data_to_show_tbl <- tibble::as_tibble(measurements_se[i_9chem_to_show, ]) |> 
  tidyr::drop_na(conc)
# Table about the selected calibration curve models
models_to_show_tbl <- purrr::imap_dfr(calcurve_models, \(m, chem_id) {
  tibble::tibble(
    chem_id = chem_id,
    best_model_name = m$best_model_name,
    R2 = max(m$R2s),
    model_r2 = paste0(best_model_name, "\nR2=", round(R2, 3)),
    
    chem_name = rowData(calcurve_se)[chem_id, ]$chem_name,
    lloq = rowData(calcurve_se)[chem_id, ]$lloq,
  )
}) |> 
  dplyr::slice(i_9chem_to_show)

# Plot the calibration curve for each chemical with the samples in the calibration curve
ggplot_calcurve_samples <- function(data) {
  ggplot(data, aes(x = conc, y = loess_norm)) +
    geom_point(aes(x = c_conc), data = calcurve_to_show_df) +     # Remaining calibration samples
    geom_point(aes(color = Class)) +                    # Samples to measure
    geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +     # Density of points at the edges of the plot
    labs(
      x = "Concentration (ng/ml)",
      y = "Peak area (Normalized)"
    ) +  
    facet_wrap(~ chem_name, scales = "free", ncol = 3) +
    geom_rect(         # Shade the region below the LLOQ
      aes(xmin = 0, xmax = Inf, ymin = 0, ymax = lloq),
      data = models_to_show_tbl,
      fill = "grey10",
      alpha = 0.5,
      inherit.aes = FALSE
    )
}
```

### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  ggplot_calcurve_samples()
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  ggplot_calcurve_samples() +
  scale_x_log10() +
  scale_y_log10()
```

### All except `QC` and `Blank` samples

```{r}
classes_excl_qc_blank <- local({
  df <- colData(measurements_se)
  unique(df$Class[!df$proc_cat %in% c("QC", "Blank")])
})
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11
geom_best_model_name <- geom_text(         # Display the best model name
  aes(x = 0, y = Inf, label = model_r2),
  data = models_to_show_tbl,
  hjust = -0.2,
  vjust = 1,
  size = 4,
  color = "red3"
)

data_to_show_tbl |> 
  dplyr::filter(Class %in% classes_excl_qc_blank) |> 
  ggplot_calcurve_samples() +
  geom_best_model_name
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  dplyr::filter(Class %in% classes_excl_qc_blank) |> 
  ggplot_calcurve_samples() +
  geom_best_model_name +
  scale_x_log10() +
  scale_y_log10()
```

# Concentration

The concentration of the samples has been computed using the calibration curve in the [section above](#calibration).
The distribution of the concentration of the samples is shown below.

```{r}
#| label: 'concentration-distribution-by-sample'
#| fig-width: 12
#| fig-height: 10
ggplot(measurements_se) +
  geom_boxplot(aes(x = given_sample_id, y = conc, fill = Class)) +
  labs(
    title = "Concentration distribution by sample", 
    x = "Sample ID",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```

```{r}
#| label: 'concentration-distribution-by-chemical'
#| fig-width: 12
#| fig-height: 20
ggplot(measurements_se) +
  geom_boxplot(aes(x = chem_name, y = conc)) +
  labs(
    title = "Concentration distribution by chemical", 
    x = "Chemical",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() + 
  facet_grid(Class~.)
```

```{r}
#| label: 'write-concentration-table'
# Write the concentration table
write_concentration_tsv <- function(sumexp, file) {
  # Prepare the chemical table
  chem_columns <- SummarizedExperiment::rowData(sumexp) |> 
    tibble::as_tibble() |> 
    dplyr::select(
      `Alignment ID` = alignment_id,
      `Chemical name` = chem_name,
      `Average Mz` = mz,
      `Average Rt(min)` = rt,
    )
  # Prepare the concentration table
  conc_df <- SummarizedExperiment::assays(sumexp)$conc |> 
    round(3) |>     # Reduce the number of decimals
    tibble::as_tibble()
  # Original sample ID
  colnames(conc_df) <- colData(sumexp)$given_sample_id
  readr::write_tsv(dplyr::bind_cols(chem_columns, conc_df), file)
}
write_concentration_tsv(measurements_se, FILE$o$conc)
```


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

