---
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-import'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  SummarizedExperiment,          # Some methods were defined using S4 system, e.g. dim
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
  io = projlib/check_io_exist,
)

# Input files
FILE <- rlang::list2(
  in_xl = "data/raw_internal/Target_Exposomics_240827.xlsx",
)
# Check if the files exist
io$chq_all_files_exist(FILE)

set.seed(7579)
```

# Given data

## Sheet "Original data (MS-DIAL)"

```{r}
#| label: 'read-given-xlsx-and-save-sumexp'
#| purl: true
#| cache: true

# Parse the data into SummarizedExperiment
parse_to_sumexp <- function(data_fr_xlsx, i_row_sample, i_col_metab) {
  # Confirm the column names are as expected
  stopifnot(identical(
    colnames(data_fr_xlsx)[i_col_metab],
    c("Alignment ID", "Average Rt(min)", "Average Mz", "Metabolite name", "Comment", "Class")
  ))
  # Uxtract metabolite info columns
  metabs <- data_fr_xlsx[-i_row_sample, i_col_metab, drop = FALSE] |> 
    # `metab_id` = Syntactically valid ID
    dplyr::mutate(metab_id = make.names(`Metabolite name`, unique = TRUE), .before = 1L) |> 
    # Not used
    dplyr::select(-Class) |> 
    # SummarizedExperiment will `make.names` change anyhow
    dplyr::rename(
      alignment_id = "Alignment ID",
      metab_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      comment = "Comment",
    )
  
  # Extract sample info rows
  sample_info <- t(data_fr_xlsx[i_row_sample, -i_col_metab, drop = FALSE])
  stopifnot(i_row_sample == 1)
  colnames(sample_info) <- "given_sample_id"
  # The row names contain the sample class
  sample_info <- tibble::as_tibble(sample_info, rownames = "Class")
  sample_info <- sample_info |> 
    dplyr::mutate(sample_id = make.names(given_sample_id, unique = TRUE)) |>   # Syntactically valid
    dplyr::select(sample_id, Class, given_sample_id)
  
  # Values of the metabolites into a matrix
  msdial_mat <- data_fr_xlsx[-i_row_sample, -i_col_metab, drop = FALSE]
  stopifnot(identical(colnames(msdial_mat), sample_info$Class))
  colnames(msdial_mat) <- sample_info$sample_id     # Update the column names with sample IDs
  msdial_mat <- lapply(msdial_mat, as.numeric) |> 
    as.data.frame(row.names = metabs$metab_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(msdial = msdial_mat),
    colData = sample_info,
    rowData = metabs,
  )
}
# Read the data in the sheet "Original data (MS-DIAL)"
msdial_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "Original data (MS-DIAL)",
  .name_repair = "minimal",     # No change of column names
)
I_COL_ANALYTE <- 1:6
I_ROW_SAMPLE <- 1           # The first row has information about the samples

msdial_sumexp <- parse_to_sumexp(msdial_data, I_ROW_SAMPLE, I_COL_ANALYTE)
colnames(SummarizedExperiment::rowData(msdial_sumexp))[6] <- "std_type"    # `Comment`
```

The sheet "Original data (MS-DIAL)" contains `r nrow(msdial_data[-I_ROW_SAMPLE, ])` rows (or metabolites) and `r ncol(msdial_data)` columns.
Among the columns, the first `r length(I_COL_ANALYTE)` columns have the analytes' information (`r knitr::combine_words(colnames(msdial_data)[I_COL_ANALYTE], before = "\x60")`),
and the rest has the data of the samples.
The column names are the sample IDs including standards for calibration curves and quality control samples.

## Samples

There are `r dplyr::n_distinct(msdial_sumexp$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
msdial_sumexp$Class |> 
  table() |> 
  knitr::kable(col.names = c("Class", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Metabolites

The data contained `r nrow(msdial_sumexp)` metabolites.
The metabolites were classified into `r dplyr::n_distinct(SummarizedExperiment::rowData(msdial_sumexp)$std_type)` types.
The number of metabolites for each type is as follows:
```{r}
SummarizedExperiment::rowData(msdial_sumexp)$std_type |> 
  table() |> 
  knitr::kable(col.names = c("Standard Type", "Number of metabolites")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

### Overall samples and metabolites

#### Linear scale

```{r}
#| label: 'distribution-values-linear'

# Distribution of the values
ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values", x = "Values")
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-log'
ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values", x = "Values") +
  scale_x_log10()
```

### By sample class and overall metabolites

```{r}
#| label: 'distribution-values-by-class'
#| fig-width: 10
#| fig-height: 6
#| out.width: '100%'

ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values by sample class", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Class, scales = "free_y", ncol = 4)
```

### By metabolite and overall samples

Randomly selected 8 metabolites are shown in the following figure.

```{r}
#| label: 'distribution-values-by-metabolite'
#| fig-width: 10
#| fig-height: 6
#| out.width: '100%'

# Randomly select 8 metabolites, excluding internal standards
i_sel <- c(1:nrow(msdial_sumexp))
i_sel <- i_sel[SummarizedExperiment::rowData(msdial_sumexp)$std_type != "Istd"]
i_sel <- sample(i_sel, 8)
ggplot(msdial_sumexp[i_sel, ]) +
  geom_histogram(aes(x = msdial, fill = Class), bins = 30) +
  labs(title = "Distribution of the values by metabolite", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ metab_name, scales = "free_x", ncol = 4)
```


----------------------------

# Normalization using internal standards

## Internal standard (IS) metabolites

```{r}
#| label: 'internal-standards'
#| purl: true

# Extract the internal standard metabolites
i_int_std <- SummarizedExperiment::rowData(msdial_sumexp)$std_type == "Istd"
internal_std_se <- msdial_sumexp[i_int_std, ]
# Sort by average retention time
rt <- SummarizedExperiment::rowData(internal_std_se)$rt
internal_std_se <- internal_std_se[order(rt), ]
```

There are `r nrow(internal_std_se)` internal standard (IS) metabolites in the data.
Those metabolites are `r knitr::combine_words(SummarizedExperiment::rowData(internal_std_se)$metab_name, before = "\x60")`, in the order of average retention time.

### Distribution of the values

Randomly selected 8 internal standard (IS) metabolites are shown in the following figure.

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 8
#| out-width: '100%'

i_sel <- sample(1:nrow(internal_std_se), 8) |> 
  sort()   # Keep the order by m/z
internal_std_se[i_sel, ] |> 
  # Below invokes `tidySummarizedExperiment:::mutate.SummarizedExperiment`
  dplyr::mutate(metab_name = forcats::fct(metab_name)) |>  # order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = msdial, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard metabolites", 
    y = "Values"
  ) +
  theme(legend.position = "none") +   # No legend for fill
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free_y", ncol = 4)
```

### Outliers

Outliers were identified based on the values of internal standard metabolites.
Those samples with values greater than 3 standard deviations from the mean after log-transformation were considered as outliers.

```{r}
#| label: 'outliers-internal-standards'
#| purl: true

internal_std_mat <- SummarizedExperiment::assays(internal_std_se)$msdial
# Log-transform the data
istd_mat_log <- log1p(internal_std_mat)
m_log <- rowMeans(istd_mat_log)
sd_log <- apply(istd_mat_log, 1, sd)

# Outliers
outliers <- (istd_mat_log > mean(m_log) + 3 * sd_log) | (istd_mat_log < mean(m_log) - 3 * sd_log)
outliers <- colSums(outliers)
is_outlier <- outliers > 0
outliers <- outliers[is_outlier]
```

There are `r length(outliers)` outlying samples in the data.

```{r}
#| label: 'outliers-table'
#| results: 'markup'
print(outliers)
```

After exclusion of those outliers, the distribution of the randomly selected 8 internal standard (IS) metabolites are shown below.

```{r}
#| label: 'distribution-values-internal-standards-no-outliers'
#| fig-width: 13
#| fig-height: 8
#| out-width: '100%'

internal_std_se[i_sel, !is_outlier, drop = FALSE] |> 
  # Below invokes `tidySummarizedExperiment:::mutate.SummarizedExperiment`
  dplyr::mutate(metab_name = forcats::fct(metab_name)) |>  # order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = msdial, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard metabolites", 
    y = "Values"
  ) +
  theme(legend.position = "none") +   # No legend for fill
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free_y", ncol = 4)
```


## Normalization

The data will be normalized using the internal standard metabolites.

### LOESS fit along RT

```{r}
#| label: 'normalize-rt-range'
#| purl: true

# Find the range of RT values
rt0 <- SummarizedExperiment::rowData(msdial_sumexp)$rt
overall_rt_rg <- range(rt0)
rt <- SummarizedExperiment::rowData(internal_std_se)$rt
```

The range of RT values of overall metabolites is `r overall_rt_rg[1]` to `r overall_rt_rg[2]`,
while the range of internal standard metabolites is `r range(rt)[1]` to `r range(rt)[2]`.

```{r}
#| label: 'normalize-rt'
#| purl: true

# Normalize the data using the internal standards
# Mean of each internal standard metabolite for overall measurement samples
# , excluding the calibration curve and blank samples
to_excl <- internal_std_se$Class %in% c("Blank", "CalCurve")
# , excluding outliers
to_excl <- to_excl | is_outlier
istd_mean_log <- rowMeans(istd_mat_log[, !to_excl])
# Subtract the mean from the data
internal_std_mat_norm_log <- istd_mat_log - istd_mean_log
# Expand to smallest RT with the mean of smallest 
i_smallest_rts <- order(rt)[1]
avg_smallest_rts <- colMeans(internal_std_mat_norm_log[i_smallest_rts, , drop = FALSE])
# Expand to largest RT with the mean of largest 
i_largest_rts <- order(rt, decreasing = TRUE)[1]
avg_largest_rts <- colMeans(internal_std_mat_norm_log[i_largest_rts, , drop = FALSE])
# Add the expanded range
internal_std_mat_norm_log <- rbind(
  avg_smallest_rts, 
  avg_largest_rts,
  internal_std_mat_norm_log
)

# LOESS fit of the internal standard metabolites along RT
loess_fit <- purrr::map(1:ncol(internal_std_mat_norm_log), function(ii) {
  # Fit a loess curve
  y <- internal_std_mat_norm_log[, ii]
  stopifnot(identical(
    names(y), c("avg_smallest_rts", "avg_largest_rts", rownames(internal_std_se))
  ))
  loess(y ~ c(overall_rt_rg, rt), span = 0.75)
})
names(loess_fit) <- colnames(internal_std_mat_norm_log)

# Normalize the data by LOESS fit along RT
rt <- SummarizedExperiment::rowData(msdial_sumexp)$rt
msdial <- SummarizedExperiment::assays(msdial_sumexp)$msdial
SummarizedExperiment::assays(msdial_sumexp)$rt_norm <- 
  sapply(
    SummarizedExperiment::colData(msdial_sumexp)$sample_id,
    function(sample_id) {
      norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
      out <- expm1(log1p(msdial[, sample_id]) - norm_factor)
      out[out < 0] <- 0
      out
    }
  )

# Extract quantification standard data
i_qstd <- SummarizedExperiment::rowData(msdial_sumexp)$std_type == "Std"
qstd_sumexp <- msdial_sumexp[i_qstd, ]
```

For each metabolite, deviation of each sample from the the average of all samples was calculated after log-transformation.
The data of the samples are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) metabolites along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all metabolites. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all metabolites.

Randomly selected 3 samples are shown below.
The line in the graph represents the LOESS fit.

```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 6
#| out.width: '100%'

# Plot the normalized and original data for randomly selected 3 samples
i_sel <- sample(c(1:ncol(msdial_sumexp))[!to_excl], 3)
# To include LOESS smooth line
x <- seq(overall_rt_rg[1], overall_rt_rg[2], length.out = 100)
loess_df <- lapply(setNames(i_sel, msdial_sumexp$sample_id[i_sel]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(rt = x) |>
  tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")

# Log-transform the data
msdial_log <- log1p(SummarizedExperiment::assays(msdial_sumexp)$msdial)
# Overall mean of every metabolite
mean_log <- rowMeans(msdial_log[, !to_excl])
# Deviation from the mean
selected_se <- msdial_sumexp[, i_sel]
SummarizedExperiment::assays(selected_se) <- SummarizedExperiment::assays(selected_se) |> 
  lapply(function(x) sweep(log1p(x), 1, mean_log))

# Plot the normalized and original data
tibble::as_tibble(selected_se) |>
  dplyr::rename(Raw = msdial, Normalized = rt_norm) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |>
  ggplot(aes(x = rt, y = Value)) +
  geom_point(aes(color = Data, shape = std_type)) +
  geom_line(data = loess_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values + 1) - log(avg + 1)"
  ) +
  scale_shape_manual(values = c("Istd" = 15, "Std" = 21)) +
  facet_wrap(~ sample_id, scales = "free_y", ncol = 3)
```


# Calibration

## Calibration curve samples

The samples for the calibration curve are identified by the sample class, `CalCurve`.
The concentration is extracted from the sample ID, which is in the format `Cal(|Std)[0-9]+_`.
The concentration is `0` for the sample ID `Cal0_`. 
The concentration is converted to a numeric value. 
The [outliers](#outliers) by the internal standards are removed from the calibration curve.

```{r}
#| label: 'calcurve-samples'
#| purl: true

# Calibration curve samples only
calcurve_se <- qstd_sumexp[, qstd_sumexp$Class == "CalCurve"]
# Remove the outliers by internal standards from the calibration curve
calcurve_se <- calcurve_se[, ! calcurve_se$sample_id %in% names(outliers)]

# Extract concentration values from the given IDs
calcurve_se$conc <- local({
  given_sample_id <- calcurve_se$given_sample_id 
  conc <- stringr::str_extract(given_sample_id, "Cal(|Std)([0-9]+)_", group = 2)
  dplyr::case_when(
    stringr::str_detect(given_sample_id, "Cal0_") ~ "0",
    stringr::str_detect(conc, "^0[0-9]") ~ stringr::str_replace(conc, "^0", "0."),
    TRUE ~ conc
  ) |> 
    as.numeric()
})
```

### Sample IDs

```{r}
#| results: 'markup'
# Present the sample IDs for the calibration curve
print(calcurve_se$given_sample_id)
```

### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
calcurve_se$conc |> 
  table() |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration", "Number of samples")
  )
```

## Calibration curve per metabolite

### Linear scale

#### Before normalization

```{r}
#| label: 'calcurve-per-metabolite-linear-before-norm'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"
set.seed(7579)
i_sel <- c(1:nrow(calcurve_se))
i_sel <- i_sel[SummarizedExperiment::rowData(calcurve_se)$std_type != "Istd"]
i_sel <- sample(i_sel, 9)
calcurve_to_show_df <- tibble::as_tibble(calcurve_se[i_sel, ])

# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = msdial)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Raw)"
  ) +  
  facet_wrap(~ metab_name, scales = "free_y", ncol = 3)
```

#### After normalization

The peak areas were normalized by the internal standards.
A line is fitted to the data points per metabolite.
Here the weight in the linear regression is inversely proportional to the squared concentration.
The concentration has been adjusted by adding 1 to avoid division by zero.

```{r}
#| label: 'calcurve-per-metabolite-linear'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

# Fitted line layer
fit_layer <- geom_smooth(
  aes(weight = 1 / (conc + 1) ^ 2),
  data = calcurve_to_show_df,
  method = "lm",
  formula = y ~ x,
  se = FALSE,
  color = "cadetblue"
)
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = rt_norm)) +
  fit_layer +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free_y", ncol = 3)
```


### Log scale

After normalization

```{r}
#| label: 'calcurve-per-metabolite-log'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

ggplot(calcurve_to_show_df, aes(x = conc, y = rt_norm)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free_y", ncol = 3)

```

## Calibration of the metabolites

The calibration curve is fitted to the [normalized data](#normalization) using a linear model with the weights of $\frac{1}{(x + 1)^2}$, where $x$ is the concentration.
The one is added to the concentration to avoid division by zero.

### All measurement samples

```{r}
#| label: 'calibration-data'
#| purl: true

# Fit the calibration curve
calcurves <- tibble::as_tibble(calcurve_se) |> 
  dplyr::summarise(
    model = list(lm(rt_norm ~ conc, weights = 1 / (conc + 1) ^ 2)),
    .by = metab_id
  ) |> 
  # Extract the coefficients
  dplyr::mutate(coef = purrr::map(model, coef), .keep = "unused") |>
  dplyr::mutate(
    intercept = purrr::map_dbl(coef, 1),
    slope = purrr::map_dbl(coef, 2),
    .keep = "unused"
  )

# Compute the concentration of the samples
classes_to_incl <- unique(qstd_sumexp$Class) |> 
  setdiff(c("Blank", "CalCurve", "RTI"))      # Exclude the control samples
measurements_se <- qstd_sumexp[, qstd_sumexp$Class %in% classes_to_incl]
stopifnot(identical(rownames(measurements_se), calcurves$metab_id))
SummarizedExperiment::rowData(measurements_se) <- cbind(
  SummarizedExperiment::rowData(measurements_se),
  tibble::column_to_rownames(calcurves, "metab_id")    # Attach the calibration curve
)
measurements_se <- measurements_se |> 
  dplyr::mutate(
    conc = (rt_norm - intercept) / slope,
    conc = ifelse(conc < 0, 0, conc),
  )
```

The concentration of the sample classes, `r knitr::combine_words(classes_to_incl)`, are computed using the calibration curve.
Any negative concentration is set to zero.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

measure_tbl <- tibble::as_tibble(measurements_se[i_sel, ])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = rt_norm)) +
  geom_point(data = calcurve_to_show_df) +
  fit_layer +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = rt_norm)) +
  geom_point(data = calcurve_to_show_df) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

### `NormSample` and `Groundwater` samples

The data of the two classes are presented on the calibration curve.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

to_incl <- measurements_se$Class %in% c("NormSample", "Groundwater")
measure_tbl <- tibble::as_tibble(measurements_se[i_sel, to_incl])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = rt_norm)) +
  geom_point(data = calcurve_to_show_df) +
  fit_layer +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = rt_norm)) +
  geom_point(data = calcurve_to_show_df) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

# Concentration

The concentration of the samples has been computed using the calibration curve in the [section above](#calibration).
The distribution of the concentration of the samples is shown below.

```{r}
#| label: 'concentration-distribution-by-sample'
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"

# Plot the concentration distribution by sample
ggplot(measurements_se) +
  geom_boxplot(aes(x = given_sample_id, y = conc, fill = Class)) +
  labs(
    title = "Concentration distribution by class", 
    x = "Sample ID",
    y = "Concentration",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```

```{r}
#| label: 'concentration-distribution-by-metabolite'
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"

# Plot the concentration distribution by metabolite
ggplot(measurements_se) +
  geom_boxplot(aes(x = metab_name, y = conc)) +
  labs(
    title = "Concentration distribution by metabolite", 
    x = "Metabolite",
    y = "Concentration",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```



```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

