---
title: "Report on {{< var title >}}"
format:
  titlepage-pdf:
    titlepage: academic
    titlepage-logo: "../projlib/images/SciLifeLab_Logotype_Green_POS.pdf"
    logo-align: center
    logo-size: 10
    #mainfont: "../projlib/fonts/Lora/Lora-VariableFont_wght.ttf"
    echo: false 
    fig-align: center  # Assure line breaks between figures
    documentclass: report
    papersize: a4
    classoption: landscape
    geometry:
      - margin=20mm
      - heightrounded
---

```{r}
#| label: 'setup'
#| purl: true
# The preprocessed data used for calibration. Default is `loess_norm_blk`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm_blk"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
IS_NON_TARGET_MODE <- SumExp::metadata(raw_se)$is_non_target_mode
# For consistency throughout the report
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$contr_cat)

# Input files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])
calib_interm_data <- qc_steps[[paste0("calibration/", DATA_ID)]]

concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
# Sort the chemicals by name
concn_se <- concn_se[order(SumExp::row_df(concn_se)$feature_name), ]
```

# Project Information

* User: `r user_inputs$user`
* University: `r user_inputs$university`
* Project: `r user_inputs$project_title`

`r user_inputs$free_text`

{{< pagebreak >}}

# General overview

* Number of samples: `r ncol(concn_se)`
* Number of targeted chemicals: `r nrow(concn_se)`

{{< pagebreak >}}

## PCA

```{r}
#| label: 'pca'
#| fig-width: 8
#| fig-height: 4
#| out-width: '100%'

plot_pca <- function(se, mat_id) {
  mat <- t(se[[mat_id]])
  mat <- mat[, colSums(mat, na.rm = TRUE) > 0]       # Scale is not possible with all zeros
  mat <- mat[, apply(mat, 2, var, na.rm = TRUE) > 0]        # Remove constant rows
  pca_res <- stats::prcomp(mat, scale. = TRUE)
  
  se_df <- SumExp::col_df(se)      # `data` in ggplot2::autoplot
  # Plot
  ggplot2::autoplot(pca_res, data = se_df, colour = "Class")
}
norm_se <- qc_steps[["normalized - blank"]]
p1 <- norm_se[, quote(! contr_cat %in% c("QC", "CalCurve", "Blank"))] |> 
  plot_pca(DATA_ID) + ggtitle("Target + Non-target")
p2 <- plot_pca(concn_se[, quote(contr_cat != "QC")], DATA_ID) +
  ggtitle("Target")
p1 + p2 + patchwork::plot_layout(guides = "collect")
```

{{< pagebreak >}}

## Chemical summary

```{r}
#| label: 'chemicals-summary'

# Summary table
summary_tbl <- show$tbl_chemical_summary(concn_se) |> 
  dplyr::mutate(      # Tidy up the table
    chem_name = stringr::str_trunc(chem_name, 40),
    perc_detf = sprintf("%5.1f%%", perc_detf),
    model_r2 = round(model_r2, 3),
    dplyr::across(c(median, mean, min, max, lod, lloq), ~ round(.x, 2))
  ) |> 
  dplyr::select(
    "Chemical" = chem_name,
    "DF (%)" = perc_detf,
    "N of sampl." = n_det,
    "LOD" = lod,
    "LLOQ" = lloq,
    "Median" = median,
    "Average" = mean,
    "Min" = min,
    "Max" = max,
    "R2" = model_r2,
    "Model" = best_model,
    "N of points" = n_conc,
  )
summary_tbl |> 
  knitr::kable(caption = "Summary of Targets") |> 
  kableExtra::kable_styling(
    latex_options = "striped", 
    full_width = FALSE, 
    repeat_header_continued = TRUE,
  ) |> 
  kableExtra::footnote(
    general = "DF: Detection frequency (percentage of samples above LOD)",
    footnote_as_chunk = TRUE
  )
```

**Note**

The unit of concentration is `r user_inputs$concentration_unit`.

In order to fit a calibration curve to a meaningful range of concentration, the range is adjusted.
The lower-limit of the range is chosen following the process described below. 
Concentration points below the limit are excluded from the calibration curve.

1. Begin with the lowest concentration point (e.g. `Cal_0`) in the calibration sample set with a detected signal (or peak area) greater than zero.
2. Examine the signal corresponding to the next highest concentration point to determine if it is statistically greater than the lowest non-zero signal, using a t-test for significance.
3. If no significant difference is found, continue to test the next higher concentration point sequentially until a significant increase in signal is identified.
4. If the calibration samples at the concentration with significantly higher signal exhibit high variability (relative standard deviation, RSD, > 20%), the lower-limit is adjusted to the lowest higher concentration point that satisfies RSD $\le$ 20%.

As quite a few concentration points in the calibration curve samples of a chemical can be too far from the range observed in the samples for quantification (non-`CalCurve` and non-`QC` sample categories), the upper-limit of the calibration range is set. 
When there are sufficient number (≥ 5) of concentration points, that concentrations that produce signal after normalization exceeding ten times the **highest** signal of the quantification samples are excluded.

If the lower-limit is set so high that fewer than three concentration points remain, that chemical is excluded from the calibration curve, because the curve cannot be fitted properly.

The limit of detection (LOD) is first calculated from a signal perspective by multiplying the standard deviation of the signals at the lowest calibration concentration point that exhibits a non-zero response by three and then adding this product to the average signal of the concentration point. 
The LOD in terms of concentration is then derived using the chosen calibration curve model.
When the LOD is below zero, it is adjusted to zero.

In a similar manner, the lower limit of quantification (LLOQ) is calculated by employing a factor of 10 instead of three, with all other steps remaining consistent with the LOD determination.

After the computation of concentration, the following steps are applied to the data:

  * Any negative concentration is adjusted to zero.
  * Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
  * Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
  * Any chemicals with values below LLOQ in all samples is removed from the data.


{{< pagebreak >}}

# Invidual chemicals

{{< pagebreak >}}

```{r}
#| label: 'individual-chemicals'
#| fig-width: 12
#| fig-height: 6
#| out-width: '100%'

#' Text box to describe a calibration curve model
#' 
textbox_of_a_width <- function(txt, width) {
  ggplot() +
    ggtext::geom_textbox(
      aes(x = 0, y = 0.9, label = txt),
      box.color = "black",
      width = width,
      fill = NA,
      hjust = 0,
      vjust = 1,
      size = 4
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_void()
}

calcurve_se <- calib_interm_data[["calcurve_se within range"]]
for(i_chem in 1:nrow(concn_se)) {
  chem_name <- SumExp::row_df(concn_se)$feature_name[i_chem]
  cat("\n\n## ", chem_name, "\n")
  
  x_se <- concn_se[i_chem, quote(! contr_cat %in% c("QC", "Blank"))]
  chem_i <- rownames(concn_se)[i_chem]
  cc_se <- calcurve_se[rownames(calcurve_se) == chem_i, ]
  # Calibration curves
  x1_se <- x_se   # For the case, lod < 0 or lloq < 0
  c_concs <- SumExp::col_df(cc_se)$c_conc
  lowest_c <- min(c_concs[c_concs > 0])
  SumExp::row_df(x1_se) <- SumExp::row_df(x1_se) |> 
    dplyr::mutate(dplyr::across(c(lod, lloq), ~ ifelse(.x < 0, lowest_c / 10, .x)))
  p1 <- show$ggplot_calcurve_samples(x1_se, 
                                     cc_se, 
                                     mat_id = DATA_ID,
                                     COLORS_OF_CLASSES) + 
    labs(
      title = "Full range",
      y = "Peak area (Normalized)"
    )
  p2 <- p1 + 
    show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID) +
    labs(title = "Zoom-in (3 cal. pts.)")
  # Distirbution of the concentration by class
  x_df <- SumExp::as_tibble(x_se)
  p3 <- ggplot(x_df) +
    geom_boxplot(aes(x = Class, y = conc, color = Class, fill = Class)) +
    labs(
      x = show$label_if_has(x_df$Class, "Class"),
      y = show$label_if_has(x_df$conc, "Concentration"),
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +    # Vertical x-axis
    theme(legend.position = "none") +      # No legend
    scale_y_log10() +
    scale_color_manual(values = COLORS_OF_CLASSES) +
    scale_fill_manual(values = COLORS_OF_CLASSES)
  
  # Text in the box
  text1 <- local(envir = show$tbl_chemical_summary(x_se), {
    perc_detf <- sprintf("%.1f%%", perc_detf)
    paste0(
      "Best model: ", best_model, "  \n",
      "R2: ", round(model_r2, 3), "  \n",
      "N of points: ", n_conc, "  \n",
      "DF: ", perc_detf, " (", n_det, "/", n_samples, ")", "  \n",
      "LOD: ", sprintf("%.2f", lod), "  \n",
      "LLOQ: ", sprintf("%.2f", lloq), "  \n",
      ""         # For trailing comma
    ) |> 
      textbox_of_a_width(width = unit(7, "cm"))
  })
  
  txt_legend <- paste0(
    "The gray areas in the calibration curve represent the regions below LOD and LLOQ.  \n",
    "DF: Detection frequency (percentage of samples above LOD)  \n",
    "LOD: Limit of detection.\n",
    "Lowest concentration at which the signal is three times the standard deviation above the lowest non-zero calibration point, as determined by the calibration model.  \n",
    "LLOQ: Lower limit of quantification\n", 
    "Ten times the standard deviation  \n",
    "The details of LOD and LLOQ are described in the text of the chemical summary table.  \n",
    ""         # For trailing comma
  ) |>
    textbox_of_a_width(width = unit(16, "cm"))
  
  # Plot
  design <- "
    112233
    112233
    445555
  "
  print(p1 + p2 + p3 + text1 + txt_legend +
          patchwork::plot_layout(design = design, guides = "collect"))
}
```

