---
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-import'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")
box::use(
  dplyr[across, case_when, case_match, coalesce, c_across, if_else, na_if],
  tidyselect[all_of, any_of, contains, ends_with, everything, matches, one_of, starts_with, where],
  ggplot2[...],        # too long to write and usually specific enough
  io = projlib/check_io_exist,
)

# Input files
FILE <- rlang::list2(
  in_xl = "data/raw_internal/Target_Exposomics_240827.xlsx",
)
# Check if the files exist
io$chq_all_files_exist(FILE)
```

# Given data

## Sheet "Original data (MS-DIAL)"

```{r}
#| label: 'read-given-xlsx-and-save-sumexp'
#| purl: true
#| cache: true

# Parse the data into SummarizedExperiment
parse_to_sumexp <- function(data_fr_xlsx, i_row_sample, i_col_metab) {
  # Confirm the column names are as expected
  stopifnot(identical(
    colnames(data_fr_xlsx)[i_col_metab],
    c("Alignment ID", "Average Rt(min)", "Average Mz", "Metabolite name", "Comment", "Class")
  ))
  id_metab <- rlang::quo(`Metabolite name`)
  # Save metabs separately
  metabs <- data_fr_xlsx[-i_row_sample, i_col_metab, drop = FALSE] |> 
    # `id` = Syntactically valid ID
    dplyr::mutate(metab_id = make.names(!!id_metab, unique = TRUE), .before = 1L) |> 
    # SummarizedExperiment will change anyhow
    dplyr::rename_with(~ make.names(.x))
  
  # Extract sample info rows
  sample_info <- t(data_fr_xlsx[i_row_sample, -i_col_metab, drop = FALSE])
  stopifnot(i_row_sample == 1)
  colnames(sample_info) <- "given_id"
  # The row names contain the sample type
  sample_info <- tibble::as_tibble(sample_info, rownames = "type")
  sample_info <- sample_info |> 
    dplyr::mutate(sample_id = make.names(given_id, unique = TRUE)) |>   # Syntactically valid
    dplyr::select(sample_id, type, given_id)
  
  # Values of the metabolites into a matrix
  msdial_mat <- data_fr_xlsx[-i_row_sample, -i_col_metab, drop = FALSE]
  stopifnot(identical(colnames(msdial_mat), sample_info$type))
  colnames(msdial_mat) <- sample_info$sample_id     # Update the column names with sample IDs
  msdial_mat <- lapply(msdial_mat, as.numeric) |> 
    as.data.frame(row.names = metabs$metab_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(msdial = msdial_mat),
    colData = sample_info,
    rowData = metabs,
  )
}
# Read the data in the sheet "Original data (MS-DIAL)"
msdial_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "Original data (MS-DIAL)",
  .name_repair = "minimal",     # No change of column names
)
I_COL_ANALYTE <- 1:6
I_ROW_SAMPLE <- 1           # The first row has information about the samples

msdial_sumexp <- parse_to_sumexp(msdial_data, I_ROW_SAMPLE, I_COL_ANALYTE)
colnames(SummarizedExperiment::rowData(msdial_sumexp))[6] <- "std_type"    # `Comment`
```

The sheet "Original data (MS-DIAL)" contained `r nrow(msdial_data[-I_ROW_SAMPLE, ])` rows (or metabolites) and `r ncol(msdial_data)` columns.
Among the columns, the first `r length(I_COL_ANALYTE)` columns were the analytes' information (`r knitr::combine_words(colnames(msdial_data)[I_COL_ANALYTE], before = "\x60")`),
and the rest were the data of the samples.
The column names were the sample IDs including standards for calibration curves and quality control samples.

* Dimension of the data: `r dim(msdial_sumexp)`


## Sheets "QStd" and "IS"

```{r}
# Read the data in the sheet "QStd"
qstd_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "QStd",
  .name_repair = "minimal",     # No change of column names
)
# Confirm the file was in the correct format
stopifnot(identical(
  colnames(qstd_data),
  c("Compound name", "", "Rt", "mz", "Comment")
))
qstd_data <- qstd_data[-2]    # No data
colnames(qstd_data) <- c("Metabolite.name", "Rt", "Mz", "Comment_Qstd")
tmp <- dplyr::left_join(
  SummarizedExperiment::rowData(msdial_sumexp) |> as.data.frame(),
  qstd_data, 
  by = "Metabolite.name"
)
# Confirm "QStd" sheet contains duplicated data
stopifnot(all(is.na(tmp$Comment_Qstd) | paste0("Q", tmp$std_type) == tmp$Comment_Qstd))

# Read the data in the sheet "IS"
is_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "IS",
  .name_repair = "minimal",     # No change of column names
)
# Confirm the file was in the correct format
stopifnot(identical(
  colnames(is_data),
  c("Compound name", "Rt", "mz", "Comment")
))
colnames(is_data) <- c("Metabolite.name", "Rt", "Mz", "Comment_IS")
is_data$Comment_IS <- "Istd"    # The values in `msdial_data` are `Istd`.
tmp <- dplyr::left_join(
  SummarizedExperiment::rowData(msdial_sumexp) |> as.data.frame(),
  is_data, 
  by = "Metabolite.name"
)
# Confirm "IS" sheet contains duplicated data
stopifnot(all(is.na(tmp$Comment_IS) | tmp$std_type == tmp$Comment_IS))
```

The sheets "QStd" and "IS" contained identical data to the sheet "Original data (MS-DIAL)". 
So the sheets are not used in the following analysis.

## Samples

There are `r dplyr::n_distinct(msdial_sumexp$type)` types of samples in the data.
The number of samples for each type is as follows:
```{r}
msdial_sumexp$type |> 
  table() |> 
  knitr::kable(col.names = c("Type", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Metabolites

The data contained `r nrow(msdial_sumexp)` metabolites.
The metabolites were classified into `r dplyr::n_distinct(SummarizedExperiment::rowData(msdial_sumexp)$std_type)` types.
The number of metabolites for each type is as follows:
```{r}
SummarizedExperiment::rowData(msdial_sumexp)$std_type |> 
  table() |> 
  knitr::kable(col.names = c("Standard Type", "Number of metabolites")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

----------------------------

# Normalization using internal standards

## Internal standard metabolites

```{r}
#| label: 'internal-standards'
#| purl: true

# Extract the internal standard metabolites
i_int_std <- SummarizedExperiment::rowData(msdial_sumexp)$std_type == "Istd"
internal_std_se <- msdial_sumexp[i_int_std, ]

# Exclude the internal standard metabolites from the data
qstd_sumexp <- msdial_sumexp[!i_int_std, ]
stopifnot(all(SummarizedExperiment::rowData(qstd_sumexp)$std_type == "Std"))

# Sort by the average m/z
intstd_analytes <- SummarizedExperiment::rowData(internal_std_se)
internal_std_se <- internal_std_se[order(intstd_analytes$Average.Mz), ]


```

There are `r nrow(internal_std_se)` internal standard metabolites in the data.
Those metabolites are `r knitr::combine_words(SummarizedExperiment::rowData(internal_std_se)$Metabolite.name, before = "\x60")`.

## Normalization




# Calibration curves

```{r}
#| label: 'divide-samples-by-type'
#| purl: true

# Divide the data by the sample type
i_by_type <- split(1:ncol(msdial_sumexp), msdial_sumexp$type)
msdial_by_type <- purrr::map(i_by_type, ~msdial_sumexp[, .x])
```


## Calibration curve samples

```{r}
#| label: 'calcurve-samples'
#| purl: true


# Extract concentration values from the given IDs
msdial_by_type[["CalCurve"]]$conc <- local({
  given_id <- msdial_by_type[["CalCurve"]]$given_id 
  conc <- stringr::str_extract(given_id, "Cal(|Std)([0-9]+)_", group = 2)
  dplyr::case_when(
    stringr::str_detect(given_id, "Cal0_") ~ "0",
    stringr::str_detect(conc, "^0[0-9]") ~ stringr::str_replace(conc, "^0", "0."),
    TRUE ~ conc
  ) |> 
    as.numeric()
})
calcurve_se <- msdial_by_type[["CalCurve"]]
```

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
calcurve_se$conc |> 
  table() |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration", "Number of samples")
  )
```

## Calibration curve per metabolite

### Linear scale

```{r}
#| label: 'calcurve-per-metabolite-linear'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"
set.seed(7579)
i_sel <- sample(1:nrow(calcurve_se), 9)
calcurve_se <- calcurve_se[i_sel, ]

# Plot calibration curves
tidySummarizedExperiment:::ggplot.SummarizedExperiment(calcurve_se, aes(x = conc, y = msdial)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area"
  ) +  
  facet_wrap(~ Metabolite.name, scales = "free_y", ncol = 3)
```

### Log scale

```{r}
#| label: 'calcurve-per-metabolite-log'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

tidySummarizedExperiment:::ggplot.SummarizedExperiment(calcurve_se, aes(x = conc, y = msdial)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ Metabolite.name, scales = "free_y", ncol = 3)

```



<!-- ```{r} -->
<!-- #| results: 'asis'         # This produces multiple chunks. -->
<!-- #| echo: false             # - knitr::knit_child, asis_output -->
<!-- titles <- lst(           # Allow a trailing comma -->
<!--   test1 = "Group1 vs. Group2", -->
<!--   test2 = "Group2 vs. Group3", -->
<!-- ) -->

<!-- sub_groups <- c("A", "B") -->

<!-- sapply(names(titles), \(ii) {   # ii <- names(titles)[1]   # for debug -->
<!--   header <- rlang::chr( -->
<!--     '\n#### {{titles[[ii]]}}', -->
<!--   ) |>  -->
<!--     knitr::knit_expand(text = _) -->

<!--   body <- lapply(sub_groups, \(jj) { -->
<!--     p_layout <- ggplot_build(vo[[ii]][[jj]]$examples)$layout$layout -->
<!--     nr <- map(p_layout[c("ROW", "COL")], ~max(.x)) -->
<!--     out_width <- sprintf("%i%%", round((nr$COL + 0.75) / 5.75 * 100)) -->
<!--     rlang::chr( -->
<!--       '\n##### {{jj}} condition', -->
<!--       # Volcano plot -->
<!--       '```{r}', -->
<!--       "#| label: '{{ii}}-{{jj}}-volcano-plot'", -->
<!--       '#| fig-width: 6',  -->
<!--       '#| fig-height: 6', -->
<!--       "#| out-width: '60%'", -->
<!--       '#| fig-cap: ', -->
<!--       'volcano_kable_examples${{ii}}${{jj}}$volcano', -->
<!--       "```", -->
<!--       # Plots to show examples -->
<!--       "```{r}", -->
<!--       "#| label: '{{ii}}-{{jj}}-examples'", -->
<!--       '#| fig-width: {{nr$COL * 2 + 1.5}}', -->
<!--       '#| fig-height: {{nr$ROW * 2 + 1}}', -->
<!--       "#| out-width: '{{out_width}}'", -->
<!--       'volcano_kable_examples${{ii}}${{jj}}$examples', -->
<!--       "```", -->
<!--     ) |>  -->
<!--       knitr::knit_expand(text = _) -->
<!--   }) -->
<!--   c(header, body) -->
<!-- }) |>  -->
<!--   knitr::knit_child(text = _, quiet = TRUE) |>  -->
<!--   knitr::asis_output()    # print `asis` -->
<!-- ``` -->


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

