---
params:
  input_file: "data/raw_internal/MMLC_ESIpos_for_Mun-Gwan.txt"
title: "Read MS-DIAL output file"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[rowData, `rowData<-`, colData, `colData<-`, assays, `assays<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
  io = projlib/check_io_exist,
)
# Input file
FILE <- list(i = params$input_file)
# Check if input file exists
io$chq_all_files_exist(FILE$i)
set.seed(7579)
```

```{r}
#| label: 'read-given-tsv-and-save-sumexp'
#| purl: true

#' Identify three column sections in the MS-DIAL output file
#' 
#' @param msdial_file Path to the MS-DIAL output file
#' @return A list of three column indices
get_three_section_indices <- function(msdial_file) {
  # Read the second (`File type`) line to identify those sections
  file_type_line <- read.delim(
    msdial_file, header = FALSE, nrow = 5L, na.strings = c("NA", "")
  )[2, ]
  i_2nd <- which(! is.na(file_type_line))
  stopifnot(file_type_line[i_2nd[1L]] == "File type")    # As expected
  list(
    "1st" = 1L:i_2nd[1L],
    "2nd" = (i_2nd[1L] + 1L):max(i_2nd),
    "3rd" = (max(i_2nd) + 1L):length(file_type_line)
  )
}

#' Fetch sample information from the first 5 lines
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the columns that contain sample information
#' @return A tibble with sample information
fetch_sample_info <- function(msdial_file, indices) {
  if (missing(indices)) indices <- get_three_section_indices(msdial_file)[[2]]
  # Read sample information lines (first 5 lines)
  sample_lines <- read.delim(
    msdial_file, 
    header = FALSE,
    nrow = 5L,
    na.strings = c("NA", "")
  )
  header <- sample_lines[, indices[1L] - 1L]
  
  # Extract sample information  
  sinfo <- t(sample_lines[, indices, drop = FALSE])
  colnames(sinfo) <- header          # There was no header in `sinfo` before
  stopifnot("Deviation from the expected names for samples" = identical(
    colnames(sinfo)[-5], c("Class", "File type", "Injection order", "Batch ID")
  ))
  colnames(sinfo)[5] <- "given_sample_id"
  # Create a tibble with the sample information
  sinfo <- tibble::as_tibble(sinfo) |> 
    dplyr::select(
      Class,
      sample_type = "File type",
      injection_order = "Injection order",
      given_sample_id,
    ) |> 
    # Syntactically valid for variable name in R
    dplyr::mutate(sample_id = make.names(given_sample_id, unique = TRUE), .before = 1L) |>
    # Control sample groups
    dplyr::mutate(
      proc_category = dplyr::case_when(
        Class == "CalCurve"    ~ "CalCurve",
        sample_type == "QC"    ~ "QC",
        sample_type == "Blank" ~ "Blank",
        TRUE ~ ""         # NA doesn't work with `==`
      )
    )
  stopifnot(
    "Calibration curve samples are required." = any(sinfo$proc_category == "CalCurve"),
    "QC samples are required." = any(sinfo$proc_category == "QC"),
    "Blank samples are required." = any(sinfo$proc_category == "Blank")
  )
  
  # Function to get short sample ID that can be used to identify repeated samples
  get_short_id <- function(ids, cat) {
    ids0 <- ids
    ids <- stringr::str_remove(ids, "^[[:digit:]]+_")    # Remove optional numeric prefix
    ids <- dplyr::case_when(
      cat == "CalCurve" ~ stringr::str_extract(ids, "^Cal_[[:digit:]-]+"),
      cat == "QC"       ~ stringr::str_extract(ids, "^QC[[:alnum:]]*"),
      # cat == "Blank"    ~ stringr::str_extract(ids, "Blank[[:digit:]]*"),
      TRUE ~ ids
    )
    if (any(tmp <- is.na(ids) | ids == "")) {
      stop("ID error in CalCurve or QC.\n", paste(ids0[tmp], collapse = ", "), "\n")
    }
    ids
  }
  sinfo <- sinfo |> 
    dplyr::mutate(short_sample_id = get_short_id(given_sample_id, proc_category))
  return(sinfo)
}

#' Fetch data of the selected columns of the MS-DIAL output file
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the selected columns
#' @return A tibble 
fetch_data_of_columns <- function(msdial_file, indices) {
  stopifnot(!missing(indices))
  # Count total number of columns in the file
  l1 <- readr::read_tsv(msdial_file, skip = 4L, n_max = 1L, col_names = FALSE, col_types = "c")
  coltypes <- rep("-", ncol(l1))      # Skip the rest columns
  coltypes[indices] <- "c"
  # Read lines that contain chemical data
  readr::read_tsv(
    msdial_file, 
    skip = 4L,      # Data starts from the 5th line
    col_types = paste0(coltypes, collapse = ""),
    name_repair = "unique_quiet"
  )
}

# Parse the data in the file into SummarizedExperiment
parse_to_sumexp <- function(msdial_file) {
  i_three_sections <- get_three_section_indices(msdial_file)
  sample_info <- fetch_sample_info(msdial_file, i_three_sections[[2]])
  
  # Read chemical data in the first section
  chems <- fetch_data_of_columns(msdial_file, i_three_sections[[1]]) |>
    dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
      alignment_id = "Alignment ID",
      chem_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      sn_ratio = "S/N average",
      std_type = "Comment",            # Type of standard, e.g. "Quant", "IS", or NA 
    ) |> 
    # `chem_id` = Syntactically valid ID
    dplyr::mutate(chem_id = make.names(chem_name, unique = TRUE), .before = 1L) |> 
    dplyr::mutate(
      dplyr::across(c(mz, rt, sn_ratio), as.numeric),
      std_type = dplyr::case_when(
        std_type == "Quant" ~ "Quant",
        std_type == "IS"    ~ "IS",
        TRUE ~ ""              # Ignore other variants
      )
    )
  
  # Measured values of the chemicals into a matrix
  raw_df <- fetch_data_of_columns(msdial_file, i_three_sections[[2]])
  stopifnot(identical(colnames(raw_df), sample_info$given_sample_id))
  colnames(raw_df) <- sample_info$sample_id     # Update with syntactically valid names
  raw_mat <- lapply(raw_df, as.numeric) |> 
    as.data.frame(row.names = chems$chem_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(raw = raw_mat),
    colData = sample_info,
    rowData = chems,
  )
}
overall_sumexp <- parse_to_sumexp(FILE$i)  # To be updated throughout the pre-processing
```

Given file: `` `r FILE$i` `` 

The given file contains data of `r nrow(overall_sumexp)` chemicals and `r ncol(overall_sumexp)` samples.

## Samples

### By class

There are `r dplyr::n_distinct(overall_sumexp$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
table(overall_sumexp$Class) |> 
  knitr::kable(col.names = c("Class", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

### Special categories for preprocessing

There are `r length(table(overall_sumexp$proc_category, exclude = ""))` special sample categories for in the data.
The data of those were used in the following processing steps.
The number of samples for each category is as follows:

```{r}
table(overall_sumexp$proc_category, exclude = "") |> 
  knitr::kable(col.names = c("Category", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

#### Calibration curve samples (`CalCurve`)

```{r}
#| label: 'calcurve-samples'
#| purl: true
# Extract calibration curve samples
extract_calcurve_samples <- function(se) {
  se[, se$proc_category == "CalCurve"]
}
# Function to find concentration values from the given IDs
find_concentration <- function(sid) {
  sid |> 
    stringr::str_remove("Cal_") |> 
    stringr::str_replace("-", ".") |>     # Replace "-" with "."
    as.numeric()
}
```

The samples for the calibration curve are identified by the sample class, `CalCurve`.
The concentration is extracted from the sample ID, which is in the format `Cal_[[:digit:]-]+`, and converted to a numeric value. 

##### Sample IDs

```{r}
# Present the sample IDs for the calibration curve
df1 <- extract_calcurve_samples(overall_sumexp) |> 
  colData() |> 
  tibble::as_tibble() |> 
  dplyr::mutate(conc = find_concentration(short_sample_id)) |> 
  dplyr::arrange(conc)
df1 |> 
  dplyr::select(given_sample_id, conc) |>
  knitr::kable(
    col.names = c("Sample ID", "Concentration (ng/ml)"),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

##### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
table(df1$conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Quality control samples (`QC`)

```{r}
#| purl: true
# Extract QC samples
extract_qc_samples <- function(se) {
  se[, se$proc_category == "QC"]
}
# Extract QC samples into list
extract_qc_samples_to_list <- function(se) {
  se <- extract_qc_samples(se)
  qc_id <- se$short_sample_id
  # Column-wise split
  lapply(setNames(nm = unique(qc_id)), \(ii) se[, qc_id == ii])
}
```

Some of the analyzed samples are quality controls.
Aliquots of the samples were measured as technical replicates.
The number of the replicates were as below.

```{r}
extract_qc_samples_to_list(overall_sumexp) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Sample ID", "Number of replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Chemicals

The data contained `r nrow(overall_sumexp)` chemicals.
Among them, `r sum(rowData(overall_sumexp)$std_type != "")` chemicals were targeted. 
The rest were not targeted.

```{r}
#| purl: true
# Extract chemicals that are standards
is_targeted <- rowData(overall_sumexp)$std_type != ""
targeted_chems_se <- overall_sumexp[is_targeted, ]
```

For those targeted chemicals, the standard type was specified.
The number of chemicals for each type is as follows:

```{r}
table(rowData(targeted_chems_se)$std_type) |> 
  knitr::kable(col.names = c("Standard Type", "Number of chemicals")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

### Overall samples and targeted chemicals

#### Linear scale

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 8 
#| out-width: '60%'
p <- ggplot(targeted_chems_se) +
  geom_histogram(aes(x = raw), bins = 50) +
  labs(title = "Distribution of the values", x = "Values")
print(p)
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-overall-log'
#| fig-width: 8
#| out-width: '60%'
p + scale_x_log10()
```

### By sample class and overall targeted chemicals

```{r}
#| label: 'distribution-values-by-class'
ggplot(targeted_chems_se) +
  geom_histogram(aes(x = raw), bins = 30) +
  labs(title = "Distribution of the values by sample class", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Class, scales = "free_y", ncol = 4)
```

### By chemical and overall samples

Randomly selected 8 targeted chemicals are shown in the following figure.

```{r}
#| purl: true
#' Randomly select `n` rows from `sumexp` with optional `subset`
#'
#' @param sumexp a SummarizedExperiment object
#' @param n the number of rows to select
#' @param subset an optional subset expression to apply to the `rowData` of `sumexp`
#' @param keep_the_order if `TRUE`, the selected rows are sorted in the previous order
#'
#' @return a SummarizedExperiment object with `n` rows randomly selected from `sumexp`
#' @export
select_rows_randomly <- function(sumexp, n, subset, keep_the_order = FALSE) {
  ii <- c(1:nrow(sumexp))
  if (!missing(subset)) {      # If given, apply it to the `rowData` of `sumexp`
    ii <- ii[eval(substitute(subset), as.data.frame(rowData(sumexp)))]
  }
  ii <- sample(ii, n)
  if (keep_the_order) ii <- sort(ii)
  sumexp[ii, ]
}
```

```{r}
#| label: 'distribution-values-by-chemical'
se <- select_rows_randomly(targeted_chems_se, 8, std_type != "IS")
ggplot(se) +
  geom_histogram(aes(x = raw, fill = Class), bins = 30) +
  labs(title = "Distribution of the values by chemical", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ chem_name, scales = "free_x", ncol = 4)
```
