---
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
# The preprocessed data used for calibration
DATA_ID <- setNames("loess_norm", nm = "Peak area (Normalized)")

# Load packages and project local libraries
options(
  box.path = "code/",           # Path to project local libraries
  box.warn.legacy = FALSE       # No warnings for source inside a module
)
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Read `params.yml` if `params` not provided
if (! exists("params")) params <- yaml::read_yaml("params.yml")
overall_sumexp <- msdial$read_parsed_msdial_data(params)
# Conversion table for sample names
sample_names_tbl <- SumExp::col_df(overall_sumexp) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
# Conversion table for chemical names
chem_names_tbl <- SumExp::row_df(overall_sumexp) |> 
  tibble::rownames_to_column("chem_id") |> 
  dplyr::select(chem_id, chem_name)
set.seed(7579)

# Extract standard chemicals
is_targeted <- SumExp::row_df(overall_sumexp)$std_type != ""
targeted_chems_se <- overall_sumexp[is_targeted, ]
```

# Given data

{{< include chapters/_read-msdial.qmd >}}

```{r}
#| purl: true
# Output file
stopifnot(    # Required parameters
  !is.null(params$input_file), 
  !is.null(params$table_dir)
)
FILE <- list(
  i = params$input_file,
  o = local({
    b <- basename(params$input_file) |> 
      tools::file_path_sans_ext()    # Without extension
    rlang::list2(
      conc = rlang::list2(
        loess_norm = file.path(params$table_dir, paste0(b, ".loess.conc.tsv")),
        closest_norm = file.path(params$table_dir, paste0(b, ".closest.conc.tsv")),
      ),
      norm = rlang::list2(
        loess_norm = file.path(params$table_dir, paste0(b, ".loess.norm.tsv")),
        closest_norm = file.path(params$table_dir, paste0(b, ".closest.norm.tsv")),
      )
    )
  })
)
```

--------------------------------------------

# Normalization using internal standards

## Internal standard (IS) chemicals

```{r}
#| purl: true
internal_std_se0 <- internal_std_se <- proc$get_internal_std_se(overall_sumexp) |> 
  proc$add_num_zeros_rsd()

# Number of outlying internal standard chemicals per sample 
n_out_istd <- proc$count_outliers_per_sample(internal_std_se, assay_id = "raw", times = 3)
is_outlier <- n_out_istd > nrow(internal_std_se) * 0.2
if (any(is_outlier)) {
  # Exclude the outlying samples
  stopifnot(identical(names(is_outlier), colnames(internal_std_se)))
  internal_std_se <- internal_std_se[, !is_outlier] |> 
    proc$add_num_zeros_rsd()
  stopifnot(identical(names(is_outlier), colnames(overall_sumexp)))
  overall_sumexp <- overall_sumexp[, !is_outlier]
}

# Failed internal standards
is_failed_istd <- SumExp::row_df(internal_std_se)$num_zeros > 0
failed_istd <- internal_std_se[is_failed_istd, ]
# Remove failed internal standard chemicals
internal_std_se <- internal_std_se[!is_failed_istd, ]
```

{{< include chapters/_preprocess-internal-stds.qmd >}}

## Normalization

The data is normalized using the internal standard chemicals.

```{r}
#| purl: true
# Find the range of RT values
overall_rt_range <- range(SumExp::row_df(overall_sumexp)$rt)
targetted_rt_range <- range(SumExp::row_df(targeted_chems_se)$rt)
rt <- SumExp::row_df(internal_std_se)$rt
```

The range of RT values of overall chemicals is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted chemicals is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and 
the range of internal standard chemicals is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard chemicals

```{r}
#| label: 'normalize-rt-closest-istd'
#| purl: true

closest_istd_raw <- proc$get_raw_of_closest_istd(overall_sumexp, internal_std_se)
# Normalize the data using closest internal standard chemicals
overall_sumexp[["closest_norm"]] <- (overall_sumexp[["raw"]] / closest_istd_raw) |> 
  labelled::set_variable_labels("Closest RT Normalized")
```

{{< include chapters/_preprocess-norm-closest.qmd >}}

### LOESS fit along RT

```{r}
#| label: 'normalize-rt'
#| purl: true

excluded_category_for_normalization <- c("Blank", "CalCurve")

loess_fit <- proc$get_loess_fit(
  istd_se = internal_std_se,
  excl_cat = excluded_category_for_normalization,
  overall_rt_range = overall_rt_range,
  span = 1
)
# Normalize the data by LOESS fit along RT
rt <- SumExp::row_df(overall_sumexp)$rt
raw <- overall_sumexp[["raw"]]
overall_sumexp[["loess_norm"]] <- sapply(
  colnames(overall_sumexp), function(sample_id) {
    norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
    exp(log(raw[, sample_id]) - norm_factor)
  }
) |> 
  labelled::set_variable_labels("LOESS Normalized")
```

{{< include chapters/_preprocess-norm-loess.qmd >}}

### Comparison of the normalization methods

{{< include chapters/_preprocess-norm-comparison.qmd >}}


```{r}
#| label: 'export-normalized-data'
for(assy_id in c("loess_norm", "closest_norm")) {
  msdial$export_data_with_chem_table_tsv(
    sumexp = overall_sumexp, 
    assay_id = assy_id,
    in_file = FILE$i, 
    out_file = FILE$o$norm[[assy_id]]
  )
}
```

# Calibration

```{r}
#| label: 'calcurve-samples-extraction'
#| purl: true
# Extract quantification standard data
quant_se <- overall_sumexp[SumExp::row_df(overall_sumexp)$std_type == "Quant", ]
# Calibration curve samples only
calcurve_se <- quant_se[, SumExp::col_df(quant_se)$proc_cat == "CalCurve"]
stopifnot("Calibration curve samples are required." = nrow(calcurve_se) > 0)
# Calibration zero samples
cal_0_se <- calcurve_se[, SumExp::col_df(calcurve_se)$c_conc == 0] 
stopifnot("Multiple `Cal_0` samples are required." = ncol(cal_0_se) > 1)
# Remove the 0 concentration samples
calcurve_se <- calcurve_se[, SumExp::col_df(calcurve_se)$c_conc != 0]

# Find the lower-limit of quantification using values of the zero concentration samples
lloq_signal <- SumExp::row_df(calcurve_se)$lloq_signal <- 
  proc$identify_llox_signal(cal_0_se, DATA_ID, times = 10)
lloq <- SumExp::row_df(calcurve_se)$lloq <- proc$identify_lloq(calcurve_se, lloq_signal, DATA_ID)
# Find the lower-limit of detection using values of the zero concentration samples
llod_signal <- SumExp::row_df(calcurve_se)$llod_signal <- 
  proc$identify_llox_signal(cal_0_se, DATA_ID, times = 3)
llod <- SumExp::row_df(calcurve_se)$llod <- proc$identify_llod(calcurve_se, llod_signal, DATA_ID)
# Calibration curve concentration upper limit
max_c_conc <- proc$identify_max_conc(calcurve_se, quant_se, DATA_ID, times = 10)
concs <- sort(unique(SumExp::col_df(calcurve_se)$c_conc))
SumExp::row_df(calcurve_se)$max_c_conc <- proc$make_sure_to_have_enough_calcurve(
  concs, lloq, max_c_conc, min_n = 3, enough_n = 5
) 
# Mark the chemicals to exclude due to LLOQ and maximum concentration
SumExp::row_df(calcurve_se)$to_exclude <- as.list(SumExp::row_df(calcurve_se)) |> 
  with(is.na(lloq) | is.na(llod) | (max_c_conc == -9) | is.na(max_c_conc))
```

{{< include chapters/_preprocess-calib-conc_range.qmd >}}

```{r}
#| label: 'compute-concentration'
#| purl: true
# Exclude the marked chemicals
is_marked <- SumExp::row_df(calcurve_se)$to_exclude
stopifnot(identical(rownames(calcurve_se), rownames(quant_se)))
calcurve_se <- calcurve_se[! is_marked, ]
quant_se <- quant_se[! is_marked, ]
# Exclude the values outside the concentration range
calcurve_se <- local({
  cc <- calcurve_se
  # Concentration values in a matrix, rows are chemicals and columns are concentrations
  conc_mat <- matrix(rep(SumExp::col_df(cc)$c_conc, each = nrow(cc)), nrow = nrow(cc))
  to_na <- conc_mat < SumExp::row_df(cc)$lloq | conc_mat > SumExp::row_df(cc)$max_c_conc
  cc[["loess_norm"]][to_na] <- NA_real_
  cc
})

# Fit the calibration curve
c_conc <- SumExp::col_df(calcurve_se)$c_conc
rt_norm_mat <- calcurve_se[["loess_norm"]]
calcurve_models <- lapply(setNames(nm = rownames(calcurve_se)), function(ii) {
  proc$fit_and_test_calcurve_model(c_conc, rt_norm_mat[ii, ])
})

# Compute the concentration of the samples skipping the calibration curve samples
measurements_se <- quant_se[, ! SumExp::col_df(quant_se)$proc_cat %in% "CalCurve"]
SumExp::col_df(measurements_se)$c_conc <- NULL
measurements_se[["conc"]] <- 
  proc$compute_concentration(measurements_se, calcurve_se, calcurve_models, DATA_ID) |> 
  labelled::set_variable_labels("conc" = "Concentration (ng/mL)")
# Maximum/minimum concentration after trimming out of each chemical
maxmin_conc <- SumExp::row_df(calcurve_se)[c("max_c_conc", "lloq", "llod")]
stopifnot(identical(rownames(maxmin_conc), rownames(calcurve_se)))
SumExp::row_df(measurements_se) <- cbind(SumExp::row_df(measurements_se), maxmin_conc)
```

{{< include chapters/_preprocess-calibration.qmd >}}

# Concentration

## Computed concentrations

{{< include chapters/_preprocess-concentration.qmd >}}

## Blank subtraction

```{r}
#| label: 'blank-subtraction'
#| purl: true

# Blank subtraction
is_blank <- SumExp::col_df(measurements_se)$proc_cat == "Blank"
blank_se <- measurements_se[, is_blank]
measurements_se <- measurements_se[, ! is_blank]
blank_mean_conc <- rowMeans(blank_se[["conc"]])
conc <- measurements_se[["conc"]] - blank_mean_conc
conc[conc < 0] <- 0
measurements_se[["conc"]] <- conc
```

{{< include chapters/_preprocess-concentration.qmd >}}

```{r}
#| label: 'export-concentration-table'
for(assy_id in c("loess_norm", "closest_norm")) {
  msdial$export_concentration_tsv(measurements_se, FILE$o$conc[[assy_id]])
}
```


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

