---
params:
  data_id: loess_norm_blk
  title: Test
title: "Data quality evaluation"
format:
  html: 
    fig-width : 10
    fig-height: 6
    out-width: "100%"
    fig-align: "default"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
#| output: false
# The processed data used for calibration.
DATA_ID <- params$data_id

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  util = projlib/msdial_utils,        # Utility functions for MS-Dial data
  projlib/proc,       # Processing functions
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
IS_NON_TARGET_MODE <- SumExp::metadata(raw_se)$is_non_target_mode
# For consistency throughout the report
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$.ctrl_cat)

# Input/Output files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Processing NOT completed." = qc_steps[["Processing Completed"]])
calib_interm_data <- qc_steps[[paste0("calibration/", DATA_ID)]]

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
feature_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(feature_id, feature_name)
set.seed(7579)
```

# Data input

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization by internal standards (IS)

## Volumetric internal standard (vIS)

```{r}
#| purl: true
is_vIS <- util$std_type(raw_se) == "vIS"
# Feature intensity to be used for normalization
mat_id_for_norm <- if (any(is_vIS)) "vol_norm" else "raw"
```

```{r}
if (any(is_vIS)) {
  cat(
    "The data contains volumetric internal standards (vIS).\n",
    "All values of all other features are normalized by the value of the vIS feature",
    "for each sample.\n\n",
    "(individual value of each feature and each sample) / (vIS value of each sample) *", 
    "(average of vIS values of all samples)\n"
  )
} else {
  cat("There is no volumetric internal standard (vIS) in the data.\n")
}
```

## Internal standards (ISs) 

{{< include chapters/_process-internal-stds.qmd >}}

## Normalization by ISs

The dataset is normalized using the provided ISs. The user can opt to choose between 2 normalization methods: 
- "Closest RT" method, whereby the peak area intensity of each feature/chemical is corrected with that of the closest IS in RT. Target classes of compounds can be specified.
- "LOESS" method, whereby the peak area intensity of each feature/chemical is corrected by a LOESS model curve built using all the IS spanning across the entire chromatographic RT.

```{r}
#| purl: true
# Find the range of RT values
wo_vIS <- raw_se[!is_vIS, ]
all_rts <- util$retention_time(wo_vIS)
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[util$is_targeted_feature(wo_vIS)])
rt <- range(all_rts[util$is_internal_std(wo_vIS)])
```

The retention time (RT) range of the dataset is between `r overall_rt_range[1]` and `r overall_rt_range[2]` min.
The RT range of the targeted chemicals to be quantified is from `r targetted_rt_range[1]` to `r targetted_rt_range[2]` min.
The RT range of the internal standards is from `r range(rt)[1]` to `r range(rt)[2]` min.

### Closest RT (IS) normalization

```{r}
#| purl: true
norm_se <- qc_steps[["normalized"]]
```

{{< include chapters/_process-norm-closest.qmd >}}

### LOESS (IS) normalization

```{r}
#| child: code/reports/chapters/_process-norm-loess.qmd

```

### Comparison of normalization methods

{{< include chapters/_process-norm-comparison.qmd >}}


### Blank subtraction

After data normalization, peak areas are blank corrected by subtracting the corresponding feature average peak area detected in the blanks (negative values set to zero).
Blank subtraction is applied to all samples in the dataset except for the calibration standards.

```{r}
#| label: target_mode
if (!IS_NON_TARGET_MODE) {
  cat("\n\n# Calibration\n\n")
  qmd <- "_process-calibration-llox_from_llox_signal.qmd"
  cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))
  cat("\n\n# Concentration\n\n")
  cat(knitr::knit_child("code/reports/chapters/_process-concentration.qmd", quiet = TRUE))
}
```

# References
