---
title: "Exploratory Data Analysis"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
# The preprocessed data used for calibration
DATA_ID <- setNames("loess_norm", nm = "LOESS normalization")

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  projlib/io,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Read `params.yml` if `params` not provided
if (! exists("params")) params <- yaml::read_yaml("params.yml")

# Input/Output files
FILE <- list(
  i = rlang::list2(
    proc = msdial$get_sumexp_file_name(params, "proc"),
    qc = msdial$get_sumexp_file_name(params, "qc_steps"),   # Intermediate state of the data
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

measurement_blank_substracted_lst <- readRDS(FILE$i$proc)
overall_sumexp <- msdial$read_parsed_msdial_data(params)
qc_steps <- readRDS(FILE$i$qc)

sample_names_tbl <- SumExp::col_df(overall_sumexp) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
# Conversion table for chemical names
chem_names_tbl <- SumExp::row_df(overall_sumexp) |> 
  tibble::rownames_to_column("chem_id") |> 
  dplyr::select(chem_id, chem_name)
set.seed(7579)
```

# Given data

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization using internal standards

## Internal standard (IS) chemicals

{{< include chapters/_preprocess-internal-stds.qmd >}}

## Normalization

The data is normalized using the internal standard chemicals.

```{r}
#| purl: true
# Find the range of RT values
all_rts <- SumExp::row_df(overall_sumexp)$rt
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[SumExp::row_df(overall_sumexp)$std_type == "Quant"])
rt <- range(all_rts[SumExp::row_df(overall_sumexp)$std_type == "IS"])
```

The range of RT values of overall chemicals is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted chemicals is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and
the range of internal standard chemicals is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard chemicals

```{r}
#| purl: true
overall_sumexp <- qc_steps[["Normalized"]]
```

{{< include chapters/_preprocess-norm-closest.qmd >}}

### LOESS fit along RT

```{r}
#| child: code/reports/chapters/_preprocess-norm-loess.qmd

```

### Comparison of the normalization methods

{{< include chapters/_preprocess-norm-comparison.qmd >}}

# Calibration

{{< include chapters/_preprocess-calib-conc_range.qmd >}}

{{< include chapters/_preprocess-calibration.qmd >}}

# Concentration

## Computed concentrations

```{r}
#| purl: true
concn_se <- qc_steps[[DATA_ID]][["concn_se within range"]]     # Used in the following 
```

{{< include chapters/_preprocess-concentration.qmd >}}

## Blank subtraction

```{r}
#| purl: true
concn_se <- qc_steps[[DATA_ID]][["concn_blank_substracted"]]     # Used in the following 
```

{{< include chapters/_preprocess-concentration.qmd >}}

```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

