---
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-import'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")
box::use(
  ggplot2[...],        # too long to write and usually specific enough
  tidySummarizedExperiment,        # S3 methods for tidyverse style of SummarizedExperiment
  io = projlib/check_io_exist,
)

# Input files
FILE <- rlang::list2(
  in_xl = "data/raw_internal/Target_Exposomics_240827.xlsx",
)
# Check if the files exist
io$chq_all_files_exist(FILE)

set.seed(7579)
```

# Given data

## Sheet "Original data (MS-DIAL)"

```{r}
#| label: 'read-given-xlsx-and-save-sumexp'
#| purl: true
#| cache: true

# Parse the data into SummarizedExperiment
parse_to_sumexp <- function(data_fr_xlsx, i_row_sample, i_col_metab) {
  # Confirm the column names are as expected
  stopifnot(identical(
    colnames(data_fr_xlsx)[i_col_metab],
    c("Alignment ID", "Average Rt(min)", "Average Mz", "Metabolite name", "Comment", "Class")
  ))
  # extract metabolite info columns
  metabs <- data_fr_xlsx[-i_row_sample, i_col_metab, drop = FALSE] |> 
    # `metab_id` = Syntactically valid ID
    dplyr::mutate(metab_id = make.names(`Metabolite name`, unique = TRUE), .before = 1L) |> 
    # SummarizedExperiment will make this change anyhow
    dplyr::rename_with(~ make.names(.x))
  
  # Extract sample info rows
  sample_info <- t(data_fr_xlsx[i_row_sample, -i_col_metab, drop = FALSE])
  stopifnot(i_row_sample == 1)
  colnames(sample_info) <- "given_sample_id"
  # The row names contain the sample type
  sample_info <- tibble::as_tibble(sample_info, rownames = "type")
  sample_info <- sample_info |> 
    dplyr::mutate(sample_id = make.names(given_sample_id, unique = TRUE)) |>   # Syntactically valid
    dplyr::select(sample_id, type, given_sample_id)
  
  # Values of the metabolites into a matrix
  msdial_mat <- data_fr_xlsx[-i_row_sample, -i_col_metab, drop = FALSE]
  stopifnot(identical(colnames(msdial_mat), sample_info$type))
  colnames(msdial_mat) <- sample_info$sample_id     # Update the column names with sample IDs
  msdial_mat <- lapply(msdial_mat, as.numeric) |> 
    as.data.frame(row.names = metabs$metab_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(msdial = msdial_mat),
    colData = sample_info,
    rowData = metabs,
  )
}
# Read the data in the sheet "Original data (MS-DIAL)"
msdial_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "Original data (MS-DIAL)",
  .name_repair = "minimal",     # No change of column names
)
I_COL_ANALYTE <- 1:6
I_ROW_SAMPLE <- 1           # The first row has information about the samples

msdial_sumexp <- parse_to_sumexp(msdial_data, I_ROW_SAMPLE, I_COL_ANALYTE)
colnames(SummarizedExperiment::rowData(msdial_sumexp))[6] <- "std_type"    # `Comment`
```

The sheet "Original data (MS-DIAL)" contained `r nrow(msdial_data[-I_ROW_SAMPLE, ])` rows (or metabolites) and `r ncol(msdial_data)` columns.
Among the columns, the first `r length(I_COL_ANALYTE)` columns were the analytes' information (`r knitr::combine_words(colnames(msdial_data)[I_COL_ANALYTE], before = "\x60")`),
and the rest were the data of the samples.
The column names were the sample IDs including standards for calibration curves and quality control samples.

* Dimension of the data: `r dim(msdial_sumexp)`


## Sheets "QStd" and "IS"

```{r}
# Read the data in the sheet "QStd"
qstd_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "QStd",
  .name_repair = "minimal",     # No change of column names
)
# Confirm the file was in the correct format
stopifnot(identical(
  colnames(qstd_data),
  c("Compound name", "", "Rt", "mz", "Comment")
))
qstd_data <- qstd_data[-2]    # No data
colnames(qstd_data) <- c("Metabolite.name", "Rt", "Mz", "Comment_Qstd")
tmp <- dplyr::left_join(
  SummarizedExperiment::rowData(msdial_sumexp) |> as.data.frame(),
  qstd_data, 
  by = "Metabolite.name"
)
# Confirm "QStd" sheet contains duplicated data
stopifnot(all(is.na(tmp$Comment_Qstd) | paste0("Q", tmp$std_type) == tmp$Comment_Qstd))

# Read the data in the sheet "IS"
is_data <- readxl::read_xlsx(
  path = FILE$in_xl, 
  sheet = "IS",
  .name_repair = "minimal",     # No change of column names
)
# Confirm the file was in the correct format
stopifnot(identical(
  colnames(is_data),
  c("Compound name", "Rt", "mz", "Comment")
))
colnames(is_data) <- c("Metabolite.name", "Rt", "Mz", "Comment_IS")
is_data$Comment_IS <- "Istd"    # The values in `msdial_data` are `Istd`.
tmp <- dplyr::left_join(
  SummarizedExperiment::rowData(msdial_sumexp) |> as.data.frame(),
  is_data, 
  by = "Metabolite.name"
)
# Confirm "IS" sheet contains duplicated data
stopifnot(all(is.na(tmp$Comment_IS) | tmp$std_type == tmp$Comment_IS))
```

The sheets "QStd" and "IS" contained identical data to the sheet "Original data (MS-DIAL)". 
So the sheets are not used in the following analysis.

## Samples

There are `r dplyr::n_distinct(msdial_sumexp$type)` types of samples in the data.
The number of samples for each type is as follows:
```{r}
msdial_sumexp$type |> 
  table() |> 
  knitr::kable(col.names = c("Type", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Metabolites

The data contained `r nrow(msdial_sumexp)` metabolites.
The metabolites were classified into `r dplyr::n_distinct(SummarizedExperiment::rowData(msdial_sumexp)$std_type)` types.
The number of metabolites for each type is as follows:
```{r}
SummarizedExperiment::rowData(msdial_sumexp)$std_type |> 
  table() |> 
  knitr::kable(col.names = c("Standard Type", "Number of metabolites")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

### Overall samples and metabolites

#### Linear scale

```{r}
#| label: 'distribution-values-linear'

# Distribution of the values
ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values", x = "Values")
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-log'
ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values", x = "Values") +
  scale_x_log10()
```

### By sample type and overall metabolites

```{r}
#| label: 'distribution-values-by-type'
#| fig-width: 10
#| fig-height: 6
#| out.width: '100%'

ggplot(msdial_sumexp) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values by sample type", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ type, scales = "free_y", ncol = 4)
```

### By metabolite and overall samples

Randomly selected 8 metabolites are shown in the following figure.

```{r}
#| label: 'distribution-values-by-metabolite'
#| fig-width: 10
#| fig-height: 6
#| out.width: '100%'

i_selected <- sample(1:nrow(msdial_sumexp), 8)
ggplot(msdial_sumexp[i_selected, ]) +
  geom_histogram(aes(x = msdial), bins = 30) +
  labs(title = "Distribution of the values by metabolite", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Metabolite.name, scales = "free_x", ncol = 4)
```


----------------------------

# Normalization using internal standards

## Internal standard (IS) metabolites

```{r}
#| label: 'internal-standards'
#| purl: true

# Extract the internal standard metabolites
i_int_std <- SummarizedExperiment::rowData(msdial_sumexp)$std_type == "Istd"
internal_std_se <- msdial_sumexp[i_int_std, ]
# Sort by average m/z
mz <- SummarizedExperiment::rowData(internal_std_se)$Average.Mz
internal_std_se <- internal_std_se[order(mz), ]

# # Exclude the internal standard metabolites from the data
# qstd_sumexp <- msdial_sumexp[!i_int_std, ]
# stopifnot(all(SummarizedExperiment::rowData(qstd_sumexp)$std_type == "Std"))
```

There are `r nrow(internal_std_se)` internal standard (IS) metabolites in the data.
Those metabolites are `r knitr::combine_words(SummarizedExperiment::rowData(internal_std_se)$Metabolite.name, before = "\x60")`, sorted by `Average m/z`.

### Distribution of the values

Randomly selected 8 internal standard (IS) metabolites are shown in the following figure.

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 8
#| fig-height: 13
#| out-width: '80%'
#| fig-cap: 'Distribution of the values of internal standard (IS) metabolites'

i_selected <- sample(1:nrow(internal_std_se), 8) |> 
  sort()
internal_std_se[i_selected, ] |> 
  # keep the order by m/z. This invokes `tidySummarizedExperiment:::mutate.SummarizedExperiment`
  dplyr::mutate(Metabolite.name = factor(Metabolite.name, unique(Metabolite.name))) |> 
  ggplot() +
  geom_histogram(aes(x = msdial, fill = type), bins = 30) +
  labs(title = "Distribution of the values of internal standard metabolites", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Metabolite.name, scales = "free_x", ncol = 2)
```

## Normalization

### LOESS along the m/z range

```{r}
#| label: 'normalize-mz-range'
#| purl: true

# Find m/z range for which LOESS fit is performed
mz0 <- SummarizedExperiment::rowData(msdial_sumexp)$Average.Mz
overall_mz_rg <- range(mz0)

internal_std_mat <- SummarizedExperiment::assays(internal_std_se)$msdial
mz <- SummarizedExperiment::rowData(internal_std_se)$Average.Mz
```

The range of m/z values of overall samples is `r overall_mz_rg[1]` to `r overall_mz_rg[2]`,
while the range of m/z values of internal standard metabolites is `r range(mz)[1]` to `r range(mz)[2]`.

```{r}
#| label: 'normalize-mz'
#| purl: true

# Normalize the data using the internal standards
# Log-transform the data
internal_std_mat_log <- log1p(internal_std_mat)
# Overall mean of the internal standard metabolites
intstd_mean_log <- rowMeans(internal_std_mat_log)
# Subtract the mean from the data
internal_std_mat_norm_log <- internal_std_mat_log - intstd_mean_log
# Extend to smallest m/z with the mean of smallest 
i_smallest_mzs <- order(mz)[1]
avg_smallest_mzs <- colMeans(internal_std_mat_norm_log[i_smallest_mzs, , drop = FALSE])
# Extend to largest m/z with the mean of largest 
i_largest_mzs <- order(mz, decreasing = TRUE)[1]
avg_largest_mzs <- colMeans(internal_std_mat_norm_log[i_largest_mzs, , drop = FALSE])
# Add the extended values
internal_std_mat_norm_log <- rbind(
  avg_smallest_mzs, 
  avg_largest_mzs,
  internal_std_mat_norm_log
)

# LOESS fit of the internal standard metabolites along m/z
loess_fit <- purrr::map(1:ncol(internal_std_mat_norm_log), function(ii) {
  # Fit a loess curve
  y <- internal_std_mat_norm_log[, ii]
  stopifnot(identical(
    names(y), c("avg_smallest_mzs", "avg_largest_mzs", rownames(internal_std_se))
  ))
  loess(y ~ c(overall_mz_rg, mz), span = 0.3)
})
names(loess_fit) <- colnames(internal_std_mat_norm_log)

# Normalize the data by m/z
mz <- SummarizedExperiment::rowData(msdial_sumexp)$Average.Mz
msdial <- SummarizedExperiment::assays(msdial_sumexp)$msdial
SummarizedExperiment::assays(msdial_sumexp)$mz_norm <- 
  sapply(
    SummarizedExperiment::colData(msdial_sumexp)$sample_id,
    function(sample_id) {
      norm_factor <- predict(loess_fit[[sample_id]], newdata = mz)
      expm1(log1p(msdial[, sample_id]) - norm_factor)
    }
  )
```

```{r}
#| label: 'normalize-mz-plot'
#| fig-width: 14
#| fig-height: 6
#| out.width: '100%'

# Plot the normalized and original data for randomly selected 3 samples
i_selected <- sample(1:ncol(msdial_sumexp), 3)
# To include LOESS smooth line
x <- seq(overall_mz_rg[1], overall_mz_rg[2], length.out = 100)
loess_df <- lapply(setNames(i_selected, msdial_sumexp$sample_id[i_selected]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(Average.Mz = x) |>
  tidyr::pivot_longer(-Average.Mz, names_to = "sample_id", values_to = "Value")

# Log-transform the data
msdial_log <- log1p(SummarizedExperiment::assays(msdial_sumexp)$msdial)
# Overall mean of every metabolite
mean_log <- rowMeans(msdial_log)
# Deviation from the mean
selected_se <- msdial_sumexp[, i_selected]
SummarizedExperiment::assays(selected_se) <- SummarizedExperiment::assays(selected_se) |> 
  lapply(function(x) sweep(log1p(x), 1, mean_log))

tidySummarizedExperiment:::as_tibble.SummarizedExperiment(selected_se) |> 
  dplyr::rename(Raw = msdial, Normalized = mz_norm) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |>
  ggplot(aes(x = Average.Mz, y = Value)) +
  geom_point(aes(color = Data, shape = std_type)) +
  geom_line(data = loess_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average m/z", 
    y = "log(Values + 1) - log(avg + 1)"
  ) +
  scale_shape_manual(values = c("Istd" = 15, "Std" = 21)) +
  facet_wrap(~ sample_id, scales = "free_y", ncol = 3)
```

### LOESS fit along RT

```{r}
#| label: 'normalize-rt-range'
#| purl: true

# Find the range of RT values
rt0 <- SummarizedExperiment::rowData(msdial_sumexp)$Average.Rt.min.
overall_rt_rg <- range(rt0)

internal_std_mat <- SummarizedExperiment::assays(internal_std_se)$msdial
rt <- SummarizedExperiment::rowData(internal_std_se)$Average.Rt.min.
```

The range of RT values of overall samples is `r overall_rt_rg[1]` to `r overall_rt_rg[2]`,
while the range of RT values of internal standard metabolites is `r range(rt)[1]` to `r range(rt)[2]`.

```{r}
#| label: 'normalize-rt'
#| purl: true

# Normalize the data using the internal standards
# Log-transform the data
internal_std_mat_log <- log1p(internal_std_mat)
# Overall mean of the internal standard metabolites
intstd_mean_log <- rowMeans(internal_std_mat_log)
# Subtract the mean from the data
internal_std_mat_norm_log <- internal_std_mat_log - intstd_mean_log
# Extend to smallest RT with the mean of smallest 
i_smallest_rts <- order(rt)[1]
avg_smallest_rts <- colMeans(internal_std_mat_norm_log[i_smallest_rts, , drop = FALSE])
# Extend to largest RT with the mean of largest 
i_largest_rts <- order(rt, decreasing = TRUE)[1]
avg_largest_rts <- colMeans(internal_std_mat_norm_log[i_largest_rts, , drop = FALSE])
# Add the extended values
internal_std_mat_norm_log <- rbind(
  avg_smallest_rts, 
  avg_largest_rts,
  internal_std_mat_norm_log
)

# LOESS fit of the internal standard metabolites along RT
loess_fit <- purrr::map(1:ncol(internal_std_mat_norm_log), function(ii) {
  # Fit a loess curve
  y <- internal_std_mat_norm_log[, ii]
  stopifnot(identical(
    names(y), c("avg_smallest_rts", "avg_largest_rts", rownames(internal_std_se))
  ))
  loess(y ~ c(overall_rt_rg, rt), span = 0.75)
})
names(loess_fit) <- colnames(internal_std_mat_norm_log)

# Normalize the data by RT
rt <- SummarizedExperiment::rowData(msdial_sumexp)$Average.Rt.min.
msdial <- SummarizedExperiment::assays(msdial_sumexp)$msdial
SummarizedExperiment::assays(msdial_sumexp)$rt_norm <- 
  sapply(
    SummarizedExperiment::colData(msdial_sumexp)$sample_id,
    function(sample_id) {
      norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
      expm1(log1p(msdial[, sample_id]) - norm_factor)
    }
  )
```

```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 6
#| out.width: '100%'

# Plot the normalized and original data for randomly selected 3 samples
# i_selected <- sample(1:ncol(msdial_sumexp), 3)
# To include LOESS smooth line
x <- seq(overall_rt_rg[1], overall_rt_rg[2], length.out = 100)
loess_df <- lapply(setNames(i_selected, msdial_sumexp$sample_id[i_selected]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(Average.Rt.min. = x) |>
  tidyr::pivot_longer(-Average.Rt.min., names_to = "sample_id", values_to = "Value")

# Log-transform the data
msdial_log <- log1p(SummarizedExperiment::assays(msdial_sumexp)$msdial)
# Overall mean of every metabolite
mean_log <- rowMeans(msdial_log)
# Deviation from the mean
selected_se <- msdial_sumexp[, i_selected]
SummarizedExperiment::assays(selected_se) <- SummarizedExperiment::assays(selected_se) |> 
  lapply(function(x) sweep(log1p(x), 1, mean_log))

# Plot the normalized and original data
selected_se |>
  tidySummarizedExperiment:::as_tibble.SummarizedExperiment() |>
  dplyr::rename(Raw = msdial, Normalized = rt_norm) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |>
  ggplot(aes(x = Average.Rt.min., y = Value)) +
  geom_point(aes(color = Data, shape = std_type)) +
  geom_line(data = loess_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values + 1) - log(avg + 1)"
  ) +
  scale_shape_manual(values = c("Istd" = 15, "Std" = 21)) +
  facet_wrap(~ sample_id, scales = "free_y", ncol = 3)
```


# Calibration curves

```{r}
#| label: 'divide-samples-by-type'
#| purl: true

# Divide the data by the sample type
i_by_type <- split(1:ncol(msdial_sumexp), msdial_sumexp$type)
msdial_by_type <- purrr::map(i_by_type, ~msdial_sumexp[, .x])
```


## Calibration curve samples

```{r}
#| label: 'calcurve-samples'
#| purl: true


# Extract concentration values from the given IDs
msdial_by_type[["CalCurve"]]$conc <- local({
  given_sample_id <- msdial_by_type[["CalCurve"]]$given_sample_id 
  conc <- stringr::str_extract(given_sample_id, "Cal(|Std)([0-9]+)_", group = 2)
  dplyr::case_when(
    stringr::str_detect(given_sample_id, "Cal0_") ~ "0",
    stringr::str_detect(conc, "^0[0-9]") ~ stringr::str_replace(conc, "^0", "0."),
    TRUE ~ conc
  ) |> 
    as.numeric()
})
calcurve_se <- msdial_by_type[["CalCurve"]]
```

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
calcurve_se$conc |> 
  table() |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration", "Number of samples")
  )
```

## Calibration curve per metabolite

### Linear scale

#### Before normalization

```{r}
#| label: 'calcurve-per-metabolite-linear-before-norm'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"
set.seed(7579)
i_sel <- sample(1:nrow(calcurve_se), 9)
calcurve_se <- calcurve_se[i_sel, ]

# Plot calibration curves
tidySummarizedExperiment:::ggplot.SummarizedExperiment(calcurve_se, aes(x = conc, y = msdial)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Raw)"
  ) +  
  facet_wrap(~ Metabolite.name, scales = "free_y", ncol = 3)
```

#### After normalization

```{r}
#| label: 'calcurve-per-metabolite-linear'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"
# Plot calibration curves
tidySummarizedExperiment:::ggplot.SummarizedExperiment(calcurve_se, aes(x = conc, y = rt_norm)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ Metabolite.name, scales = "free_y", ncol = 3) +
  geom_smooth(
    aes(weight = 1 / (conc + 1) ^ 2),
    method = "lm",
    se = FALSE,
    alpha = 0.1,
    color = "cadetblue"
  )
```


### Log scale

After normalization

```{r}
#| label: 'calcurve-per-metabolite-log'
#| fig-width: 12
#| fig-height: 11
#| out-width: "100%"

tidySummarizedExperiment:::ggplot.SummarizedExperiment(calcurve_se, aes(x = conc, y = rt_norm)) +
  geom_point() +
  labs(
    x = "Concentration",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ Metabolite.name, scales = "free_y", ncol = 3)

```


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

