---
params:
  norm_method: closest_norm
  project_title: "Test"
title: "Report on `r params$project_title`"
format:
  titlepage-pdf:
    titlepage: academic
    titlepage-logo: "../projlib/images/SciLifeLab_Logotype_Green_POS.pdf"
    logo-align: center
    logo-size: 10
    #mainfont: cursive
    echo: false 
    fig-align: center  # Assure line breaks between figures
    documentclass: report
    papersize: a4
    classoption: landscape
    geometry:
      - margin=20mm
      - heightrounded
---

```{r}
#| label: 'setup'
#| purl: true
# The processed data used for calibration. Blank subtracted
if (!exists("params")) params <- list(norm_method = "closest_norm") # For "Run Cell" functionality

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  util = projlib/msdial_utils,        # Utility functions for MS-Dial data
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
DATA_ID <- util$mat_id_of_blank_subtracted(params$norm_method)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
IS_NON_TARGET_MODE <- SumExp::metadata(raw_se)$is_non_target_mode
# For consistency throughout the report
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$.ctrl_cat)

# Input files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Processing NOT completed." = qc_steps[["Completed"]])
calib_interm_data <- qc_steps[[paste0("calibration/", DATA_ID)]]

concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
# Sort the chemicals by name
concn_se <- concn_se[order(SumExp::row_df(concn_se)$feature_name), ]
# Exclude control samples
concn_se <- util$exclude_ctrl_smpl_cat(concn_se, c("QC"))
```

# Project Information

* User: `r user_inputs$user`
* University: `r user_inputs$university`
* Project: `r user_inputs$project_title`

`r user_inputs$free_text`

{{< pagebreak >}}

# General overview

* Number of samples: `r ncol(concn_se)`
* Number of targeted chemicals: `r nrow(concn_se)`

```{r}
show$kable_number_of(SumExp::col_df(concn_se)$Class)
```

{{< pagebreak >}}

## PCA

```{r}
#| label: 'pca'
#| fig-width: 8
#| fig-height: 4
#| out-width: '100%'

plot_pca <- function(se, mat_id) {
  mat <- t(se[[mat_id]])
  # Scale is not possible with all zeros
  mat <- mat[, colSums(mat, na.rm = TRUE) > 0, drop = FALSE]
  # Remove constant rows
  mat <- mat[, apply(mat, 2, var, na.rm = TRUE) > 0, drop = FALSE]
  # If there are no features, return an empty plot
  if (ncol(mat) <= 1) {
    return(ggplot() + ggtitle("No features for PCA"))
  }
  pca_res <- stats::prcomp(mat, scale. = TRUE)
  
  se_df <- SumExp::col_df(se)      # `data` in ggplot2::autoplot
  # Plot
  ggplot2::autoplot(pca_res, data = se_df, colour = "Class")
}
both_se <- qc_steps[["normalized - blank"]] |>
  util$exclude_ctrl_smpl_cat(c("QC", "CalCurve", "Blank"))
both_se <- both_se[!util$is_targeted_feature(both_se), ]       # Exclude internal standards
p1 <- plot_pca(both_se, DATA_ID) + ggtitle("Target + Non-target")
p2 <- plot_pca(concn_se, DATA_ID) +
  ggtitle("Target")
p1 + p2 + patchwork::plot_layout(guides = "collect")
```

{{< pagebreak >}}

## Chemical summary

```{r}
#| label: 'chemicals-summary'

# Summary table
summary_tbl <- show$tbl_chemical_summary(concn_se) |> 
  dplyr::mutate(      # Tidy up the table
    chem_name = stringr::str_trunc(chem_name, 40),
    perc_detf = round(perc_detf, 1),
    model_r2 = round(model_r2, 3),
    dplyr::across(c(median, mean, min, max, lod, lloq), ~ round(.x, 2))
  ) |> 
  dplyr::select(
    "Chemical" = chem_name,
    "DF%" = perc_detf,
    "N of sampl." = n_det,
    "LOD" = lod,
    "LLOQ" = lloq,
    "Median" = median,
    "Average" = mean,
    "Min" = min,
    "Max" = max,
    "R2" = model_r2,
    "Model" = best_model,
    "N of points" = n_conc,
  )
summary_tbl |> 
  knitr::kable(caption = "Summary of Targets") |> 
  kableExtra::kable_styling(
    latex_options = "striped", 
    full_width = FALSE, 
    repeat_header_continued = TRUE,
  ) |> 
  kableExtra::footnote(
    general = "DF: Detection frequency (percentage of samples above LOD)",
    footnote_as_chunk = TRUE
  )
```

**Note**

The unit of concentration is `r user_inputs$concentration_unit`.

```{r}
#| label: 'lod-lloq-limits-text'
qmd <- "_external-llox_from_llox_signal.qmd"
cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))

```

After the computation of concentration, the following steps are applied to the data:

  * Any negative concentration is adjusted to zero.
  * Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
  * Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
  * Any chemicals with values below LLOQ in all samples is removed from the data.


{{< pagebreak >}}

# Invidual chemicals

{{< pagebreak >}}

```{r}
#| label: 'individual-chemicals'
#| fig-width: 12
#| fig-height: 6
#| out-width: '100%'

#' Text box to describe a calibration curve model
#' 
textbox_of_a_width <- function(txt, width) {
  ggplot() +
    ggtext::geom_textbox(
      aes(x = 0, y = 0.9, label = txt),
      box.color = "black",
      width = width,
      fill = NA,
      hjust = 0,
      vjust = 1,
      size = 4
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_void()
}

calcurve_se <- calib_interm_data[["calcurve_se within range"]]
c_concs <- util$spiked_conc_pts(calcurve_se)
lowest_c <- min(c_concs[c_concs > 0])

for (i_chem in seq_len(nrow(concn_se))) {       # For each chemical
  chem_name <- SumExp::row_df(concn_se)$feature_name[i_chem]
  cat("\n\n## ", chem_name, "\n")
  
  x_se <- concn_se[i_chem, ]
  # Calibration curves
  x1_se <- x_se   # For the case, lod < 0 or lloq < 0
  SumExp::row_df(x1_se) <- SumExp::row_df(x1_se) |> 
    dplyr::mutate(dplyr::across(c(lod, lloq), ~ ifelse(.x < 0, lowest_c / 10, .x)))
  p1 <- show$ggplot_calcurve_samples(x1_se, 
                                     calcurve_se, 
                                     mat_id = DATA_ID,
                                     COLORS_OF_CLASSES) + 
    labs(
      title = "Full range",
      y = "Peak area (Normalized)"
    )
  p2 <- p1 + 
    show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID) +
    labs(title = "Zoom-in (3 cal. pts.)")
  # Distribution of the concentration by class
  x_df <- SumExp::as_tibble(x_se)
  p3 <- ggplot(x_df) +
    geom_boxplot(aes(x = Class, y = conc, color = Class, fill = Class)) +
    labs(
      x = show$label_if_has(x_df$Class, "Class"),
      y = show$label_if_has(x_df$conc, "Concentration"),
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +    # Vertical x-axis
    theme(legend.position = "none") +      # No legend
    scale_y_log10() +
    scale_color_manual(values = COLORS_OF_CLASSES) +
    scale_fill_manual(values = COLORS_OF_CLASSES)
  
  # Text in the box
  text1 <- local(envir = show$tbl_chemical_summary(x_se), {
    perc_detf <- sprintf("%.1f%%", perc_detf)
    paste0(
      "Best model: ", best_model, "  \n",
      "R2: ", round(model_r2, 3), "  \n",
      "N of points: ", n_conc, "  \n",
      "DF: ", perc_detf, " (", n_det, "/", n_samples, ")", "  \n",
      "LOD: ", sprintf("%.2f", lod), "  \n",
      "LLOQ: ", sprintf("%.2f", lloq), "  \n",
      ""         # For trailing comma
    ) |> 
      textbox_of_a_width(width = unit(7, "cm"))
  })
  
  txt_legend <- paste0(
    "The gray areas in the calibration curve represent the regions below LOD and LLOQ.  \n",
    "DF: Detection frequency (percentage of samples above LOD)  \n",
    "LOD: Limit of detection.\n",
    "Lowest concentration at which the signal is three times the standard deviation above the lowest non-zero calibration point, as determined by the calibration model.  \n",
    "LLOQ: Lower limit of quantification\n", 
    "Ten times the standard deviation  \n",
    "The details of LOD and LLOQ are described in the text of the chemical summary table.  \n",
    ""         # For trailing comma
  ) |>
    textbox_of_a_width(width = unit(16, "cm"))
  
  # Plot
  design <- "
    112233
    112233
    445555
  "
  print(p1 + p2 + p3 + text1 + txt_legend +
          patchwork::plot_layout(design = design, guides = "collect"))
}
```

