---
title: "Report on {{< var title >}}"
format:
  titlepage-pdf:
    titlepage: academic
    titlepage-logo: "../projlib/images/SciLifeLab_Logotype_Green_POS.pdf"
    logo-align: center
    logo-size: 10
    #mainfont: "../projlib/fonts/Lora/Lora-VariableFont_wght.ttf"
    echo: false 
    fig-align: center  # Assure line breaks between figures
    documentclass: report
    papersize: a4
    classoption: landscape
    geometry:
      - margin=20mm
      - heightrounded
---

```{r}
#| label: 'setup'
#| purl: true
# The preprocessed data used for calibration. Default is `loess_norm_blk`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm_blk"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
IS_NON_TARGET_MODE <- SumExp::metadata(raw_se)$is_non_target_mode
# For consistency throughout the report
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$contr_cat)

# Input files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])

concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
CHEMICAL_IDS <- SumExp::rownames(concn_se)#[1:2]

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
chem_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(chem_id = feature_id, chem_name = feature_name)
```

# Project Information

* User: `r user_inputs$user`
* University: `r user_inputs$university`
* Project: `r user_inputs$project_title`

`r user_inputs$free_text`

{{< pagebreak >}}

# General overview

* Number of samples: `r ncol(concn_se)`
* Number of targeted chemicals: `r nrow(concn_se)`

{{< pagebreak >}}

## PCA

```{r}
#| label: 'pca'
#| fig-width: 8
#| fig-height: 4
#| out-width: '100%'

plot_pca <- function(se, mat_id) {
  mat <- t(se[[mat_id]])
  mat <- mat[, colSums(mat, na.rm = TRUE) > 0]       # Scale is not possible with all zeros
  mat <- mat[, apply(mat, 2, var, na.rm = TRUE) > 0]        # Remove constant rows
  pca_res <- stats::prcomp(mat, scale. = TRUE)
  
  se_df <- SumExp::col_df(se)      # `data` in ggplot2::autoplot
  # Plot
  ggplot2::autoplot(pca_res, data = se_df, colour = "Class")
}
norm_se <- qc_steps[["normalized - blank"]]
p1 <- norm_se[, quote(! contr_cat %in% c("QC", "CalCurve", "Blank"))] |> 
  plot_pca(DATA_ID) + ggtitle("Target + Non-target")
p2 <- plot_pca(concn_se[, quote(contr_cat != "QC")], DATA_ID) +
  ggtitle("Target")
p1 + p2 + patchwork::plot_layout(guides = "collect")
```

{{< pagebreak >}}

## Chemical summary

```{r}
#| label: 'chemicals-summary'

# Information about the chemicals
chemicals <- SumExp::row_df(concn_se) |> 
  tibble::as_tibble(rownames = "chem_id") |> 
  dplyr::rename(chem_name = "feature_name")
# Concentration
mat <- concn_se[["conc"]]
# Summary about the concentration ranges
conc_summary <- tibble::tibble(
  chem_id = chemicals$chem_id,
  llod = chemicals$llod,
  lloq = chemicals$lloq,
  df_n = sapply(1:nrow(mat), \(i) sum(mat[i, ] >= llod[i])),
  df_perc = df_n / ncol(mat) * 100,
  median = apply(mat, 1, median),
  mean = rowMeans(mat),
  min = apply(mat, 1, min),
  max = apply(mat, 1, max),
)
conc_summary <- conc_summary |> 
  dplyr::mutate(
    df_perc = scales::percent(df_perc / 100),
    median = round(median, 2),
    mean = round(mean, 2),
    min = round(min, 2),
    max = round(max, 2)
  )
# Summary about the calibration curve models
model_summary <- qc_steps$calibration[[DATA_ID]][["calcurve_models"]] |> 
  lapply(\(m) {
    type <- m$best_model_name
    R2 <- round(m$R2s[[type]], 3)
    n_conc <- m$n_conc
    data.frame(R2, type, n_conc) 
  }) |> 
  purrr::list_rbind(names_to = "chem_id")
# Merge
summary_tbl <- list(
  chemicals[, c("chem_id", "chem_name")],
  conc_summary,
  model_summary
) |> 
  purrr::reduce(dplyr::left_join, by = "chem_id") |> 
  dplyr::arrange(chem_name) |> 
  dplyr::select(
    "Chemical" = chem_name,
    "DF (%)" = df_perc,
    "N of sampl." = df_n,
    "LLOD" = llod,
    "LLOQ" = lloq,
    "Median" = median,
    "Average" = mean,
    "Min" = min,
    "Max" = max,
    "R2" = R2,
    "Model" = type,
    "N of points" = n_conc,
  )

# Summary table
summary_tbl |> 
  knitr::kable(caption = "Summary of Targets") |> 
  kableExtra::kable_styling(
    latex_options = "striped", 
    full_width = FALSE, 
    repeat_header_continued = TRUE,
  ) |> 
  kableExtra::footnote(
    general = "DF: Detection frequency (percentage of samples above LLOD)",
    footnote_as_chunk = TRUE
  )
```

**Note**

* The unit of concentration is `r user_inputs$concentration_unit`.
* The lower-limit of quantification (LLOQ) and lower-limit of detection (LLOD) are calculated using zero concentration samples.
The LLOD is defined as the lowest concentration at which the average signal of calibration samples exceeds three times the mean of the zero concentration samples. 
Similarly, the LLOQ is set at ten times the mean of the zero concentration samples.
If the calibration samples at the LLOQ concentration exhibit high variability (relative standard deviation, RSD, > 20%), the LLOQ is adjusted to the lowest higher concentration that satisfies RSD ≤ 20%.

{{< pagebreak >}}

# Invidual chemicals

{{< pagebreak >}}

```{r}
#| label: 'individual-chemicals'
#| fig-width: 12
#| fig-height: 6
#| out-width: '100%'
calcurve_se <- qc_steps$calibration[[DATA_ID]][["calcurve_se within range"]]
calcurve_models <- qc_steps$calibration[[DATA_ID]][["calcurve_models"]]
cc_models <- lapply(calcurve_models, \(.x) .x$inv_model)
for(chem_i in stringr::str_sort(CHEMICAL_IDS)) {
  chem_name <- chem_id_name_tbl$chem_name[chem_id_name_tbl$chem_id == chem_i]
  cat("\n\n## ", chem_name, "\n")
  
  x_se <- concn_se[rownames(concn_se) == chem_i, quote(! contr_cat %in% c("QC", "Blank"))]
  cc_se <- calcurve_se[rownames(calcurve_se) == chem_i, ]
  # Calibration curves
  p1 <- show$ggplot_calcurve_samples(x_se, 
                                     calcurve_se, 
                                     cc_models, 
                                     mat_id = DATA_ID,
                                     COLORS_OF_CLASSES) + 
    labs(
      title = "Linear scale",
      y = "Peak area (Normalized)"
    )
  p2 <- p1 + scale_x_log10() + scale_y_log10() + 
    labs(title = "Log10 scale")
  # Distirbution of the concentration by class
  x_df <- SumExp::as_tibble(x_se)
  p3 <- ggplot(x_df) +
    geom_boxplot(aes(x = Class, y = conc, color = Class, fill = Class)) +
    labs(
      x = show$label_if_has(x_df$Class, "Class"),
      y = show$label_if_has(x_df$conc, "Concentration"),
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +    # Vertical x-axis
    theme(legend.position = "none") +      # No legend
    scale_y_log10() +
    scale_color_manual(values = COLORS_OF_CLASSES) +
    scale_fill_manual(values = COLORS_OF_CLASSES)
  # Text
  mdl <- calcurve_models[[chem_i]]
  fts <- SumExp::row_df(x_se)
  conc <- x_se[["conc"]][1, ]
  text_df <- tibble::tibble(
    x = 0,
    y = 0.9,
    label = paste0(
      "Best model: ", mdl$best_model_name, "  \n",
      "R2: ", round(summary(mdl$inv_model)$r.squared, 3), "  \n",
      "N of points: ", mdl$n_conc, "  \n",
      "DF: ", scales::percent(sum(conc >= fts$llod) / length(conc)), 
      " (", sum(conc >= fts$llod), "/", length(conc), ")", "  \n",
      "LLOD: ", fts$llod, "  \n",
      "LLOQ: ", fts$lloq, "  \n",
      ""         # For trailing comma
    )
  )
  text1 <- ggplot(text_df) +
    ggtext::geom_textbox(
      aes(x, y, label = label),
      box.color = "black",
      width = unit(7, "cm"),
      fill = NA,
      hjust = 0,
      vjust = 1,
      size = 4
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_void()
  
  txt_legend <- tibble::tibble(
    x = 0, y = 0.9, 
    label = paste0(
      "The gray areas in the calibration curve represent the regions below LLOD and LLOQ.  \n",
      "DF: Detection frequency (percentage of samples above LLOD)  \n",
      "LLOD: Lower limit of detection.\n",
      "Lowest concentration at which the average signal of calibration samples exceeds three times the mean of the zero concentration sample  \n",
      "LLOQ: Lower limit of quantification\n", 
      "Ten times the mean of the zero concentration samples  \n",
      ""         # For trailing comma
    )
  )
  txt_legend <- ggplot(txt_legend) +
    ggtext::geom_textbox(
      aes(x, y, label = label),
      box.color = "black",
      width = unit(16, "cm"),
      fill = NA,
      hjust = 0,
      vjust = 1,
      size = 4
    ) +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_void()
  
  # Plot
  design <- "
    112233
    112233
    445555
  "
  print(p1 + p2 + p3 + text1 + txt_legend +
          patchwork::plot_layout(design = design, guides = "collect"))
}
```

