---
title: "Exploratory data analysis"
format:
  html: 
    fig-width : 10
    fig-height: 6
    out-width: "100%"
    fig-align: "default"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
#| output: false
# The preprocessed data used for calibration. Default is `loess_norm_blk`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm_blk"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
IS_NON_TARGET_MODE <- SumExp::metadata(raw_se)$is_non_target_mode
# For consistency throughout the report
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$contr_cat)

# Input/Output files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])
calib_interm_data <- qc_steps$calibration[[DATA_ID]]

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
feature_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(feature_id, feature_name)
set.seed(7579)
```

# Given data

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization using internal standards

## Volumetric internal standard

```{r}
#| purl: true
is_vIS <- SumExp::row_df(raw_se)$std_type == "vIS"
# Feature intensity to be used for normalization
mat_id_for_norm <- if (any(is_vIS)) "vol_norm" else "raw"
```

```{r}
if (any(is_vIS)) {
  cat(
    "The data contains volumetric internal standards (vIS).\n",
    "All values of all other features are normalized by the value of the vIS feature",
    "for each sample.\n\n",
    "(individual value of each feature and each sample) / (vIS value of each sample) *", 
    "(average of vIS values of all samples)\n"
  )
} else {
  cat("There is no volumetric internal standard (vIS) in the data.\n")
}
```

## Internal standard (IS) features

{{< include chapters/_preprocess-internal-stds.qmd >}}

## Normalization

The data is normalized using the internal standard features.

```{r}
#| purl: true
# Find the range of RT values
wo_vIS_df <- SumExp::row_df(raw_se[!is_vIS, ])
all_rts <- wo_vIS_df$rt
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[wo_vIS_df$std_type == "Quant"])
rt <- range(all_rts[wo_vIS_df$std_type == "IS"])
```

The range of RT values of overall features is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted features is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and
the range of internal standard features is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard features

```{r}
#| purl: true
norm_se <- qc_steps[["normalized"]]
```

{{< include chapters/_preprocess-norm-closest.qmd >}}

### LOESS fit along RT

```{r}
#| child: code/reports/chapters/_preprocess-norm-loess.qmd

```

### Comparison of the normalization methods

{{< include chapters/_preprocess-norm-comparison.qmd >}}


### Blank subtraction

The normalized values were subtracted by the average of blank samples to remove the background signals.
As calibration samples were prepared differently from all the other samples including the blank samples, blank subtraction was not applied to the calibration samples.
Any values below 0 are replaced with 0 to avoid any negative values. 

```{r}
#| label: target_mode
if (!IS_NON_TARGET_MODE) {
  cat("\n\n# Calibration\n\n")
  cat(knitr::knit_child("code/reports/chapters/_preprocess-calib-conc_range.qmd", quiet = TRUE))
  cat(knitr::knit_child("code/reports/chapters/_preprocess-calibration.qmd", quiet = TRUE))
  cat("\n\n# Concentration\n\n")
  cat(knitr::knit_child("code/reports/chapters/_preprocess-concentration.qmd", quiet = TRUE))
}
```

# References
