---
title: "Exploratory data analysis"
format:
  html: 
    fig-width : 10
    fig-height: 6
    out-width: "100%"
    fig-align: "default"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true
# The preprocessed data used for calibration. Default is `loess_norm`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Read `params.yml` to get various project parameters
params <- yaml::read_yaml("params.yml")
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors

# Input/Output files
FILE <- list(
  i = rlang::list2(
    proc = msdial$get_sumexp_file_name(params, "proc"),
    qc = msdial$get_sumexp_file_name(params, "qc_steps"),   # Intermediate state of the data
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

measurement_blank_substracted_lst <- readRDS(FILE$i$proc)
overall_sumexp <- msdial$read_parsed_msdial_data(params)
colors_of_classes <- show$get_colors_of_classes(overall_sumexp, COLORS$contr_cat)
# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing was not completed." = qc_steps[["Preprocessing Completed"]])
# Conversion tables
sample_id_name_tbl <- qc_steps[["sample_id_name_tbl"]]
feature_id_name_tbl <- qc_steps[["feature_id_name_tbl"]]
set.seed(7579)
```

# Given data

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization using internal standards

## Internal standard (IS) features

{{< include chapters/_preprocess-internal-stds.qmd >}}

## Normalization

The data is normalized using the internal standard features.

```{r}
#| purl: true
# Find the range of RT values
all_rts <- SumExp::row_df(overall_sumexp)$rt
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[SumExp::row_df(overall_sumexp)$std_type == "Quant"])
rt <- range(all_rts[SumExp::row_df(overall_sumexp)$std_type == "IS"])
```

The range of RT values of overall features is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted features is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and
the range of internal standard features is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard features

```{r}
#| purl: true
overall_sumexp <- qc_steps[["Normalized"]]
```

{{< include chapters/_preprocess-norm-closest.qmd >}}

### LOESS fit along RT

```{r}
#| child: code/reports/chapters/_preprocess-norm-loess.qmd

```

### Comparison of the normalization methods

{{< include chapters/_preprocess-norm-comparison.qmd >}}

# Calibration

{{< include chapters/_preprocess-calib-conc_range.qmd >}}

```{r}
#| child: code/reports/chapters/_preprocess-calibration.qmd

```

# Concentration

## Computed concentrations

```{r}
#| purl: true
concn_se <- qc_steps[[DATA_ID]][["concn_se with conc"]]     # Used in the following 
```

{{< include chapters/_preprocess-concentration.qmd >}}

## Blank subtraction

```{r}
#| purl: true
concn_se <- qc_steps[[DATA_ID]][["concn_minus_blk_se"]]     # Used in the following 
```

{{< include chapters/_preprocess-concentration.qmd >}}

# References
