---
title: "Exploratory Data Analysis - Exposome"
tbl-cap-location: top
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true

# Load packages and project local libraries
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[rowData, `rowData<-`, colData, `colData<-`, assays, `assays<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
)
# Input file
FILE <- list(i = "data/raw_internal/water_samples_original_ESipos_MSdial_output (not blank filtered).txt")
set.seed(7579)
```

```{r}
#| label: 'read-given-tsv-and-save-sumexp'
#| purl: true

#' Identify three column sections in the MS-DIAL output file
#' 
#' @param msdial_file Path to the MS-DIAL output file
#' @return A list of three column indices
get_three_section_indices <- function(msdial_file) {
  # Read the second (`File type`) line to identify three sections
  file_type_line <- read.delim(
    msdial_file, header = FALSE, nrow = 5L, na.strings = c("NA", "")
  )[2, ]
  i_2nd <- which(! is.na(file_type_line))    # Not NA is the 2nd column sections
  stopifnot(file_type_line[i_2nd[1L]] == "File type")    # As expected
  list(
    "1st" = 1L:i_2nd[1L],
    "2nd" = (i_2nd[1L] + 1L):max(i_2nd),         # "+ 1L" for the title column of the rows
    "3rd" = (max(i_2nd) + 1L):length(file_type_line)
  )
}

#' Fetch sample information from the first 5 lines
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the columns that contain sample information
#' @return A tibble with sample information
fetch_sample_info <- function(msdial_file, indices) {
  if (missing(indices)) indices <- get_three_section_indices(msdial_file)[[2]]
  # Read sample information lines (first 5 lines)
  sample_lines <- read.delim(
    msdial_file, 
    header = FALSE,
    nrow = 5L,
    na.strings = c("NA", "")
  )
  header <- sample_lines[, indices[1L] - 1L]
  
  # Extract sample information  
  sinfo <- t(sample_lines[, indices, drop = FALSE])
  colnames(sinfo) <- header          # There was no header in `sinfo` before
  stopifnot("Deviation from the expected names for samples" = identical(
    colnames(sinfo)[-5], c("Class", "File type", "Injection order", "Batch ID")
  ))
  colnames(sinfo)[5] <- "given_sample_id"
  # Create a tibble with the sample information
  sinfo <- tibble::as_tibble(sinfo) |> 
    dplyr::select(
      Class,
      sample_type = "File type",
      injection_order = "Injection order",
      given_sample_id,
    ) |> 
    # Syntactically valid for variable name in R
    dplyr::mutate(sample_id = make.names(given_sample_id, unique = TRUE), .before = 1L) |>
    # Control sample groups
    dplyr::mutate(
      proc_cat = dplyr::case_when(
        sample_type == "Standard" & Class == "CalCurve" ~ "CalCurve",
        sample_type == "QC"    ~ "QC",
        sample_type == "Blank" ~ "Blank",
        TRUE ~ ""         # # NA does not behave predictably with `==`
      )
    )
  stopifnot(
    "Calibration curve samples are required." = any(sinfo$proc_cat == "CalCurve"),
    "QC samples are required." = any(sinfo$proc_cat == "QC"),
    "Blank samples are required." = any(sinfo$proc_cat == "Blank")
  )
  
  # Function to find concentration values from the given IDs
  find_concentration <- function(sid) {
    sid |> 
      stringr::str_extract("Cal_([[:digit:]-]+)", group = 1) |> 
      stringr::str_replace("-", ".") |>     # Replace "-" with "."
      as.numeric()
  }
  sinfo <- sinfo |> 
    dplyr::mutate(conc = ifelse(proc_cat == "CalCurve", 
                                find_concentration(given_sample_id),
                                NA_real_))
  stopifnot(
    "Error in Calibration sample IDs" = all(! is.na(sinfo$conc[sinfo$proc_cat == "CalCurve"]))
  )
  return(sinfo)
}

#' Fetch data of the selected columns of the MS-DIAL output file
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the selected columns
#' @return A tibble 
fetch_data_of_columns <- function(msdial_file, indices) {
  stopifnot(!missing(indices))
  # Count total number of columns in the file
  l1 <- readr::read_tsv(msdial_file, skip = 4L, n_max = 1L, col_names = FALSE, col_types = "c")
  coltypes <- rep("-", ncol(l1))      # Skip the rest columns
  coltypes[indices] <- "c"
  # Read lines that contain chemical data
  readr::read_tsv(
    msdial_file, 
    skip = 4L,      # Data starts from the 5th line
    col_types = paste0(coltypes, collapse = ""),
    name_repair = "unique_quiet"
  )
}

# Parse the data in the file into SummarizedExperiment
parse_to_sumexp <- function(msdial_file) {
  i_three_sections <- get_three_section_indices(msdial_file)
  sample_info <- fetch_sample_info(msdial_file, i_three_sections[[2]])
  
  # Read chemical data in the first section
  chems <- fetch_data_of_columns(msdial_file, i_three_sections[[1]]) |>
    dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
      alignment_id = "Alignment ID",
      chem_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      sn_ratio = "S/N average",
      std_type = "Comment",            # Type of standard, e.g. "Quant", "IS", or NA 
    ) |> 
    # `chem_id` = Syntactically valid ID
    dplyr::mutate(chem_id = make.names(chem_name, unique = TRUE), .before = 1L) |> 
    dplyr::mutate(
      dplyr::across(c(mz, rt, sn_ratio), as.numeric),
      std_type = dplyr::case_when(
        std_type == "Quant" ~ "Quant",
        std_type == "IS"    ~ "IS",
        TRUE ~ ""              # Ignore other variants
      )
    )
  
  # Measured values of the chemicals into a matrix
  raw_df <- fetch_data_of_columns(msdial_file, i_three_sections[[2]])
  stopifnot(identical(colnames(raw_df), sample_info$given_sample_id))
  colnames(raw_df) <- sample_info$sample_id     # Update with syntactically valid names
  raw_mat <- lapply(raw_df, as.numeric) |> 
    as.data.frame(row.names = chems$chem_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(raw = raw_mat),
    colData = sample_info,
    rowData = chems,
  )
}
overall_sumexp <- parse_to_sumexp(FILE$i)  # To be updated throughout the pre-processing
```

Given file: `` `r FILE$i` `` 

The given file contains data of `r nrow(overall_sumexp)` chemicals and `r ncol(overall_sumexp)` samples.

## Samples

### By class

There are `r dplyr::n_distinct(overall_sumexp$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
table(overall_sumexp$Class) |> 
  knitr::kable(col.names = c("Class", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

### Special categories for preprocessing

There are `r length(table(overall_sumexp$proc_cat, exclude = ""))` special sample categories for in the data.
The data of those are important in the following processing steps.
The number of samples for each category is as follows:

```{r}
table(overall_sumexp$proc_cat, exclude = "") |> 
  knitr::kable(col.names = c("Category", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

#### Calibration curve samples

The samples for the calibration curve are identified by the sample type `Standard` and class `CalCurve`.
The concentration is extracted from the provided sample ID.
The IDs are in the format `Cal_[[:digit:]-]+` with optional numbers and/or characters before and after of it. 
Concentration is extracted from the numeric part in the format after the replacement of `-` with `.`.

##### Sample IDs

```{r}
# Present the sample IDs for the calibration curve
df_show <- colData(overall_sumexp[, overall_sumexp$proc_cat == "CalCurve"]) |> 
  tibble::as_tibble() |> 
  dplyr::arrange(conc)
df_show |> 
  dplyr::select(given_sample_id, conc) |>
  knitr::kable(
    col.names = c("Sample ID", "Concentration (ng/ml)"),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

##### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
table(df_show$conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Quality control samples

```{r}
#| purl: true
# Extract QC samples into list
extract_qc_samples_to_list <- function(se) {
  se <- se[, se$proc_cat == "QC"]
  qc_id <- se$Class
  # Column-wise split
  lapply(setNames(nm = unique(qc_id)), \(ii) se[, qc_id == ii])
}
```

Aliquots of some samples were measured as technical replicates for quality control.
The number of the replicates were as below.

```{r}
extract_qc_samples_to_list(overall_sumexp) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Sample ID", "Number of replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Chemicals

The data contained `r nrow(overall_sumexp)` chemicals.
Among them, `r sum(rowData(overall_sumexp)$std_type != "")` chemicals were targeted. 
The rest were not targeted.

```{r}
#| purl: true
# Extract standard chemicals
is_targeted <- rowData(overall_sumexp)$std_type != ""
targeted_chems_se <- overall_sumexp[is_targeted, ]
```

For those targeted chemicals, the standard type was specified.
The number of chemicals for each type is as follows:

```{r}
table(rowData(targeted_chems_se)$std_type) |> 
  knitr::kable(col.names = c("Standard Type", "Number of chemicals")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

### Overall samples and targeted chemicals

#### Linear scale

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 8 
#| out-width: '60%'
p <- ggplot(targeted_chems_se) +
  geom_histogram(aes(x = raw), bins = 50) +
  labs(title = "Distribution of the values", x = "Values")
print(p)
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-overall-log'
#| fig-width: 8
#| out-width: '60%'
p + scale_x_log10()
```

### By sample class and overall targeted chemicals

```{r}
#| label: 'distribution-values-by-class'
ggplot(targeted_chems_se) +
  geom_histogram(aes(x = raw), bins = 30) +
  labs(title = "Distribution of the values by sample class", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Class, scales = "free_y", ncol = 4)
```

### By chemical and overall samples

Randomly selected 8 targeted chemicals are shown in the following figure.

```{r}
#| purl: true
#' Randomly select `n` rows from `sumexp` with optional `subset`
#'
#' @param sumexp a SummarizedExperiment object
#' @param n the number of rows to select
#' @param subset an optional subset expression to apply to the `rowData` of `sumexp`
#' @param keep_the_order if `TRUE`, the selected rows are sorted in the previous order
#'
#' @return a SummarizedExperiment object with `n` rows randomly selected from `sumexp`
#' @export
select_rows_randomly <- function(sumexp, n, subset, keep_the_order = FALSE) {
  ii <- c(1:nrow(sumexp))
  if (!missing(subset)) {      # If given, apply it to the `rowData` of `sumexp`
    ii <- ii[eval(substitute(subset), as.data.frame(rowData(sumexp)))]
  }
  ii <- sample(ii, n)
  if (keep_the_order) ii <- sort(ii)
  sumexp[ii, ]
}
```

```{r}
#| label: 'distribution-values-by-chemical'
se <- select_rows_randomly(targeted_chems_se, 8, std_type != "IS")
ggplot(se) +
  geom_histogram(aes(x = raw, fill = Class), bins = 30) +
  labs(title = "Distribution of the values by chemical", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ chem_name, scales = "free_x", ncol = 4)
```

----------------------------

# Normalization using internal standards

## Internal standard (IS) metabolites

```{r}
#| purl: true

# Extract the internal standard metabolites
internal_std_se <- given_data_sumexp |> 
  dplyr::filter(std_type == "Istd")
# Sort by average retention time
rt <- rowData(internal_std_se)$rt
internal_std_se <- internal_std_se[order(rt), ]
```

There are `r nrow(internal_std_se)` internal standard (IS) metabolites in the data.
Those metabolites are `r knitr::combine_words(rowData(internal_std_se)$metab_name, before = "\x60")`, in the order of average retention time.

### Distribution of the values

Randomly selected 8 internal standard (IS) metabolites are shown in the following figure.

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 8

# Keep the order by m/z
istd_to_show_se <- select_rows_randomly(internal_std_se, 8, in_previous_order = TRUE)
istd_to_show_se |> 
  # Below invokes `tidySummarizedExperiment:::mutate.SummarizedExperiment`
  dplyr::mutate(metab_name = forcats::fct(metab_name)) |>  # Keep the order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard metabolites", 
    y = "Peak area",
  ) +
  theme(legend.position = "none") +   # No legend for fill
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Rotate x-axis labels
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "fixed", ncol = 4)
```

### Outliers

Outliers were identified based on the values of internal standard metabolites.
Those samples with values greater than 3 standard deviations from the mean after log-transformation ($log(1 + x)$) were considered as outliers.

```{r}
#| purl: true

# Identify outliers
identify_outliers <- function(x) {
  m <- mean(x)
  sd <- sd(x)
  return((x > m + 3 * sd) | (x < m - 3 * sd))
}
# Number of outliers per sample
n_outlier <- assays(internal_std_se)$raw |> 
  log1p() |>              # Log-transform the data
  apply(1, identify_outliers) |>
  rowSums()         # Number of outliers per sample (transposed by `apply` above)
# Exclude the outliers
stopifnot(identical(names(n_outlier), colnames(internal_std_se)))
internal_std_se <- internal_std_se[, n_outlier == 0, drop = FALSE]
stopifnot(identical(names(n_outlier), colnames(given_data_sumexp)))
given_data_sumexp$is_outlier <- n_outlier > 0
no_outlier_se <- given_data_sumexp[, n_outlier == 0, drop = FALSE]
```

There are `r sum(n_outlier > 0)` outlying samples in the data.
The data of those outliers were excluded in the following analysis. 

```{r}
#| label: 'outliers-table'
#| results: 'markup'
print(n_outlier[n_outlier > 0])
```

After exclusion of those outliers, the distribution of the randomly selected 8 internal standard (IS) metabolites are shown below.

```{r}
#| label: 'distribution-values-internal-standards-no-outliers'
#| fig-width: 13
#| fig-height: 8

internal_std_se[rownames(istd_to_show_se), ] |>
  dplyr::mutate(metab_name = forcats::fct(metab_name)) |>  # Order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard metabolites (no outliers)", 
    y = "Values"
  ) +
  theme(legend.position = "none") +   # No legend for fill
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +    # Rotate x-axis labels
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "fixed", ncol = 4)
```

### raw RSD%

The relative standard deviation (RSD%) of the raw values of the internal standard metabolites are calculated.

#### By quality control groups

```{r}
#| purl: true
#' Calculate RSD%
#'
#' @param x A numeric vector
#' @return A numeric value of RSD%
#' @export
rsd_perc <- function(x) {
  100 * sd(x) / mean(x)
}
```

```{r}
#| label: 'rsd-internal-standards-by-qc-groups'

# Divide the internal standard metabolites by quality control groups
# Calculate RSD% per internal standard metabolite and per quality control groups
sapply(setNames(nm = unique(na.omit(internal_std_se$qc_group))), \(grp) {
  in_grp <- ! is.na(internal_std_se$qc_group) & internal_std_se$qc_group == grp
  se1 <- internal_std_se[, in_grp]
  apply(assay(se1, "raw"), 1, \(.x) round(rsd_perc(.x), 2))
}) |>
  knitr::kable(caption = "RSD% of internal standard metabolites by quality control groups") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### By sample classes

```{r}
sapply(setNames(nm = unique(internal_std_se$Class)), function(cls) {
  se1 <- internal_std_se[, internal_std_se$Class == cls]
  apply(assay(se1, "raw"), 1, \(.x) round(rsd_perc(.x), 2))
}) |> 
  knitr::kable(caption = "RSD% of internal standard metabolites by sample classes") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Normalization

The data will be normalized using the internal standard metabolites.

```{r}
#| purl: true

# Find the range of RT values
overall_rt_range <- range(rowData(no_outlier_se)$rt)
rt <- rowData(internal_std_se)$rt
```

The range of RT values of overall metabolites is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of internal standard metabolites is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard metabolites

The closest internal standard metabolites to each metabolite will be identified based on the retention time.
The measured peak area of each metabolite will be divided by the peak area of the closest internal standard metabolite.

```{r}
#| label: 'normalize-rt-closest-istd'
#| purl: true

# Get the values of the closest internal standard metabolites
get_raw_of_closest_istd <- function(rt_x, internal_std_se) {
  # Find the closest internal standard metabolite
  closest_istd <- which.min(abs(rowData(internal_std_se)$rt - rt_x))
  return(assays(internal_std_se)$raw[closest_istd, ])
}
assays(no_outlier_se)$closest_istd_raw <- rowData(no_outlier_se)$rt |> 
  setNames(rownames(no_outlier_se)) |>          # To keep the names
  sapply(get_raw_of_closest_istd, internal_std_se) |> 
  t()     # Rows are metabolites

# Normalize the data using closest internal standard metabolites
assays(no_outlier_se)$closest_norm <- 
  assays(no_outlier_se)$raw / assays(no_outlier_se)$closest_istd_raw
```

#### Evaluation of normalization

```{r}
#| purl: true

# Extract quantitative standards in QC samples
extract_quant_qc <- function(sumexp) {
  unique(na.omit(sumexp$qc_group)) |> 
    setNames(nm = _) |> 
    lapply(\(grp) {
      in_grp <- ! is.na(sumexp$qc_group) & sumexp$qc_group == grp
      is_qstd <- rowData(sumexp)$std_type == "Std"
      sumexp[is_qstd, in_grp]
    })
}
# Calculate RSD of the quantification standard samples
calc_rsd_qstd <- function(qc_se, assay_ids) {
  lapply(qc_se, \(qc1) {
    sapply(setNames(nm = assay_ids), function(nm) {
      apply(assays(qc1)[[nm]], 1, rsd_perc)
    }) |>    # Rows are metabolites, columns are assay IDs
      tibble::as_tibble(rownames = "metab_id")
  }) |> 
    # `QC` has the name of the QC samples
    dplyr::bind_rows(.id = "QC")
}
# Plot the RSD% of various data sets of metabolites
ggplot_rsdp_metab <- function(rsd_df, assay_ids) {
  nms <- names(assay_ids)
  ids <- unname(assay_ids)    # Duplicated by `dplyr::all_of`
  rsd_df <- rsd_df |> 
    dplyr::mutate(
      # NA was generated when all values were zero
      dplyr::across(dplyr::all_of(ids), \(.x) ifelse(is.na(.x), 0, .x)),
    ) |> 
    tidyr::pivot_longer(dplyr::all_of(ids), names_to = "Data", values_to = "RSD%") |> 
    dplyr::mutate(Data = factor(Data, levels = ids, labels = nms))
  ggplot(rsd_df) +
    aes(x = Data, y = `RSD%`) +
    geom_line(aes(group = metab_id), alpha = 0.5)
}
```
```{r}
#| label: 'evaluation-normalization-closest-istd'
#| out-width: '80%'

# Quantification standard samples
qc_se <- extract_quant_qc(no_outlier_se)

assay_ids <- setNames(c("raw", "closest_norm"), nm = c("Raw", "Closest IS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > closest_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of metabolites with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'evaluation-normalization-closest-istd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```


### LOESS fit along RT

```{r}
#| label: 'normalize-rt'
#| purl: true

loess_fit <- local({
  # Log-transform the data
  istd_log <- log(assays(internal_std_se)$raw)
  rt <- rowData(internal_std_se)$rt
  
  # Normalize the data using the internal standards
  # Mean of each internal standard metabolite for overall measurement samples
  # , excluding the calibration curve and blank samples
  to_excl <- internal_std_se$Class %in% c("Blank", "CalCurve")
  # The data of the samples with measurements
  mat <- istd_log[, !to_excl, drop = FALSE]
  stopifnot(all(is.finite(mat)))
  istd_m <- rowMeans(mat)
  # Subtract the mean from the data
  fr_m_log <- istd_log - istd_m
  
  # Expand to smallest RT 
  istd_smallest_rt <- fr_m_log[which.min(rt), ]
  # Expand to largest RT 
  istd_largest_rt <- fr_m_log[which.max(rt), ]
  # Add the expanded range
  fr_m_log <- rbind(
    istd_smallest_rt, 
    istd_largest_rt,
    fr_m_log
  )
  
  # LOESS fit of the internal standard metabolites along RT
  loess_fit <- purrr::map(1:ncol(fr_m_log), function(ii) {
    # Fit a loess curve
    y <- fr_m_log[, ii]
    y[is.finite(y) == FALSE] <- NA
    stopifnot(identical(
      names(y), c("istd_smallest_rt", "istd_largest_rt", rownames(internal_std_se))
    ))
    suppressWarnings(loess(y ~ c(overall_rt_range, rt), span = 1))
  })
  names(loess_fit) <- colnames(fr_m_log)
  loess_fit
})

# Normalize the data by LOESS fit along RT
rt <- rowData(no_outlier_se)$rt
raw <- assays(no_outlier_se)$raw
assays(no_outlier_se)$loess_norm <- sapply(
  colnames(no_outlier_se), function(sample_id) {
    norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
    exp(log(raw[, sample_id]) - norm_factor)
  }
)

# Extract quantification standard data
qstd_sumexp <- no_outlier_se |> 
  dplyr::filter(std_type == "Std")
```

For each metabolite, the average of all samples excluding `Blank` and `CalCurve` samples was calculated after log-transformation.
The deviation of each sample from the average was calculated as normalization factor.
The data of the sample are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) metabolites along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all metabolites. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all metabolites.
The normalized data are calculated by subtracting the normalization factor from the log-transformed data.

Randomly selected 3 samples are shown below.

```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 12

to_excl <- no_outlier_se$Class %in% c("Blank", "CalCurve")
# Plot the normalized and original data for randomly selected 3 samples
i_sel <- sample(c(1:ncol(no_outlier_se))[!to_excl], 3)

# To include LOESS smooth line
x <- seq(overall_rt_range[1], overall_rt_range[2], length.out = 100)
loess_curve_df <- lapply(setNames(i_sel, colnames(no_outlier_se)[i_sel]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(rt = x) |>
  tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")
loess_curve_df$sample_name <-
  no_outlier_se$sample_name[match(loess_curve_df$sample_id, no_outlier_se$sample_name)]

# Log-transform the data
raw_log <- log(assays(no_outlier_se)$raw)
# Overall mean of every metabolite
mat <- raw_log[, !to_excl, drop = FALSE]
mat[is.finite(mat) == FALSE] <- NA    # Visualization purpose only
mean_log <- rowMeans(mat, na.rm = TRUE)
# Deviation from the mean
selected_se <- no_outlier_se[, i_sel]
assays(selected_se) <- assays(selected_se) |> 
  lapply(\(x) sweep(log(x), 1, mean_log, FUN = "-"))

df_to_show <- tibble::as_tibble(selected_se) |>
  dplyr::rename(Raw = raw, Normalized = loess_norm) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value")

# Plot the normalized and original data
ggplot(df_to_show, aes(x = rt, y = Value)) +
  geom_point(aes(color = Data, shape = std_type)) +
  geom_line(data = loess_curve_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values) - log(avg)"
  ) +
  scale_shape_manual(values = c("Istd" = 15, "Std" = 21)) +
  facet_wrap(~ sample_name, scales = "free_y", ncol = 3)
```

The line in the graph represents the LOESS fit.
The points at the bottom boundary are the **0**s, which were not normalized by this process.
Those zeros were kept as zeros in the normalized data.

#### Evaluation of normalization

The normalized data are evaluated by the changes of relative standard deviation (RSD) of the quantification standard samples.

```{r}
#| label: 'rsd-eval'
#| out-width: '80%'

# Quantification standard samples
qc_se <- extract_quant_qc(qstd_sumexp)
# Calculate RSD of the quantification standard samples
assay_ids <- setNames(c("raw", "loess_norm"), nm = c("Raw", "LOESS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > loess_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of metabolites with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'rsd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

### Comparison of the normalization methods

The RSD% of the quantification standard samples is calculated for the raw data and the normalized data.

```{r}
#| label: 'norm-method-comparison'
#| out-width: '80%'

assay_ids <- setNames(c("raw", "closest_norm", "loess_norm"), nm = c("Raw", "Closest", "LOESS"))
df_to_show <- extract_quant_qc(qstd_sumexp) |> 
  calc_rsd_qstd(assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(loess_norm <= closest_norm, "↓ LOESS", "↓ Closest"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓ LOESS", "↓ Closest", "All 0s"))
  ) 
ggplot_rsdp_metab(df_to_show, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of metabolites with lower RSD after each normalization is counted below.

```{r}
#| label: 'rsd-comparison-count'
#| results: 'markup'
# Count the number of metabolites with lower RSD
df_to_show |> 
  janitor::tabyl(QC, sign)
```

The number of compounds above or below 20 RSD% is counted below.

```{r}
#| label: 'rsd-comparison-count-20'
df_to_show |> 
  dplyr::summarize(
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x > 20, na.rm = TRUE), 
      .names = "{.col}_gt_20"
    ),
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x <= 20, na.rm = TRUE), 
      .names = "{.col}_eq_lt_20"
    ),
    .by = QC
  ) |>  
  knitr::kable(
    caption = "Number of metabolites with RSD% above or below 20", 
    col.names = c("QC", rep(names(assay_ids), 2))
  ) |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "RSD > 20%" = 3, "RSD ≤ 20%" = 3)) 
```


The RSD% values are presented in the table below.

```{r}
#| label: 'rsd-comparison-table'

df_to_show |> 
  dplyr::left_join(as.data.frame(rowData(qstd_sumexp)), by = "metab_id") |>   # Add metabolite names
  dplyr::select(QC, metab_name, dplyr::all_of(unname(assay_ids))) |> 
  dplyr::rename(all_of(assay_ids), `Metabolite` = metab_name) |> 
  # Caption at the top
  knitr::kable(caption = "RSD% of the quantification standard samples", digits = 2) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |>
  kableExtra::scroll_box(height = "400px")
```


# Calibration

## Calibration curve samples

The samples for the calibration curve are identified by the sample class, `CalCurve`.
The data of `0` concentration (sample ID `Cal0_`) has been ignored. 
The concentration is extracted from the sample ID, which is in the format `CalStd[0-9]+_`.
The concentration is converted to a numeric value. 

```{r}
#| label: 'calcurve-samples'
#| purl: true

# Calibration curve samples only
calcurve_se <- qstd_sumexp[, qstd_sumexp$Class == "CalCurve"]
# Remove the 0 concentration sample
calcurve_se <- calcurve_se[, ! stringr::str_detect(calcurve_se$sample_name, "Cal0_")]

# Extract concentration values from the given IDs
extract_conc <- function(sample_name) {
  conc <- stringr::str_extract(sample_name, "CalStd([0-9]+)_", group = 1)
  dplyr::case_when(
    # Replace the leading `0` with `0.`
    stringr::str_detect(conc, "^0[0-9]") ~ stringr::str_replace(conc, "^0", "0."),
    TRUE ~ conc
  ) |> 
    as.numeric()
}
calcurve_se$conc <- extract_conc(calcurve_se$sample_name)
```

### Sample IDs

```{r}
#| results: 'markup'
# Present the sample IDs for the calibration curve
print(calcurve_se$sample_name)
```

### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
calcurve_se$conc |> 
  table() |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  )
```

## Concentration ranges

```{r}
#| purl: true
classes_not_for_measurement <- c("CalCurve", "QC", "Qcspiked")
```

Quite a few concentrations of calibration curve samples of a compound may be too far from the range of the measured samples. 
Hence the concentrations were excluded for which the peak areas after normalization were more than 10 times higher than the maximum peak area of the samples of `r knitr::combine_words(classes_not_for_measurement, before = "\x60")` classes.

```{r}
#| purl: true

# Find maximum peak area of the samples of the classes for measurement
rowData(calcurve_se)$max_loess_norm <- local({
  stopifnot(identical(rownames(qstd_sumexp), rownames(calcurve_se)))
  sumexp <- qstd_sumexp[, ! qstd_sumexp$Class %in% classes_not_for_measurement]
  apply(assays(sumexp)$loess_norm, 1, max, na.rm = TRUE)
})
# Function to replace .x with NA if it is more than 10 times higher than .y
replace_if_10x_higher <- function(.x, .y) {
  ifelse(.x > 10 * .y, NA_real_, .x)
}
# Exclude the concentrations that are too far from the measured samples
assays(calcurve_se)$loess_norm <- assays(calcurve_se)$loess_norm |> 
  replace_if_10x_higher(rowData(calcurve_se)$max_loess_norm)
rowData(calcurve_se)$max_conc <- assays(calcurve_se)$loess_norm |> 
  apply(1, \(.x) {
    if (all(is.na(.x))) return(NA_real_)
    max(calcurve_se$conc[!is.na(.x)])
  })
```

Remaining concentration range of each compound is presented below.

```{r}
#| label: 'calcurve-conc-range-table'

rowData(calcurve_se) |> 
  tibble::as_tibble() |> 
  dplyr::select(metab_name, max_conc) |>
  dplyr::mutate(
    max_conc = dplyr::case_when(
      is.na(max_conc) ~ "All 0s, no valid",
      TRUE ~ scales::number(max_conc)
    )
  ) |>
  knitr::kable(
    caption = "Concentration range of the calibration curve samples",
    col.names = c("Metabolite", "Maximum concentration (ng/ml)"),
    align = c("l", "r"),
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "400px")
```


## Calibration curve per metabolite

```{r}
#| purl: true

set.seed(7579)
i_sel <- c(1:nrow(calcurve_se))
i_sel <- i_sel[rowData(calcurve_se)$std_type != "Istd"]
i_sel <- sample(i_sel, 9)
calcurve_to_show_df0 <- tibble::as_tibble(calcurve_se[i_sel, ])
# Linear model fitting
calcurve_to_show_lmfit <- calcurve_to_show_df0 |> 
  tidyr::nest(.by = c(metab_id, metab_name)) |> 
  dplyr::mutate(
    lmfit = lapply(data, \(.x) lm(formula = raw ~ conc, data = .x)),# weight = 1 / (conc ^ 2))),
    rsq = sapply(lmfit, \(.x) summary(.x)$r.squared),
  )

# Fitted line layer
fit_layer <- geom_smooth(
  # aes(weight = 1 / (conc ^ 2)),
  data = calcurve_to_show_df0,
  method = "lm",
  formula = y ~ x + I(x^2),
  se = FALSE,
  color = "cadetblue"
)

calcurve_to_show_df <- calcurve_to_show_df0 |> 
  tidyr::drop_na(loess_norm)
```

### Linear scale

#### Before normalization

```{r}
#| label: 'calcurve-per-metabolite-linear-before-norm'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df0, aes(x = conc, y = raw)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Raw)"
  ) +  
  facet_wrap(~ metab_name, scales = "free_y", ncol = 3)
```

#### After normalization

The peak areas were normalized by the internal standards.
A line is fitted to the data points per metabolite.
~~Here the weight in the linear regression is inversely proportional to the squared concentration.~~

```{r}
#| label: 'calcurve-per-metabolite-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = loess_norm)) +
  geom_text(
    aes(x = Inf, y = -Inf, label = sprintf("R2 = %.2f", rsq)),
    data = calcurve_to_show_lmfit,
    hjust = 1.1,
    vjust = -0.1,
  ) +
  fit_layer +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```


### Log scale

After normalization

```{r}
#| label: 'calcurve-per-metabolite-log'
#| fig-width: 12
#| fig-height: 11
ggplot(calcurve_to_show_df, aes(x = conc, y = loess_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free", ncol = 3)

```

#### After normalization by closest

```{r}
#| label: 'calcurve-per-metabolite-linear-closest'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = closest_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```


## Calibration of the metabolites

The calibration curve is fitted to the [normalized data](#normalization) using a linear model ~~with the weights of $\frac{1}{x^2}$, where $x$ is the concentration~~.

### All measurement samples

```{r}
#| label: 'calibration-data'
#| purl: true

build_calcurve_model <- function(conc, value) {
  if (all(is.na(value))) {
    return(function(x) rep(NA, length(x)))
  }
  lmfit <- lm(value ~ conc + I(conc^2))#, weights = 1 / (conc ^ 2))
  beta <- coef(lmfit)
  a <- beta[["I(conc^2)"]]
  b <- beta[["conc"]]
  cc <- beta[["(Intercept)"]]
  function(x) {
    det <- b ^ 2 - 4 * a * (cc - x)
    out <- suppressWarnings((-b + sqrt(det)) / (2 * a))      # det < 0
    out[det < 0 | out < 0] <- 0
    out
  }
}

#build_calcurve_model <- function(conc, value) {
#  if (all(is.na(value))) {
#    return(function(x) rep(NA, length(x)))
#  }
#  lmfit <- lm(value ~ conc, weights = 1 / (conc ^ 2))
#  beta <- coef(lmfit)
#  b1 <- beta[["conc"]]
#  b0 <- beta[["(Intercept)"]]
#  function(x) {
#    (x - b0) / b1
#  }
#}

# Fit the calibration curve
conc <- calcurve_se$conc
rt_norm_mat <- assays(calcurve_se)$loess_norm
calcurve_fit <- lapply(setNames(nm = rownames(calcurve_se)), function(ii) {
  build_calcurve_model(conc, rt_norm_mat[ii, ])
})

# Compute the concentration of the samples
classes_to_incl <- unique(qstd_sumexp$Class) |> 
  setdiff(c("CalCurve", "RTI"))      # Exclude the control samples
measurements_se <- qstd_sumexp[, qstd_sumexp$Class %in% classes_to_incl]
assays(measurements_se)$conc <- rowData(measurements_se)$metab_id |> 
  sapply(function(metab_id) {
    v <- assays(measurements_se)$loess_norm[metab_id, ]
    calcurve_fit[[metab_id]](v)
  }) |>
    t()          # Metabolites to rows
```

The concentration of the sample classes, `r knitr::combine_words(classes_to_incl)`, are computed using the calibration curve.
Any negative concentration is set to zero.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
measure_tbl <- tibble::as_tibble(measurements_se[i_sel, ])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  fit_layer +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
# Retrieve the coefficients of the linear model for the calibration curve
metab_ids <- unique(measure_tbl$metab_id)
# beta <- lapply(metab_ids, function(m_id) {
#   eval(quote(c(cc, b, a)), environment(calcurve_fit[[m_id]]))
# }) |> 
#   do.call(what = rbind) |> 
#   as.data.frame() |>
#   setNames(c("cc", "b", "a")) |> 
#   dplyr::mutate(
#     metab_id = metab_ids, 
#     metab_name = measure_tbl$metab_name[match(metab_id, measure_tbl$metab_id)]
#   )
# 
# # All combination of metabolites and concentrations across all range
# rg <- range(calcurve_to_show_df$conc, na.rm = TRUE)
# calcurve_smooth <- tidyr::expand_grid(
#   beta,
#   conc = exp(seq(log(rg[1]), log(rg[2]), length.out = 100))
# ) |> 
#   dplyr::mutate(
#     loess_norm = a * (conc^2) + b * conc + cc,
#   ) |> 
#   dplyr::filter(loess_norm > 0)

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  # geom_line(data = calcurve_smooth, color = "cadetblue", alpha = 0.5) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

### `NormSample` and `Groundwater` samples

The data of the two classes are presented on the calibration curve.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11
to_incl <- measurements_se$Class %in% c("NormSample", "Groundwater")
measure_tbl <- tibble::as_tibble(measurements_se[i_sel, to_incl])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  fit_layer +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  # geom_line(data = calcurve_smooth, color = "cadetblue", alpha = 0.5) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ metab_name, scales = "free", ncol = 3)
```

# Concentration

The concentration of the samples has been computed using the calibration curve in the [section above](#calibration).
The distribution of the concentration of the samples is shown below.

```{r}
#| label: 'concentration-distribution-by-sample'
#| fig-width: 12
ggplot(measurements_se) +
  geom_boxplot(aes(x = sample_name, y = conc, fill = Class)) +
  labs(
    title = "Concentration distribution by class", 
    x = "Sample ID",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```

```{r}
#| label: 'concentration-distribution-by-metabolite'
#| fig-width: 12
#| fig-height: 20
ggplot(measurements_se) +
  geom_boxplot(aes(x = metab_name, y = conc)) +
  labs(
    title = "Concentration distribution by metabolite", 
    x = "Metabolite",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() + 
  facet_grid(Class~.)
```

```{r}
#| label: 'write-concentration-table'
# Write the concentration table
write_concentration_tsv <- function(sumexp, file) {
  # Prepare the metabolite table
  metab_columns <- SummarizedExperiment::rowData(sumexp) |> 
    tibble::as_tibble() |> 
    dplyr::select(
      `Alignment ID` = alignment_id,
      `Metabolite name` = metab_name,
      `Average Mz` = mz,
      `Average Rt(min)` = rt,
    )
  # Prepare the concentration table
  conc_df <- SummarizedExperiment::assays(sumexp)$conc |> 
    round(3) |>     # Reduce the number of decimals
    tibble::as_tibble()
  # Original sample ID
  colnames(conc_df) <- SummarizedExperiment::colData(sumexp)$sample_name
  readr::write_tsv(dplyr::bind_cols(metab_columns, conc_df), file)
}
write_concentration_tsv(measurements_se, FILE$o)
```


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

