---
title: "Exploratory Data Analysis - Exposome"
tbl-cap-location: top
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[row_df, `row_df<-`, col_df, `col_df<-`, assays, `assays<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
)
# Input file
box::use(io = projlib/check_io_exist)       # Handle snakemake parameters and input/output files
FILE <- rlang::list2(
  i = "data/raw_internal/water_samples_original_ESipos_MSdial_output (not blank filtered).txt",
)
io$chq_all_files_exist(FILE)
set.seed(7579)
```

```{r}
#| label: 'read-given-tsv-and-save-sumexp'
#| purl: true

#' Identify three column sections in the MS-DIAL output file
#' 
#' @param msdial_file Path to the MS-DIAL output file
#' @return A list of three column indices
get_three_section_indices <- function(msdial_file) {
  # Read the second (`File type`) line to identify three sections
  file_type_line <- read.delim(
    msdial_file, header = FALSE, nrow = 5L, na.strings = c("NA", "")
  )[2, ]
  i_2nd <- which(! is.na(file_type_line))    # Not NA is the 2nd column sections
  stopifnot(file_type_line[i_2nd[1L]] == "File type")    # As expected
  list(
    "1st" = 1L:i_2nd[1L],
    "2nd" = (i_2nd[1L] + 1L):max(i_2nd),         # "+ 1L" for the title column of the rows
    "3rd" = (max(i_2nd) + 1L):length(file_type_line)
  )
}

#' Fetch sample information from the first 5 lines
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the columns that contain sample information
#' @return A tibble with sample information
fetch_sample_info <- function(msdial_file, indices) {
  if (missing(indices)) indices <- get_three_section_indices(msdial_file)[[2]]
  # Read sample information lines (first 5 lines)
  sample_lines <- read.delim(
    msdial_file, 
    header = FALSE,
    nrow = 5L,
    na.strings = c("NA", "")
  )
  header <- sample_lines[, indices[1L] - 1L]
  
  # Extract sample information  
  sinfo <- t(sample_lines[, indices, drop = FALSE])
  colnames(sinfo) <- header          # There was no header in `sinfo` before
  stopifnot("Deviation from the expected names for samples" = identical(
    colnames(sinfo)[-5], c("Class", "File type", "Injection order", "Batch ID")
  ))
  colnames(sinfo)[5] <- "sample_name"
  # Create a tibble with the sample information
  sinfo <- tibble::as_tibble(sinfo) |> 
    dplyr::select(
      Class,
      sample_type = "File type",
      injection_order = "Injection order",
      sample_name,
    ) |> 
    # Syntactically valid for variable name in R
    dplyr::mutate(sample_id = make.names(sample_name, unique = TRUE), .before = 1L) |>
    # Control sample groups
    dplyr::mutate(
      contr_cat = dplyr::case_when(
        sample_type == "Standard" & Class == "CalCurve" ~ "CalCurve",
        sample_type == "QC"    ~ "QC",
        sample_type == "Blank" ~ "Blank",
        TRUE ~ ""         # # NA does not behave predictably with `==`
      )
    )
  stopifnot(
    "Calibration curve samples are required." = any(sinfo$contr_cat == "CalCurve"),
    "QC samples are required." = any(sinfo$contr_cat == "QC"),
    "Blank samples are required." = any(sinfo$contr_cat == "Blank")
  )
  
  # Function to find concentration values from the given IDs
  find_concentration <- function(sid) {
    sid |> 
      stringr::str_extract("Cal_([[:digit:]-]+)", group = 1) |> 
      stringr::str_replace("-", ".") |>     # Replace "-" with "."
      as.numeric()
  }
  sinfo <- sinfo |> 
    dplyr::mutate(
      c_conc = ifelse(contr_cat == "CalCurve", find_concentration(sample_name), NA_real_),
      injection_order = as.integer(injection_order),
    )
  stopifnot(
    "Error in Calibration sample IDs" = all(! is.na(sinfo$c_conc[sinfo$contr_cat == "CalCurve"]))
  )
  return(sinfo)
}

#' Fetch data of the selected columns of the MS-DIAL output file
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the selected columns
#' @return A tibble 
fetch_data_of_columns <- function(msdial_file, indices) {
  stopifnot(!missing(indices))
  # Count total number of columns in the file
  l1 <- readr::read_tsv(msdial_file, skip = 4L, n_max = 1L, col_names = FALSE, col_types = "c")
  coltypes <- rep("-", ncol(l1))      # Skip the rest columns
  coltypes[indices] <- "c"
  # Read lines that contain feature data
  readr::read_tsv(
    msdial_file, 
    skip = 4L,      # Data starts from the 5th line
    col_types = paste0(coltypes, collapse = ""),
    name_repair = "unique_quiet"
  )
}

# Parse the data in the file into SummarizedExperiment
parse_to_sumexp <- function(msdial_file) {
  i_three_sections <- get_three_section_indices(msdial_file)
  sample_info <- fetch_sample_info(msdial_file, i_three_sections[[2]])
  
  # Read feature data in the first section
  features <- fetch_data_of_columns(msdial_file, i_three_sections[[1]]) |>
    dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
      alignment_id = "Alignment ID",
      feature_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      sn_ratio = "S/N average",
      std_type = "Comment",            # Type of standard, e.g. "Quant", "IS", or NA 
    ) |> 
    # `feature_id` = Syntactically valid ID
    dplyr::mutate(feature_id = make.names(feature_name, unique = TRUE), .before = 1L) |> 
    dplyr::mutate(
      dplyr::across(c(mz, rt, sn_ratio), as.numeric),
      std_type = dplyr::case_when(
        std_type == "Quant" ~ "Quant",
        std_type == "IS"    ~ "IS",
        TRUE ~ ""              # Ignore other variants
      )
    )
  
  # Measured values of the features into a matrix
  raw_df <- fetch_data_of_columns(msdial_file, i_three_sections[[2]])
  stopifnot(identical(colnames(raw_df), sample_info$sample_name))
  colnames(raw_df) <- sample_info$sample_id     # Update with syntactically valid names
  raw_mat <- lapply(raw_df, as.numeric) |> 
    as.data.frame(row.names = features$feature_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(raw = raw_mat),
    col_df = sample_info,
    row_df = features,
  )
}
overall_sumexp <- parse_to_sumexp(FILE$i)  # To be updated throughout the pre-processing
```

Given file: `` `r FILE$i` `` 

The given file contains data of `r nrow(overall_sumexp)` features and `r ncol(overall_sumexp)` samples.

## Samples

### By class

There are `r dplyr::n_distinct(SumExp::col_df(overall_sumexp)$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
table(SumExp::col_df(overall_sumexp)$Class) |> 
  knitr::kable(col.names = c("Class", "Number of Samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

### Special categories for preprocessing

There are `r length(table(SumExp::col_df(overall_sumexp)$contr_cat, exclude = ""))` special sample categories for in the data.
The data of those are important in the following processing steps.
The number of samples for each category is as follows:

```{r}
table(SumExp::col_df(overall_sumexp)$contr_cat, exclude = "") |> 
  knitr::kable(col.names = c("Category", "Number of Samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

#### Calibration curve samples

The samples for the calibration curve are identified by the sample type `Standard` and class `CalCurve`.
The concentration is extracted from the provided sample ID.
The IDs are in the format `Cal_[[:digit:]-]+` with optional numbers and/or characters before and after of it. 
Concentration is extracted from the numeric part in the format after the replacement of `-` with `.`.

##### Sample IDs

```{r}
# Present the sample IDs for the calibration curve
df_show <- col_df(overall_sumexp[, SumExp::col_df(overall_sumexp)$contr_cat == "CalCurve"]) |> 
  tibble::as_tibble() |> 
  dplyr::arrange(c_conc)
df_show |> 
  dplyr::select(sample_name, c_conc) |>
  knitr::kable(
    col.names = c("Sample ID", "Concentration (ng/ml)"),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

##### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
table(df_show$c_conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Quality control samples

```{r}
#| purl: true
# Extract QC samples into list
extract_qc_samples_to_list <- function(se) {
  se <- se[, se$contr_cat == "QC"]
  qc_id <- se$Class
  # Column-wise split
  lapply(setNames(nm = unique(qc_id)), \(ii) se[, qc_id == ii])
}
```

Aliquots of some samples were measured as technical replicates for quality control.
The number of the replicates were as below.

```{r}
extract_qc_samples_to_list(overall_sumexp) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Sample ID", "Number of Replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Features

The data contained `r nrow(overall_sumexp)` features.
Among them, `r sum(SumExp::row_df(overall_sumexp)$std_type != "")` features were targeted. 
The rest were not targeted.

```{r}
#| purl: true
# Extract standard features
is_targeted <- SumExp::row_df(overall_sumexp)$std_type != ""
targeted_features_se <- overall_sumexp[is_targeted, ]
```

For those targeted features, the standard type was specified.
The number of features for each type is as follows:

```{r}
table(SumExp::row_df(targeted_features_se)$std_type) |> 
  knitr::kable(col.names = c("Standard Type", "Number of Features")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

### Overall samples and targeted features

#### Linear scale

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 8 
#| out-width: '60%'
p <- ggplot(targeted_features_se) +
  geom_histogram(aes(x = raw), bins = 50) +
  labs(title = "Distribution of the values", x = "Values")
print(p)
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-overall-log'
#| fig-width: 8
#| out-width: '60%'
p + scale_x_log10()
```

### By sample class and overall targeted features

```{r}
#| label: 'distribution-values-by-class'
ggplot(targeted_features_se) +
  geom_histogram(aes(x = raw), bins = 30) +
  labs(title = "Distribution of the values by sample class", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Class, scales = "free_y", ncol = 4)
```

### By feature and overall samples

Randomly selected 8 targeted features are shown in the following figure.

```{r}
#| purl: true
#' Randomly select `n` rows from `sumexp` with optional `subset`
#'
#' @param sumexp a SumExp object
#' @param n the number of rows to select
#' @param subset an optional subset expression to apply to the `row_df` of `sumexp`
#' @param keep_the_order if `TRUE`, the selected rows are sorted in the previous order
#'
#' @return a SumExp object with `n` rows randomly selected from `sumexp`
#' @export
select_rows_randomly <- function(sumexp, n, subset, keep_the_order = FALSE) {
  ii <- c(1:nrow(sumexp))
  if (!missing(subset)) {      # If given, apply it to the `row_df` of `sumexp`
    ii <- ii[eval(substitute(subset), as.data.frame(SumExp::row_df(sumexp)))]
  }
  ii <- sample(ii, n)
  if (keep_the_order) ii <- sort(ii)
  sumexp[ii, ]
}
```

```{r}
#| label: 'distribution-values-by-feature'
se <- select_rows_randomly(targeted_features_se, 8, std_type != "IS")
ggplot(se) +
  geom_histogram(aes(x = raw, fill = Class), bins = 30) +
  labs(title = "Distribution of the values by feature", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ feature_name, scales = "free_x", ncol = 4)
```

----------------------------

# Normalization using internal standards

## Internal standard (IS) features

```{r}
#| purl: true
#' Calculate RSD%
#'
#' @param x A numeric vector
#' @param digits An integer for rounding
#' @return A numeric value of RSD%
rsd_perc <- function(x, digits = 2) round(100 * sd(x) / mean(x), digits)

internal_std_se <- local({
  se <- overall_sumexp
  # Extract the internal standard features
  se <- se[SumExp::row_df(se)$std_type == "IS", ]
  # Sort by average retention time
  se <- se[order(SumExp::row_df(se)$rt), ]
  # Number of zeros
  SumExp::row_df(se)$num_zeros <- rowSums(assays(se)$raw == 0)
  # RSD% of the internal standard features across all samples
  SumExp::row_df(se)$rsd <- apply(assays(se)$raw, 1, rsd_perc)
  return(se)
})
```

There are `r nrow(internal_std_se)` internal standard (IS) features in the data.
Those features are listed below, in the order of average retention time.

```{r}
#| purl: true
# Table of internal standard features
kable_internal_std <- function(se) {
  df <- SumExp::row_df(se)[, c("feature_name", "rt", "mz", "sn_ratio", "rsd", "num_zeros"), drop = FALSE]
  df$num_zeros <- ifelse(df$num_zeros == 0, "", df$num_zeros)
  knitr::kable(
    df, 
    row.names = FALSE,
    col.names = c("Feature Name", "Avg Retention Time (min)", "m/z", "S/N Ratio", 
                  "RSD%", "Number of zeros"),
    align = "lrrrrr"
  ) |> 
    kableExtra::kable_styling(full_width = FALSE)
}
kable_internal_std(internal_std_se)
```

### Outliers

Samples deviating too much from others may be considered as outliers based on the values of internal standard (IS) features, which were supposed to be the same across all the samples. 
For each internal standard, the samples with values greater than 3 standard deviations from the mean of all samples after log-transformation ($log(1 + x)$) were marked as outlying samples for the feature.
In the other way, the feature was marked as an outlier for the sample.
When the number of outlying IS features of a sample exceeds 20% of the total number of IS, the sample was considered as outliers and eventually excluded from the data.

```{r}
#| purl: true
# Identify outliers
identify_outliers <- function(x) {
  m <- mean(x)
  sd <- sd(x)
  return((x > m + 3 * sd) | (x < m - 3 * sd))
}
# Number of outlying internal standard features per sample 
n_out_istd <- assays(internal_std_se)$raw |> 
  log1p() |>              # Log-transform the data
  apply(1, identify_outliers) |>
  rowSums()      # Transposed by `apply` above
is_outlier <- n_out_istd > nrow(internal_std_se) * 0.2
if (all(! is_outlier)) {
  # Exclude the outlying samples
  stopifnot(identical(names(is_outlier), colnames(internal_std_se)))
  internal_std_se <- internal_std_se[, !is_outlier]
  stopifnot(identical(names(is_outlier), colnames(overall_sumexp)))
  overall_sumexp <- overall_sumexp[, !is_outlier]
}
```

```{r}
#| label: 'outliers-table'

if (all(! is_outlier)) {
  cat("No outliers were identified.\n")
} else {
  cat("The following samples were identified as outliers.\n\n")
  stopifnot(identical(names(is_outlier), rownames(n_out_istd)))
  tibble::as_tibble(n_out_istd, rownames = "sample_id") |> 
    dplyr::mutate(is_outlier = is_outlier) |>
    dplyr::left_join(as.data.frame(col_df(overall_sumexp)), by = "sample_id") |> 
    dplyr::filter(is_outlier) |> 
    dplyr::select(
      "Sample ID" = sample_name,
      Class,
      "File type" = sample_type,
      "Number of Outlying IS" = value
    ) |> 
    knitr::kable(align = "lccr") |> 
    kableExtra::kable_classic(full_width = FALSE)
  
  # Update the table of internal standard features
  cat("\n\nAfter excluding the outliers,", 
      "the table of internal standard features is as follows.\n\n")
  # Number of zeros
  SumExp::row_df(internal_std_se)$num_zeros <- rowSums(assays(internal_std_se)$raw == 0)
  # RSD% of the internal standard features across all samples
  SumExp::row_df(internal_std_se)$rsd <- apply(assays(internal_std_se)$raw, 1, rsd_perc)
  # Updated table of internal standard features
  kable_internal_std(internal_std_se)
}
```

### Failed internal standards

Internal standard (IS) features that satisfy the following conditions were considered as failed IS:

* One or more samples have zero values.

```{r}
#| label: 'failed-internal-standards'
#| purl: true
is_failed_istd <- SumExp::row_df(internal_std_se)$num_zeros > 0
```

```{r}
#| out-width: 100%
box_scatter_plot_of_one_feature <- function(se, ii, widths) {
  feature_name <- SumExp::row_df(se)$feature_name[ii]
  se_tbl <- tibble::as_tibble(se[ii, ])
  # Box plot
  p1 <- ggplot(se_tbl) +
    geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
    labs(title = feature_name, y = "Peak area") +
    theme(legend.position = "none") +   # No legend for fill
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Rotate x-axis labels
    scale_y_log10()
  # Scatter plot across injection order
  p2 <- ggplot(se_tbl) +
    geom_point(aes(x = injection_order, y = raw, color = Class)) +
    labs(title = "",  y = "", x = "Injection order") +
    theme(legend.position = "none") +   # No legend for fill
    # No y-axis labels
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_log10()
  # Requires patchwork package
  p1 + p2 + patchwork::plot_layout(widths = widths)
}

if (all(! is_failed_istd)) {
  cat("No failed internal standard features were identified.\n")
} else {
  cat("The following internal standard features were identified as failed IS,",
      "which were removed.\n\n")
  failed <- internal_std_se[is_failed_istd, ]
  # Table of failed internal standard features
  print(kable_internal_std(failed))
  # Box and scatter plots of the failed internal standard features 
  for(ii in 1:nrow(failed)) {
    print(box_scatter_plot_of_one_feature(failed, ii, widths = c(0.3, 0.7)))
  }
}
```

```{r}
#| purl: true
# Remove failed internal standard features
internal_std_se <- internal_std_se[!is_failed_istd, ]
```

### Distribution of the values of individual IS features

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 4
#| out-width: 100%
# Internal standard features to show
se_show <- internal_std_se
for(ii in 1:nrow(se_show)) {
  print(box_scatter_plot_of_one_feature(se_show, ii, widths = c(0.3, 0.7)))
}
```

### Raw RSD%

The relative standard deviation (RSD%) of the raw values of the internal standard features are calculated.

#### `QC` samples

As these `QC` samples are technical replicates, the RSD% values of the samples are expected to be low.

```{r}
#| label: 'rsd-internal-standards-by-qc-samples'
# Divide the internal standard features by quality control sample
# Calculate RSD% per internal standard feature and per quality control sample
df1 <- extract_qc_samples_to_list(internal_std_se) |> 
  sapply(\(se) apply(assays(se)$raw, 1, \(.x) round(rsd_perc(.x), 2)))
stopifnot(identical(rownames(df1), rownames(internal_std_se)))
rownames(df1) <- SumExp::row_df(internal_std_se)$feature_name
knitr::kable(df1, caption = "RSD% of internal standard features by quality control sample") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### By sample classes

```{r}
se <- internal_std_se
se <- se[, se$contr_cat != "QC"]
df1 <- sapply(setNames(nm = unique(se$Class)), function(cls) {
  se <- se[, se$Class == cls]
  apply(assays(se)$raw, 1, \(.x) round(rsd_perc(.x), 2))
})
stopifnot(identical(rownames(df1), rownames(se)))
rownames(df1) <- SumExp::row_df(se)$feature_name
knitr::kable(df1, caption = "RSD% of internal standard features by sample classes") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Normalization

The data is normalized using the internal standard features.

```{r}
#| purl: true
# Find the range of RT values
overall_rt_range <- range(SumExp::row_df(overall_sumexp)$rt)
targetted_rt_range <- range(SumExp::row_df(targeted_features_se)$rt)
rt <- SumExp::row_df(internal_std_se)$rt
```

The range of RT values of overall features is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of targeted features is `r targetted_rt_range[1]` to `r targetted_rt_range[2]` and 
the range of internal standard features is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard features

The closest internal standard features to each feature will be identified based on the retention time.
The measured peak area of each feature will be divided by the peak area of the closest internal standard feature.

```{r}
#| label: 'normalize-rt-closest-istd'
#| purl: true
# Get the values of the closest internal standard features
get_raw_of_closest_istd <- function(se_x, internal_std_se) {
  rt_x <- SumExp::row_df(se_x)$rt
  rt <- SumExp::row_df(internal_std_se)$rt
  # Find the closest internal standard feature
  i_closest <- sapply(rt_x, \(.x) which.min(abs(rt - .x)))
  out <- assays(internal_std_se)$raw[i_closest, ]
  rownames(out) <- rownames(se_x)
  out
}
closest_istd_raw <- get_raw_of_closest_istd(overall_sumexp, internal_std_se)

# Normalize the data using closest internal standard features
assays(overall_sumexp)$closest_norm <- assays(overall_sumexp)$raw / closest_istd_raw
```

#### Evaluation of normalization

```{r}
#| purl: true

# Extract quantitative standards in QC samples
extract_quant_qc <- function(sumexp) {
  se_lst <- extract_qc_samples_to_list(sumexp)
  lapply(se_lst, \(se) se[SumExp::row_df(se)$std_type == "Quant", ])
}
# Calculate RSD of the quantification standard samples
calc_rsd_qstd <- function(qc_se, assay_ids) {
  lapply(qc_se, \(qc1) {
    sapply(setNames(nm = assay_ids), function(nm) {
      apply(assays(qc1)[[nm]], 1, rsd_perc)
    }) |>    # Rows are features, columns are assay IDs
      tibble::as_tibble(rownames = "feature_id")
  }) |> 
    # `QC` has the name of the QC samples
    dplyr::bind_rows(.id = "QC")
}
# Plot the RSD% of various data sets of features
ggplot_rsdp_metab <- function(rsd_df, assay_ids) {
  nms <- names(assay_ids)
  ids <- unname(assay_ids)    # Duplicated by `dplyr::all_of`
  rsd_df <- rsd_df |> 
    dplyr::mutate(
      # NA was generated when all values were zero
      dplyr::across(dplyr::all_of(ids), \(.x) ifelse(is.na(.x), 0, .x)),
    ) |> 
    tidyr::pivot_longer(dplyr::all_of(ids), names_to = "Data", values_to = "RSD%") |> 
    dplyr::mutate(Data = factor(Data, levels = ids, labels = nms))
  ggplot(rsd_df) +
    aes(x = Data, y = `RSD%`) +
    geom_line(aes(group = feature_id), alpha = 0.5)
}
```
```{r}
#| label: 'evaluation-normalization-closest-istd'
#| out-width: '80%'

# Quantification standard samples
qc_se <- extract_quant_qc(overall_sumexp)

assay_ids <- setNames(c("raw", "closest_norm"), nm = c("Raw", "Closest IS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > closest_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'evaluation-normalization-closest-istd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```


### LOESS fit along RT

```{r}
#| label: 'normalize-rt'
#| purl: true

excluded_category_for_normalization <- c("Blank", "CalCurve")

loess_fit <- local({
  # Log-transform the data
  istd_log <- log(assays(internal_std_se)$raw)
  rt <- SumExp::row_df(internal_std_se)$rt
  
  # Normalize the data using the internal standards
  # Mean of each internal standard feature for overall measurement samples
  # , excluding the calibration curve and blank samples
  to_excl <- internal_std_se$contr_cat %in% excluded_category_for_normalization
  # The data of the samples with measurements
  mat <- istd_log[, !to_excl, drop = FALSE]
  stopifnot("Failed interncal control(s)! value = 0" = all(is.finite(mat)))
  istd_m <- rowMeans(mat)
  # Subtract the mean from the data
  fr_m_log <- istd_log - istd_m
  
  # Expand to smallest RT 
  istd_smallest_rt <- fr_m_log[which.min(rt), ]
  # Expand to largest RT 
  istd_largest_rt <- fr_m_log[which.max(rt), ]
  # Add the expanded range
  fr_m_log <- rbind(
    istd_smallest_rt, 
    istd_largest_rt,
    fr_m_log
  )
  
  # LOESS fit of the internal standard features along RT
  loess_fit <- purrr::map(1:ncol(fr_m_log), function(ii) {
    # Fit a loess curve
    y <- fr_m_log[, ii]
    y[is.finite(y) == FALSE] <- NA
    stopifnot(identical(
      names(y), c("istd_smallest_rt", "istd_largest_rt", rownames(internal_std_se))
    ))
    suppressWarnings(loess(y ~ c(overall_rt_range, rt), span = 1))
  })
  names(loess_fit) <- colnames(fr_m_log)
  loess_fit
})

# Normalize the data by LOESS fit along RT
rt <- SumExp::row_df(overall_sumexp)$rt
raw <- assays(overall_sumexp)$raw
assays(overall_sumexp)$loess_norm <- sapply(
  colnames(overall_sumexp), function(sample_id) {
    norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
    exp(log(raw[, sample_id]) - norm_factor)
  }
)
```

For each feature, the average of all samples excluding 
`r knitr::combine_words(excluded_category_for_normalization, before = "\x60")`
samples was calculated after log-transformation.
The deviation of each sample from the average was calculated as normalization factor.
The data of the sample are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) features along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all features. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all features.
The normalized data are calculated by subtracting the normalization factor from the log-transformed data.

Randomly selected 3 samples are shown below.
```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 12

to_excl <- SumExp::col_df(overall_sumexp)$contr_cat %in% excluded_category_for_normalization
# Plot the normalized and original data for randomly selected 3 samples
i_sel <- sample(c(1:ncol(overall_sumexp))[!to_excl], 3)

# To include LOESS smooth line
x <- seq(overall_rt_range[1], overall_rt_range[2], length.out = 100)
loess_curve_df <- lapply(setNames(i_sel, colnames(overall_sumexp)[i_sel]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(rt = x) |>
  tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")
loess_curve_df$sample_name <-
  SumExp::col_df(overall_sumexp)$sample_name[match(loess_curve_df$sample_id, SumExp::col_df(overall_sumexp)$sample_id)]

# Log-transform the data
raw_log <- log(assays(overall_sumexp)$raw)
# Overall mean of every feature
mat <- raw_log[, !to_excl, drop = FALSE]
mat[is.finite(mat) == FALSE] <- NA    # Visualization purpose only
mean_log <- rowMeans(mat, na.rm = TRUE)
# Deviation from the mean
selected_se <- overall_sumexp[, i_sel]
assays(selected_se) <- assays(selected_se) |> 
  lapply(\(x) sweep(log(x), 1, mean_log, FUN = "-"))

df_to_show <- tibble::as_tibble(selected_se) |>
  dplyr::rename(Raw = raw, Normalized = loess_norm, sample_name = sample_name) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(
    Type = ifelse(std_type == "", "non-target", std_type),
    Type = factor(Type, levels = c("IS", "Quant", "non-target"))
  ) |> 
  dplyr::arrange(desc(Type)) # To show IS on top

# Plot the normalized and original data
ggplot(df_to_show, aes(x = rt, y = Value)) + 
  geom_point(aes(color = Data, shape = Type, size = Type, alpha = Type)) +
  geom_line(data = loess_curve_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values) - log(avg)"
  ) +
  scale_shape_manual(values = c("IS" = 15, "Quant" = 21, "non-target" = 20)) +
  scale_size_manual(values = c("IS" = 3, "Quant" = 3, "non-target" = 1)) +
  scale_alpha_manual(values = c("IS" = 1, "Quant" = 1, "non-target" = 0.1)) +
  facet_wrap(~ sample_name, scales = "free_y", ncol = 3)
```

The line in the graph represents the LOESS fit.
The points at the bottom boundary are the **0**s, which were not normalized by this process.
Those zeros were kept as zeros in the normalized data.

#### Evaluation of normalization

The normalized data are evaluated by the changes of relative standard deviation (RSD) of the quantification standard features of the QC samples.
```{r}
#| label: 'rsd-eval'
#| out-width: '80%'

# Quantification standard features of QC samples
qc_se <- extract_quant_qc(overall_sumexp)
# Calculate RSD of the quantification standard samples
assay_ids <- setNames(c("raw", "loess_norm"), nm = c("Raw", "LOESS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > loess_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'rsd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

Bar charts of the values of standard features are compared for raw and normalized data.

```{r}
#| fig-width: 14
#| fig-height: 6
#| out-width: '100%'

se <- overall_sumexp
df_show <- se[SumExp::row_df(se)$std_type %in% c("Quant", "IS"), ] |> 
  tibble::as_tibble() |> 
  tidyr::pivot_longer(cols = c(raw, loess_norm), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(
    Data = factor(Data, levels = c("raw", "loess_norm"), labels = c("Raw", "LOESS Normalized"))
  )
df_show |> 
  dplyr::filter(std_type == "Quant", contr_cat == "") |> 
  ggplot() +
  geom_col(aes(x = injection_order, y = Value, fill = Class)) +
  labs(title = "Quantification standards", x = "Injection order") +
  facet_wrap(~ Data, scales = "free_y")

df1 <- df_show |> 
  dplyr::filter(std_type == "IS") |> 
  dplyr::mutate(sum_value = sum(Value, na.rm = TRUE), .by = c(injection_order, Data))
ggplot(df1, aes(x = injection_order)) +
  geom_col(aes(y = Value, fill = Class)) +
  labs(title = "Internal standards", x = "Injection order") +
  geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
  facet_wrap(~ Data, scales = "free_y")

ggplot(df1, aes(x = injection_order)) +
  geom_col(aes(y = Value, fill = feature_name)) +
  labs(title = "Internal standards", x = "Injection order") +
  geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
  facet_wrap(~ Data, scales = "free_y")
```

### Comparison of the normalization methods

The RSD% of the quantification standard samples is calculated for the raw data and the normalized data.

```{r}
#| label: 'norm-method-comparison'
#| out-width: '80%'

assay_ids <- setNames(c("raw", "closest_norm", "loess_norm"), nm = c("Raw", "Closest", "LOESS"))
df_to_show <- extract_quant_qc(overall_sumexp) |> 
  calc_rsd_qstd(assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(loess_norm <= closest_norm, "↓ LOESS", "↓ Closest"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓ LOESS", "↓ Closest", "All 0s"))
  ) 
ggplot_rsdp_metab(df_to_show, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with lower RSD after each normalization is counted below.

```{r}
#| label: 'rsd-comparison-count'
#| results: 'markup'
# Count the number of features with lower RSD
df_to_show |> 
  janitor::tabyl(QC, sign)
```

The number of compounds above or below 20 RSD% is counted below.

```{r}
#| label: 'rsd-comparison-count-20'
df_to_show |> 
  dplyr::summarize(
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x > 20, na.rm = TRUE), 
      .names = "{.col}_gt_20"
    ),
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x <= 20, na.rm = TRUE), 
      .names = "{.col}_eq_lt_20"
    ),
    .by = QC
  ) |>  
  knitr::kable(
    caption = "Number of features with RSD% above or below 20", 
    col.names = c("QC", rep(names(assay_ids), 2))
  ) |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "RSD > 20%" = 3, "RSD ≤ 20%" = 3)) 
```

The RSD% values are presented in the table below.

```{r}
#| label: 'rsd-comparison-table'
df_to_show |> 
  dplyr::left_join(as.data.frame(SumExp::row_df(overall_sumexp)), by = "feature_id") |>   # Add feature names
  dplyr::select(QC, feature_name, dplyr::all_of(unname(assay_ids))) |> 
  dplyr::rename(all_of(assay_ids), `Feature` = feature_name) |> 
  knitr::kable(caption = "RSD% of the quantification standard samples", digits = 2) |> 
  kableExtra::kable_classic(bootstrap_options = "striped", full_width = FALSE) |>
  kableExtra::scroll_box(height = "300px")
```

```{r}
#| label: 'write-loess-normalized-data'
#| purl: true

# Write the LOESS normalized data
write_normalized_data_tsv <- function(sumexp, in_file, out_file) {
  # Prepare the feature table
  i_sec <- get_three_section_indices(in_file)
  intact_feature_cols <- fetch_data_of_columns(in_file, i_sec[[1]])
  stopifnot(identical(intact_feature_cols$"Alignment ID", SumExp::row_df(sumexp)$alignment_id))
  # Prepare the normalized data table
  norm_df <- assays(sumexp)$loess_norm |> 
    round(0) |>     # Reduce the number of decimals
    tibble::as_tibble()
  # Original sample ID
  colnames(norm_df) <- col_df(sumexp)$sample_name
  readr::write_tsv(dplyr::bind_cols(intact_feature_cols, norm_df), out_file)
}
write_normalized_data_tsv(overall_sumexp, FILE$i, FILE$o$norm)
```

# Calibration

## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The samples with a concentration of `0` are separated as the zero samples.
The data of the zeros is used to calculate lower-limits of quantification (LLOQ).

```{r}
#| label: 'calcurve-samples-extraction'
#| purl: true
# Extract quantification standard data
quant_se <- overall_sumexp[SumExp::row_df(overall_sumexp)$std_type == "Quant", ]
# Calibration curve samples only
calcurve_se <- quant_se[, quant_se$contr_cat == "CalCurve"]
stopifnot("Calibration curve samples are required." = nrow(calcurve_se) > 0)
# Calibration zero samples
cal_0_se <- calcurve_se[, calcurve_se$c_conc == 0] 
stopifnot("Multiple `Cal_0` samples are required." = ncol(cal_0_se) > 1)
# Remove the 0 concentration samples
calcurve_se <- calcurve_se[, calcurve_se$c_conc != 0]
```

The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
table(calcurve_se$c_conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Concentration ranges

Quite a few concentrations of calibration curve samples of a compound may be too far from the range of the measured samples. 
Hence the concentrations were excluded for which the peak areas after normalization were more than 10 times higher than the **maximum** peak area of the samples excluding `CalCurve` and `QC` categories.

```{r}
#| label: 'calcurve-upper-limit-concentration-range'
#| purl: true

SumExp::row_df(calcurve_se)$max_conc <- local({
  stopifnot(rownames(quant_se) == rownames(calcurve_se))
  se1 <- quant_se
  se1 <- se1[, se1$contr_cat %in% c("CalCurve", "QC") == FALSE]
  # Find maximum peak area of the samples of the classes for measurement
  max_loess_norm <- apply(assays(se1)$loess_norm, 1, max, na.rm = TRUE)
  
  cc <- calcurve_se
  # Find the maximum concentration of the samples that are not too far (10x) from the measured
  apply(assays(cc)$loess_norm <= 10 * max_loess_norm, 1, \(ea_row) {
    if (all(!ea_row)) return(NA_real_)
    max(cc$c_conc[ea_row])
  })
})
```

When the obtained peak area of low concentration of the calibration samples is too low, the concentrations are excluded.
The value is considered **too low** when the average of the values of one concentration is less than 10 times of the standard deviation from the mean of zero concentration samples. 

```{r}
#| label: 'calcurve-lower-limit-concentration-range'
#| purl: true

# Function: find the mean + `times` of standard deviation 
get_mean_plus_sd <- function(mat, times) {
  m <- apply(mat, 1, mean, na.rm = TRUE)
  s <- apply(mat, 1, sd, na.rm = TRUE)
  m + times * s
}
# Find the lower-limit of quantification using values of the zero concentration samples
SumExp::row_df(calcurve_se)$lloq <- get_mean_plus_sd(assays(cal_0_se)$loess_norm, times = 10)
SumExp::row_df(calcurve_se)$min_conc <- local({
  cc <- calcurve_se
  # Find the minimum concentration of the calibration samples
  apply(assays(cc)$loess_norm > SumExp::row_df(cc)$lloq, 1, \(ea_row) {
    if (all(!ea_row)) return(NA_real_)
    min(cc$c_conc[ea_row])
  })
})

# Mark the features that have the minimum concentration equal to or greater than the maximum concentration
SumExp::row_df(calcurve_se)$to_exclude <- as.data.frame(SumExp::row_df(calcurve_se)) |> 
  with(is.na(min_conc) | is.na(max_conc) | min_conc >= max_conc)
```

When the minimum concentration is equal to or greater than maximum concentration, the feature is excluded from the calibration curve.
`r english::Words(sum(SumExp::row_df(calcurve_se)$to_exclude))` features are excluded from the calibration curve.

Remaining concentration range of each feature is presented below.

```{r}
#| label: 'calcurve-conc-range-table'

SumExp::row_df(calcurve_se) |> 
  tibble::as_tibble() |> 
  dplyr::mutate(
    min_conc = dplyr::case_when(
      is.na(min_conc) ~ paste0("All < LLOQ (", scales::scientific(lloq, 2), ")"), 
      is.na(max_conc) ~ "",
      TRUE ~ scales::number(min_conc)
    ),
    max_conc = ifelse(is.na(max_conc), "All 0s, no valid", scales::number(max_conc)),
    to_exclude = ifelse(to_exclude, "Excluded", "")
  ) |>
  dplyr::select(feature_name, min_conc, max_conc, to_exclude) |>
  knitr::kable(
    caption = "Concentration range of the calibration curve samples",
    col.names = c("Feature",
                  "Minimum (ng/ml)",
                  "Maximum concentration (ng/ml)",
                  "Excluded?"), 
    align = c("l", "r", "r", "r"),
  ) |> 
  kableExtra::kable_classic(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px")
```

```{r}
#| label: 'calcurve-exclude-data-by-concentration-range'
#| purl: true

# Exclude the marked features
is_marked <- SumExp::row_df(calcurve_se)$to_exclude
stopifnot(rownames(calcurve_se) == rownames(quant_se))
calcurve_se <- calcurve_se[is_marked == FALSE, ]
quant_se <- quant_se[is_marked == FALSE, ]
SumExp::row_df(quant_se)$lloq <- SumExp::row_df(calcurve_se)$lloq    # Copy the LLOQ values
# Exclude the values outside the concentration range
calcurve_se <- local({
  cc <- calcurve_se
  # Concentration values in a matrix, rows are features and columns are concentrations
  conc_mat <- matrix(rep(cc$c_conc, each = nrow(cc)), nrow = nrow(cc))
  to_na <- conc_mat < SumExp::row_df(cc)$min_conc | conc_mat > SumExp::row_df(cc)$max_conc
  assays(cc)$loess_norm[to_na] <- NA_real_
  cc
})
```

## Calibration curve per feature

The calibration curve of each of randomly selected 9 features is presented below.

```{r}
#| purl: true

set.seed(7579)
# Randomly selected 9 features to present in calibration curve plots
i_9feature_to_show <- local({
  ii <- c(1:nrow(calcurve_se))
  ii <- ii[SumExp::row_df(calcurve_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve_to_show_df0 <- tibble::as_tibble(calcurve_se[i_9feature_to_show, ])
calcurve_to_show_df <- calcurve_to_show_df0 |> 
  tidyr::drop_na(loess_norm)        # Drop concentrations outside the range
```

### Unprocessed data

The peak areas of the calibration samples are presented in the linear scale.
The presented data includes all the concentrations that were trimmed out.

```{r}
#| label: 'calcurve-per-feature-linear-before-norm'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df0, aes(x = c_conc, y = raw)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Raw)"
  ) +  
  facet_wrap(~ feature_name, scales = "free_y", ncol = 3)
```

### After normalization and narrowing the concentration range

The peak areas were normalized by the internal standards applying the LOESS normalization.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

#### Linear scale

```{r}
#| label: 'calcurve-per-feature-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
p <- ggplot(calcurve_to_show_df, aes(x = c_conc, y = loess_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ feature_name, scales = "free", ncol = 3)
print(p)
```

#### Log scale

```{r}
#| label: 'calcurve-per-feature-log'
#| fig-width: 12
#| fig-height: 11
p + scale_x_log10() + scale_y_log10()
```

### After normalization by closest IS in RT

```{r}
#| label: 'calcurve-per-feature-linear-closest'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = c_conc, y = closest_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ feature_name, scales = "free", ncol = 3)
```

## Calibration of the features

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.

```{r}
#| label: 'calibration-data'
#| purl: true

quadratic_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc + I(conc^2), weights = weights)
  beta <- coef(lmfit)
  a <- beta[["I(conc^2)"]]
  b <- beta[["conc"]]
  cc <- beta[["(Intercept)"]]
  out <- function(x) {
    det <- b ^ 2 - 4 * a * (cc - x)
    conc <- suppressWarnings((-b + sqrt(det)) / (2 * a))      # det < 0
    conc[conc < 0 | conc < 0] <- 0
    conc
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}
linear_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc, weights = weights)
  beta <- coef(lmfit)
  b1 <- beta[["conc"]]
  b0 <- beta[["(Intercept)"]]
  out <- function(x) {
    (x - b0) / b1
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}

build_calcurve_model <- function(conc, value) {
  if (all(is.na(value) | value == 0)) {
    return(list(
      "best_model" = function(x) x,
      "R2s" = NA,
      "best_model_name" = NA_character_
    ))
  }
  # Weight alternatives
  weights_alt <- rlang::list2(
    "1" = rep(1, length(conc)),
    "1/x" = 1 / conc,
    "1/x2" = 1 / (conc ^ 2),
  )
  # Linear and quadratic models
  lmodels <- lapply(weights_alt, \(w) linear_calcurve_model(conc, value, weights = w))
  names(lmodels) <- paste("linear", names(lmodels), sep = "-")
  qmodels <- lapply(weights_alt, \(w) quadratic_calcurve_model(conc, value, weights = w))
  names(qmodels) <- paste("quadratic", names(qmodels), sep = "-")
  models <- c(lmodels, qmodels)
  R2s <- sapply(models, \(f) attr(f, "R2"))
  list(
    "best_model" = models[[which.max(R2s)]],        # Best model by R2
    "R2s" = R2s,
    "best_model_name" = names(models)[which.max(R2s)]
  )
}

# Fit the calibration curve
to_excl <- rownames(calcurve_se)[SumExp::row_df(calcurve_se)$to_exclude]
calcurve_se <- calcurve_se[]
c_conc <- calcurve_se$c_conc
rt_norm_mat <- assays(calcurve_se)$loess_norm
calcurve_models <- lapply(setNames(nm = rownames(calcurve_se)), function(ii) {
  build_calcurve_model(c_conc, rt_norm_mat[ii, ])
})
```

The distribution of $R^2$ values by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'

R2s <- lapply(calcurve_models, \(m) m$R2s)
R2s <- R2s[!is.na(R2s)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(
    x = "Model",
    y = expression(R^2)
  )
```

The count of features favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-feature'
#| results: 'markup'
table(sapply(calcurve_models, \(m) m$best_model_name))
```

```{r}
#| label: 'concentration-computation'
#| purl: true
# Compute the concentration of the samples skipping the calibration curve samples
measurements_se <- quant_se[, ! quant_se$contr_cat %in% "CalCurve"]
SumExp::row_df(measurements_se)$c_conc <- NULL
assays(measurements_se)$conc <- SumExp::row_df(measurements_se)$feature_id |> 
  sapply(function(feature_id) {
    v <- assays(measurements_se)$loess_norm[feature_id, ]
    calcurve_models[[feature_id]]$best_model(v)
  }) |>
    t()          # Features to rows
# Negative concentrations are adjusted to zero
assays(measurements_se)$conc[assays(measurements_se)$conc < 0] <- 0
# Maximum/minimum concentration after trimming out of each feature
maxmin_conc <- SumExp::row_df(calcurve_se)[c("max_conc", "min_conc")]
stopifnot(rownames(maxmin_conc) == rownames(calcurve_se))
SumExp::row_df(measurements_se) <- cbind(SumExp::row_df(measurements_se), maxmin_conc)
```

The concentration of the sample classes, `r knitr::combine_words(unique(measurements_se$Class), before = "\x60")`,
are computed using the calibration curve.
Any negative concentration is adjusted to zero.

```{r}
#| purl: true

# Table of the samples to show in the calibration curve
data_to_show_tbl <- tibble::as_tibble(measurements_se[i_9feature_to_show, ]) |> 
  tidyr::drop_na(conc)
# Table about the selected calibration curve models
models_to_show_tbl <- purrr::imap_dfr(calcurve_models, \(m, feature_id) {
  tibble::tibble(
    feature_id = feature_id,
    best_model_name = m$best_model_name,
    R2 = max(m$R2s),
    model_r2 = paste0(best_model_name, "\nR2=", round(R2, 2)),
    
    feature_name = SumExp::row_df(calcurve_se)[feature_id, ]$feature_name,
    lloq = SumExp::row_df(calcurve_se)[feature_id, ]$lloq,
  )
}) |> 
  dplyr::slice(i_9feature_to_show)

# Plot the calibration curve for each feature with the samples in the calibration curve
ggplot_calcurve_samples_facet <- function(data) {
  ggplot(data, aes(x = conc, y = loess_norm)) +
    geom_point(aes(x = c_conc), data = calcurve_to_show_df) +     # Remaining calibration samples
    geom_point(aes(color = Class)) +                    # Samples to measure
    geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +     # Density of points at the edges of the plot
    labs(
      x = "Concentration (ng/ml)",
      y = "Peak area (Normalized)"
    ) +  
    facet_wrap(~ feature_name, scales = "free", ncol = 3) +
    geom_rect(         # Shade the region below the LLOQ
      aes(xmin = 0, xmax = Inf, ymin = 0, ymax = lloq),
      data = models_to_show_tbl,
      fill = "grey10",
      alpha = 0.5,
      inherit.aes = FALSE
    )
}
```

### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each feature.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  ggplot_calcurve_samples_facet()
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  ggplot_calcurve_samples_facet() +
  scale_x_log10() +
  scale_y_log10()
```

### All except `QC` and `Blank` samples

```{r}
classes_excl_qc_blank <- local({
  df <- col_df(measurements_se)
  unique(df$Class[!df$contr_cat %in% c("QC", "Blank")])
})
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each feature.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11
geom_best_model_name <- geom_text(         # Display the best model name
  aes(x = Inf, y = 0, label = model_r2),
  data = models_to_show_tbl,
  hjust = 1.1,
  vjust = 0,
  size = 4,
  color = "cadetblue"
)

data_to_show_tbl |> 
  dplyr::filter(Class %in% classes_excl_qc_blank) |> 
  ggplot_calcurve_samples_facet() +
  geom_best_model_name
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
data_to_show_tbl |> 
  dplyr::filter(Class %in% classes_excl_qc_blank) |> 
  ggplot_calcurve_samples_facet() +
  geom_best_model_name +
  scale_x_log10() +
  scale_y_log10()
```

# Concentration

The concentration of the samples has been computed using the calibration curve in the [section above](#calibration).
The distribution of the concentration of the samples is shown below.

```{r}
#| label: 'concentration-distribution-by-sample'
#| fig-width: 12
#| fig-height: 10
ggplot(measurements_se) +
  geom_boxplot(aes(x = sample_name, y = conc, fill = Class)) +
  labs(
    title = "Concentration distribution by sample", 
    x = "Sample ID",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```

```{r}
#| label: 'concentration-distribution-by-feature'
#| fig-width: 12
#| fig-height: 20
ggplot(measurements_se) +
  geom_boxplot(aes(x = feature_name, y = conc)) +
  labs(
    title = "Concentration distribution by feature", 
    x = "Feature",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() + 
  facet_grid(Class~.)
```

```{r}
#| label: 'write-concentration-table'
# Write the concentration table
write_concentration_tsv <- function(sumexp, file) {
  # Prepare the feature table
  feature_columns <- SumExp::row_df(sumexp) |> 
    tibble::as_tibble() |> 
    dplyr::select(
      `Alignment ID` = alignment_id,
      `Feature name` = feature_name,
      `Average Mz` = mz,
      `Average Rt(min)` = rt,
    )
  # Prepare the concentration table
  conc_df <- SummarizedExperiment::assays(sumexp)$conc |> 
    round(3) |>     # Reduce the number of decimals
    tibble::as_tibble()
  # Original sample ID
  colnames(conc_df) <- col_df(sumexp)$sample_name
  readr::write_tsv(dplyr::bind_cols(feature_columns, conc_df), file)
}
write_concentration_tsv(measurements_se, FILE$o$conc)
```


```{r}
#| echo: false
#| output: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

