---
params:
  input_file: "data/raw_internal/water_samples_original_ESipos_MSdial_output (not blank filtered).txt"
  output_dir: "data/processed"
title: "Exploratory Data Analysis - Exposome"
---

```{r}
#| label: 'setup-import'
#| purl: true

# Load packages and project local libraries
options(box.path = "code/")
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[rowData, `rowData<-`, assays, `assays<-`, assay, `assay<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
  io = projlib/check_io_exist,
)
# Input file
FILE <- rlang::list2(
  i = params$input_file,
)
# Output file
FILE$o <- local({
  b <- basename(FILE$i) |> 
    tools::file_path_sans_ext()    # Without extension
  file.path(params$output_dir, paste0(b, ".conc.tsv"))
})
# Check if input files and output directories exist
io$check_io_exist(FILE)

set.seed(7579)
```

# Given data

```{r}
#| label: 'read-given-tsv-and-save-sumexp'
#| purl: true

# Parse the data in the file into SummarizedExperiment
parse_to_sumexp <- function(msdial_file) {
  # Function to extract sample information from the first 5 lines
  extract_sample_info <- function(msdial_file) {
    # Read sample information lines (first 5 lines)
    sample_lines <- read.delim(
      msdial_file, 
      header = FALSE,
      nrow = 5L,
      na.strings = c("NA", "")
    )
    # Identify the columns of sample information by the second row ("File type" row)
    i_col <- which(! is.na(sample_lines[2, ]))
    header <- sample_lines[, i_col[1L]]
    i_col <- i_col[-1L]    # Exclude the header column
    # Extract sample information  
    sinfo <- t(sample_lines[, i_col, drop = FALSE])
    colnames(sinfo) <- header
    stopifnot(      # Confirm the column names are as expected
      identical(colnames(sinfo)[-5], c("Class", "File type", "Injection order", "Batch ID"))
    )
    colnames(sinfo)[5] <- "given_sample_id"
    # Function to simplify the sample IDs removing repeated characters
    simplify_sample_ids <- function(ids, class) {
      ids <- stringr::str_remove(ids, "^[[:digit:]]+_")    # Remove the numeric prefix
      ids <- dplyr::case_when(
        class == "CalCurve" ~ stringr::str_extract(ids, "Cal_[[:digit:]-]+_"),
        class == "QC"       ~ stringr::str_extract(ids, "QC[[:alnum:]]*_"),
        class == "Blank"    ~ stringr::str_extract(ids, "Blank[[:digit:]]*_"),
        TRUE ~ ids
      ) |> 
        stringr::str_remove("_$")    # Remove trailing underscore
      stopifnot(! any(ids == ""))
      ids
    }
    # Create a tibble with the sample information
    sinfo <- tibble::as_tibble(sinfo) |> 
      dplyr::mutate(
        sample_id = make.names(given_sample_id, unique = TRUE),    # Syntactically valid
        short_sample_id = simplify_sample_ids(given_sample_id, Class)
      ) |>
      dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
        sample_id, 
        Class,
        sample_type = "File type",
        injection_order = "Injection order",
        given_sample_id,
        short_sample_id,
      )
    attr(sinfo, "i_col_in_msdial_file") <- i_col
    return(sinfo)
  }
  sample_info <- extract_sample_info(msdial_file)
  i_col_in_msdial_file <- attr(sample_info, "i_col_in_msdial_file")
  
  # Read lines that contain chemical data
  chem_lines <- readr::read_tsv(
    msdial_file, 
    skip = 4L,
    col_types = readr::cols(.default = readr::col_character()),
    name_repair = "unique_quiet"
  )
  # Extract chemical info columns
  chems <- chem_lines |> 
    dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
      alignment_id = "Alignment ID",
      chem_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      sn_ratio = "S/N average",
      std_type = "Comment",            # Type of standard, e.g. "Quant", "IS", or NA 
    ) |> 
    # `chem_id` = Syntactically valid ID
    dplyr::mutate(chem_id = make.names(chem_name, unique = TRUE), .before = 1L)
  chems <- chems |> 
    dplyr::mutate(dplyr::across(c(mz, rt, sn_ratio), as.numeric))
  
  # Values of the chemicals into a matrix
  raw_df <- chem_lines[, i_col_in_msdial_file, drop = FALSE]
  stopifnot(identical(colnames(raw_df), sample_info$given_sample_id))
  colnames(raw_df) <- sample_info$sample_id     # Update with syntactically valid names
  raw_mat <- lapply(raw_df, as.numeric) |> 
    as.data.frame(row.names = chems$chem_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(raw = raw_mat),
    colData = sample_info,
    rowData = chems,
  )
}
given_data_sumexp <- parse_to_sumexp(FILE$i)
```

The given file `` `r FILE$i` `` contains data of `r nrow(given_data_sumexp)` chemicals and `r ncol(given_data_sumexp)` samples.

## Samples

There are `r dplyr::n_distinct(given_data_sumexp$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
table(given_data_sumexp$Class) |> 
  knitr::kable(col.names = c("Class", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

Some of them are quality control samples.
Aliquots of the samples were measured as technical replicates.
The replicates of the samples were as below.

```{r}
coldata <- SummarizedExperiment::colData(given_data_sumexp)
table(coldata$short_sample_id[coldata$Class == "QC"]) |> 
  knitr::kable(col.names = c("Sample ID", "Number of replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Metabolites

The data contained `r nrow(given_data_sumexp)` chemicals.
There were `r dplyr::n_distinct(rowData(given_data_sumexp)$std_type)` standard types of chemicals.
The number of chemicals for each type is as follows:

```{r}
table(rowData(given_data_sumexp)$std_type) |> 
  knitr::kable(col.names = c("Standard Type", "Number of chemicals")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r}
#| label: 'extract-standard-chemicals'
#| purl: true

# Extract chemicals that are standards
is_std <- ! is.na(rowData(given_data_sumexp)$std_type)
standard_chems <- given_data_sumexp[is_std, ]
```

## Distribution of the values

### Overall samples and chemicals

#### Linear scale

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 8 
#| out-width: '60%'
p <- ggplot(given_data_sumexp) +
  geom_histogram(aes(x = raw), bins = 50) +
  labs(title = "Distribution of the values", x = "Values")
print(p)
```

#### Logarithmic scale

```{r}
#| label: 'distribution-values-overall-log'
#| fig-width: 8
#| out-width: '60%'
p + scale_x_log10()
```

### By sample class and overall chemicals

```{r}
#| label: 'distribution-values-by-class'

ggplot(given_data_sumexp) +
  geom_histogram(aes(x = raw), bins = 30) +
  labs(title = "Distribution of the values by sample class", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ Class, scales = "free_y", ncol = 4)
```

### By chemical and overall samples

Randomly selected 8 standard chemicals are shown in the following figure.

```{r}
#| purl: true
#' Randomly select `n` rows from `sumexp` with optional `subset`
#'
#' @param sumexp a SummarizedExperiment object
#' @param n the number of rows to select
#' @param subset an optional subset expression to apply to the `rowData` of `sumexp`
#' @param in_previous_order if `TRUE`, the selected rows are sorted in the previous order
#'
#' @return a SummarizedExperiment object with `n` rows randomly selected from `sumexp`
#' @export
select_rows_randomly <- function(sumexp, n, subset, in_previous_order = FALSE) {
  ii <- c(1:nrow(sumexp))
  if (!missing(subset)) {      # If given, apply it to the `rowData` of `sumexp`
    ii <- ii[eval(substitute(subset), as.data.frame(rowData(sumexp)))]
  }
  ii <- sample(ii, n)
  if (in_previous_order) ii <- sort(ii)
  sumexp[ii, ]
}
```

```{r}
#| label: 'distribution-values-by-chemical'
select_rows_randomly(standard_chems, 8, std_type != "IS") |> 
  ggplot() +
  geom_histogram(aes(x = raw, fill = Class), bins = 30) +
  labs(title = "Distribution of the values by chemical", x = "Values") +
  scale_x_log10() +
  facet_wrap(~ chem_name, scales = "free_x", ncol = 4)
```

----------------------------

# Normalization using internal standards

## Internal standard (IS) chemicals

```{r}
#| purl: true

# Extract the internal standard chemicals
internal_std_se <- standard_chems |> 
  dplyr::filter(std_type == "IS")
# Sort by average retention time
rt <- rowData(internal_std_se)$rt
internal_std_se <- internal_std_se[order(rt), ]
```

There are `r nrow(internal_std_se)` internal standard (IS) chemicals in the data.
Those chemicals are `r knitr::combine_words(rowData(internal_std_se)$chem_name, before = "\x60")`, in the order of average retention time.

### Distribution of the values

Randomly selected 8 internal standard (IS) chemicals are shown in the following figure.

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 8

# Keep the order by m/z
istd_to_show_se <- select_rows_randomly(internal_std_se, 8, in_previous_order = TRUE)
istd_to_show_se |> 
  # Below invokes `tidySummarizedExperiment:::mutate.SummarizedExperiment`
  dplyr::mutate(chem_name = forcats::fct(chem_name)) |>  # Keep the order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard chemicals", 
    y = "Peak area",
  ) +
  theme(legend.position = "none") +   # No legend for fill
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Rotate x-axis labels
  scale_y_log10() +
  facet_wrap(~ chem_name, scales = "fixed", ncol = 4)
```

### Outliers

Outliers were identified based on the values of internal standard chemicals.
Those samples with values greater than 3 standard deviations from the mean after log-transformation ($log(1 + x)$) were considered as outliers.

```{r}
#| purl: true

# Identify outliers
identify_outliers <- function(x) {
  m <- mean(x)
  sd <- sd(x)
  return((x > m + 3 * sd) | (x < m - 3 * sd))
}
# Number of outliers per sample
n_outlier <- assays(internal_std_se)$raw |> 
  log1p() |>              # Log-transform the data
  apply(1, identify_outliers) |>
  rowSums()         # Number of outliers per sample (transposed by `apply` above)
# Exclude the outliers
stopifnot(identical(names(n_outlier), colnames(internal_std_se)))
internal_std_se <- internal_std_se[, n_outlier == 0, drop = FALSE]
stopifnot(identical(names(n_outlier), colnames(standard_chems)))
standard_chems$is_outlier <- n_outlier > 0
no_outlier_se <- standard_chems[, n_outlier == 0, drop = FALSE]
```

There are `r sum(n_outlier > 0)` outlying samples in the data.
The data of those outliers were excluded in the following analysis. 

```{r}
#| label: 'outliers-table'
#| results: 'markup'
print(n_outlier[n_outlier > 0])
```

After exclusion of those outliers, the distribution of the randomly selected 8 internal standard (IS) chemicals are shown below.

```{r}
#| label: 'distribution-values-internal-standards-no-outliers'
#| fig-width: 13
#| fig-height: 8

internal_std_se[rownames(istd_to_show_se), ] |>
  dplyr::mutate(chem_name = forcats::fct(chem_name)) |>  # Order by m/z 
  ggplot() +
  geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
  labs(
    title = "Distribution of the values of internal standard chemicals (no outliers)", 
    y = "Values"
  ) +
  theme(legend.position = "none") +   # No legend for fill
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +    # Rotate x-axis labels
  scale_y_log10() +
  facet_wrap(~ chem_name, scales = "fixed", ncol = 4)
```

### raw RSD%

The relative standard deviation (RSD%) of the raw values of the internal standard chemicals are calculated.

#### By quality control groups

```{r}
#| purl: true
#' Calculate RSD%
#'
#' @param x A numeric vector
#' @return A numeric value of RSD%
#' @export
rsd_perc <- function(x) {
  100 * sd(x) / mean(x)
}
# Extract QC samples into list
extract_qc_samples_to_list <- function(se) {
  se <- se[, se$Class == "QC"]
  # Column-wise split
  lapply(setNames(nm = unique(se$short_sample_id)), \(ea) se[, se$short_sample_id == ea])
}
```

```{r}
#| label: 'rsd-internal-standards-by-qc-groups'

# Divide the internal standard chemicals by quality control groups
# Calculate RSD% per internal standard chemical and per quality control groups
extract_qc_samples_to_list(internal_std_se) |> 
  sapply(\(se) {
    apply(assay(se, "raw"), 1, \(.x) round(rsd_perc(.x), 2))
  }) |>
  knitr::kable(caption = "RSD% of internal standard chemicals by quality control groups") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### By sample classes

```{r}
sapply(setNames(nm = unique(internal_std_se$Class)), function(cls) {
  se1 <- internal_std_se[, internal_std_se$Class == cls]
  apply(assay(se1, "raw"), 1, \(.x) round(rsd_perc(.x), 2))
}) |> 
  knitr::kable(caption = "RSD% of internal standard chemicals by sample classes") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Normalization

The data will be normalized using the internal standard chemicals.

```{r}
#| purl: true

# Find the range of RT values
overall_rt_range <- range(rowData(no_outlier_se)$rt)
rt <- rowData(internal_std_se)$rt
```

The range of RT values of overall chemicals is `r overall_rt_range[1]` to `r overall_rt_range[2]`,
while the range of internal standard chemicals is `r range(rt)[1]` to `r range(rt)[2]`.

### Closest internal standard chemicals

The closest internal standard chemicals to each chemical will be identified based on the retention time.
The measured peak area of each chemical will be divided by the peak area of the closest internal standard chemical.

```{r}
#| label: 'normalize-rt-closest-istd'
#| purl: true

# Get the values of the closest internal standard chemicals
get_raw_of_closest_istd <- function(rt_x, internal_std_se) {
  # Find the closest internal standard chemical
  closest_istd <- which.min(abs(rowData(internal_std_se)$rt - rt_x))
  return(assays(internal_std_se)$raw[closest_istd, ])
}
assays(no_outlier_se)$closest_istd_raw <- rowData(no_outlier_se)$rt |> 
  setNames(rownames(no_outlier_se)) |>          # To keep the names
  sapply(get_raw_of_closest_istd, internal_std_se) |> 
  t()     # Rows are chemicals

# Normalize the data using closest internal standard chemicals
assays(no_outlier_se)$closest_norm <- 
  assays(no_outlier_se)$raw / assays(no_outlier_se)$closest_istd_raw
```

#### Evaluation of normalization

```{r}
#| purl: true

# Extract quantitative standards in QC samples
extract_quant_qc <- function(sumexp) {
  se_lst <- extract_qc_samples_to_list(sumexp)
  lapply(se_lst, \(se) {
    is_qstd <- rowData(se)$std_type == "Quant"
    se[is_qstd, ]
  })
}
# Calculate RSD of the quantification standard samples
calc_rsd_qstd <- function(qc_se, assay_ids) {
  lapply(qc_se, \(qc1) {
    sapply(setNames(nm = assay_ids), function(nm) {
      apply(assays(qc1)[[nm]], 1, rsd_perc)
    }) |>    # Rows are chemicals, columns are assay IDs
      tibble::as_tibble(rownames = "chem_id")
  }) |> 
    # `QC` has the name of the QC samples
    dplyr::bind_rows(.id = "QC")
}
# Plot the RSD% of various data sets of chemicals
ggplot_rsdp_metab <- function(rsd_df, assay_ids) {
  nms <- names(assay_ids)
  ids <- unname(assay_ids)    # Duplicated by `dplyr::all_of`
  rsd_df <- rsd_df |> 
    dplyr::mutate(
      # NA was generated when all values were zero
      dplyr::across(dplyr::all_of(ids), \(.x) ifelse(is.na(.x), 0, .x)),
    ) |> 
    tidyr::pivot_longer(dplyr::all_of(ids), names_to = "Data", values_to = "RSD%") |> 
    dplyr::mutate(Data = factor(Data, levels = ids, labels = nms))
  ggplot(rsd_df) +
    aes(x = Data, y = `RSD%`) +
    geom_line(aes(group = chem_id), alpha = 0.5)
}
```
```{r}
#| label: 'evaluation-normalization-closest-istd'
#| out-width: '80%'

# Quantification standard samples
qc_se <- extract_quant_qc(no_outlier_se)

assay_ids <- setNames(c("raw", "closest_norm"), nm = c("Raw", "Closest IS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > closest_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of chemicals with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'evaluation-normalization-closest-istd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```


### LOESS fit along RT

```{r}
#| label: 'normalize-rt'
#| purl: true

excluded_classes_for_normalization <- c("Blank", "ProcBlank", "CalCurve")

loess_fit <- local({
  # Log-transform the data
  istd_log <- log(assays(internal_std_se)$raw)
  rt <- rowData(internal_std_se)$rt
  
  # Normalize the data using the internal standards
  # Mean of each internal standard chemical for overall measurement samples
  # , excluding the calibration curve and blank samples
  to_excl <- internal_std_se$Class %in% excluded_classes_for_normalization
  # The data of the samples with measurements
  mat <- istd_log[, !to_excl, drop = FALSE]
  stopifnot(all(is.finite(mat)))
  istd_m <- rowMeans(mat)
  # Subtract the mean from the data
  fr_m_log <- istd_log - istd_m
  
  # Expand to smallest RT 
  istd_smallest_rt <- fr_m_log[which.min(rt), ]
  # Expand to largest RT 
  istd_largest_rt <- fr_m_log[which.max(rt), ]
  # Add the expanded range
  fr_m_log <- rbind(
    istd_smallest_rt, 
    istd_largest_rt,
    fr_m_log
  )
  
  # LOESS fit of the internal standard chemicals along RT
  loess_fit <- purrr::map(1:ncol(fr_m_log), function(ii) {
    # Fit a loess curve
    y <- fr_m_log[, ii]
    y[is.finite(y) == FALSE] <- NA
    stopifnot(identical(
      names(y), c("istd_smallest_rt", "istd_largest_rt", rownames(internal_std_se))
    ))
    suppressWarnings(loess(y ~ c(overall_rt_range, rt), span = 1))
  })
  names(loess_fit) <- colnames(fr_m_log)
  loess_fit
})

# Normalize the data by LOESS fit along RT
rt <- rowData(no_outlier_se)$rt
raw <- assays(no_outlier_se)$raw
assays(no_outlier_se)$loess_norm <- sapply(
  colnames(no_outlier_se), function(sample_id) {
    norm_factor <- predict(loess_fit[[sample_id]], newdata = rt)
    exp(log(raw[, sample_id]) - norm_factor)
  }
)

# Extract quantification standard data
quant_sumexp <- no_outlier_se |> 
  dplyr::filter(std_type == "Quant")
```

For each chemical, the average of all samples excluding 
`r knitr::combine_words(excluded_classes_for_normalization, before = "\x60")`
samples was calculated after log-transformation.
The deviation of each sample from the average was calculated as normalization factor.
The data of the sample are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) chemicals along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all chemicals. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all chemicals.
The normalized data are calculated by subtracting the normalization factor from the log-transformed data.

Randomly selected 3 samples are shown below.

```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 12

to_excl <- no_outlier_se$Class %in% excluded_classes_for_normalization
# Plot the normalized and original data for randomly selected 3 samples
i_sel <- sample(c(1:ncol(no_outlier_se))[!to_excl], 3)

# To include LOESS smooth line
x <- seq(overall_rt_range[1], overall_rt_range[2], length.out = 100)
loess_curve_df <- lapply(setNames(i_sel, colnames(no_outlier_se)[i_sel]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(rt = x) |>
  tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")
loess_curve_df$sample_name <-
  no_outlier_se$short_sample_id[match(loess_curve_df$sample_id, no_outlier_se$sample_id)]

# Log-transform the data
raw_log <- log(assays(no_outlier_se)$raw)
# Overall mean of every chemical
mat <- raw_log[, !to_excl, drop = FALSE]
mat[is.finite(mat) == FALSE] <- NA    # Visualization purpose only
mean_log <- rowMeans(mat, na.rm = TRUE)
# Deviation from the mean
selected_se <- no_outlier_se[, i_sel]
assays(selected_se) <- assays(selected_se) |> 
  lapply(\(x) sweep(log(x), 1, mean_log, FUN = "-"))

df_to_show <- tibble::as_tibble(selected_se) |>
  dplyr::rename(Raw = raw, Normalized = loess_norm, sample_name = short_sample_id) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value")

# Plot the normalized and original data
ggplot(df_to_show, aes(x = rt, y = Value)) +
  geom_point(aes(color = Data, shape = std_type)) +
  geom_line(data = loess_curve_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values) - log(avg)"
  ) +
  scale_shape_manual(values = c("IS" = 15, "Quant" = 21)) +
  facet_wrap(~ sample_name, scales = "free_y", ncol = 3)
```

The line in the graph represents the LOESS fit.
The points at the bottom boundary are the **0**s, which were not normalized by this process.
Those zeros were kept as zeros in the normalized data.

#### Evaluation of normalization

The normalized data are evaluated by the changes of relative standard deviation (RSD) of the quantification standard samples.

```{r}
#| label: 'rsd-eval'
#| out-width: '80%'

# Quantification standard samples
qc_se <- extract_quant_qc(quant_sumexp)
# Calculate RSD of the quantification standard samples
assay_ids <- setNames(c("raw", "loess_norm"), nm = c("Raw", "LOESS Normalized"))
rsd_qstd <- calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > loess_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of chemicals with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'rsd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

### Comparison of the normalization methods

The RSD% of the quantification standard samples is calculated for the raw data and the normalized data.

```{r}
#| label: 'norm-method-comparison'
#| out-width: '80%'

assay_ids <- setNames(c("raw", "closest_norm", "loess_norm"), nm = c("Raw", "Closest", "LOESS"))
df_to_show <- extract_quant_qc(quant_sumexp) |> 
  calc_rsd_qstd(assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(loess_norm <= closest_norm, "↓ LOESS", "↓ Closest"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓ LOESS", "↓ Closest", "All 0s"))
  ) 
ggplot_rsdp_metab(df_to_show, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of chemicals with lower RSD after each normalization is counted below.

```{r}
#| label: 'rsd-comparison-count'
#| results: 'markup'
# Count the number of chemicals with lower RSD
df_to_show |> 
  janitor::tabyl(QC, sign)
```

The number of compounds above or below 20 RSD% is counted below.

```{r}
#| label: 'rsd-comparison-count-20'
df_to_show |> 
  dplyr::summarize(
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x > 20, na.rm = TRUE), 
      .names = "{.col}_gt_20"
    ),
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x <= 20, na.rm = TRUE), 
      .names = "{.col}_eq_lt_20"
    ),
    .by = QC
  ) |>  
  knitr::kable(
    caption = "Number of chemicals with RSD% above or below 20", 
    col.names = c("QC", rep(names(assay_ids), 2))
  ) |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "RSD > 20%" = 3, "RSD ≤ 20%" = 3)) 
```


The RSD% values are presented in the table below.

```{r}
#| label: 'rsd-comparison-table'

df_to_show |> 
  dplyr::left_join(as.data.frame(rowData(quant_sumexp)), by = "chem_id") |>   # Add chemical names
  dplyr::select(QC, chem_name, dplyr::all_of(unname(assay_ids))) |> 
  dplyr::rename(all_of(assay_ids), `Metabolite` = chem_name) |> 
  # Caption at the top
  knitr::kable(caption = "RSD% of the quantification standard samples", digits = 2) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |>
  kableExtra::scroll_box(height = "400px")
```

# Calibration

## Calibration curve samples

The samples for the calibration curve are identified by the sample class, `CalCurve`.
The concentration is extracted from the sample ID, which is in the format `Cal_[[:digit:]-]+_`, and converted to a numeric value. 
The samples with a concentration of `0` are separated as the zero samples.

```{r}
#| label: 'calcurve-samples'
#| purl: true

# Calibration curve samples only
calcurve_se <- quant_sumexp[, quant_sumexp$Class == "CalCurve"]
stopifnot("Calibration curve samples are required." = nrow(calcurve_se) > 0)
# Function to fetch concentration values from the given IDs
fetch_concentration <- function(sid) {
  sid |> 
    stringr::str_remove("Cal_") |> 
    stringr::str_replace("-", ".") |>     # Replace "-" with "."
    as.numeric()
}
calcurve_se$conc <- fetch_concentration(calcurve_se$short_sample_id)
# Calibration zero samples
cal_0_se <- calcurve_se[, calcurve_se$conc == 0] 
stopifnot("Multiple `Cal_0` samples are required." = nrow(cal_0_se) > 1)
# Remove the 0 concentration samples
calcurve_se <- calcurve_se[, calcurve_se$conc != 0]
```

### Sample IDs

```{r}
#| results: 'markup'
# Present the sample IDs for the calibration curve
SummarizedExperiment::colData(calcurve_se)[, c("given_sample_id", "conc")] |> 
  tibble::as_tibble() |> 
  dplyr::arrange(conc) |> 
  knitr::kable(
    col.names = c("Sample ID", "Concentration (ng/ml)"),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
calcurve_se$conc |> 
  table() |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Concentration ranges

```{r}
#| purl: true
classes_not_for_measurement <- c("CalCurve", "QC")
```

Quite a few concentrations of calibration curve samples of a compound may be too far from the range of the measured samples. 
Hence the concentrations were excluded for which the peak areas after normalization were more than 10 times higher than the maximum peak area of the samples except `r knitr::combine_words(classes_not_for_measurement, before = "\x60")` classes.

```{r}
#| label: 'calcurve-upper-limit-concentration-range'
#| purl: true

# Find maximum peak area of the samples of the classes for measurement
rowData(calcurve_se)$max_loess_norm <- local({
  stopifnot(identical(rownames(quant_sumexp), rownames(calcurve_se)))
  sumexp <- quant_sumexp[, ! quant_sumexp$Class %in% classes_not_for_measurement]
  apply(assays(sumexp)$loess_norm, 1, max, na.rm = TRUE)
})
# Function to replace .x with NA if it is more than 10 times higher than .y
replace_if_10x_higher <- function(.x, .y) {
  ifelse(.x > 10 * .y, NA_real_, .x)
}
# Exclude the concentrations that are too far from the measured samples
assays(calcurve_se)$loess_norm <- assays(calcurve_se)$loess_norm |> 
  replace_if_10x_higher(rowData(calcurve_se)$max_loess_norm)
rowData(calcurve_se)$max_conc <- assays(calcurve_se)$loess_norm |> 
  apply(1, \(.x) {
    if (all(is.na(.x))) return(NA_real_)
    max(calcurve_se$conc[!is.na(.x)])
  })
```

When the obtained peak area of low concentration of the calibration samples is too low, the concentrations are excluded.
The value is considered too low when the average of the values of one concentration is less than 10 times of the standard deviation from the mean of zero concentration samples. 

```{r}
#| label: 'calcurve-lower-limit-concentration-range'
#| purl: true

# Function: find the mean + `times` of standard deviation 
get_mean_plus_sd <- function(mat, times = 10) {
  m <- apply(mat, 1, mean, na.rm = TRUE)
  s <- apply(mat, 1, sd, na.rm = TRUE)
  m + times * s
}
# Function to replace .x with NA if it is less than .y
replace_if_lower <- function(.x, .y) {
  ifelse(.x < .y, NA_real_, .x)
}

# Find the mean and standard deviation of the zero concentration samples
thres <- get_mean_plus_sd(assays(cal_0_se)$loess_norm)
# Exclude the concentrations that are too low
assays(calcurve_se)$loess_norm <- replace_if_lower(assays(calcurve_se)$loess_norm, thres)
rowData(calcurve_se)$min_conc <- assays(calcurve_se)$loess_norm |>
  apply(1, \(.x) {
    if (all(is.na(.x))) return(NA_real_)
    min(calcurve_se$conc[!is.na(.x)])
  })
```

Remaining concentration range of each compound is presented below.

```{r}
#| label: 'calcurve-conc-range-table'

rowData(calcurve_se) |> 
  tibble::as_tibble() |> 
  dplyr::select(chem_name, min_conc, max_conc) |>
  dplyr::mutate(
    min_conc = dplyr::case_when(
      is.na(min_conc) ~ "",
      TRUE ~ scales::number(min_conc)
    ),
    max_conc = dplyr::case_when(
      is.na(max_conc) ~ "All 0s, no valid",
      TRUE ~ scales::number(max_conc)
    )
  ) |>
  knitr::kable(
    caption = "Concentration range of the calibration curve samples",
    col.names = c("Metabolite",
                  "Minimum (ng/ml)",
                  "Maximum concentration (ng/ml)"), 
    align = c("l", "r", "r"),
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "400px")
```


## Calibration curve per chemical

```{r}
#| label: 'fit_layer'
#| purl: true

set.seed(7579)
# Randomly select 9 calibration curve samples
i_samples_in_calcurv <- local({
  ii <- c(1:nrow(calcurve_se))
  ii[rowData(calcurve_se)$std_type != "IS"] |> 
    sample(9)
})
calcurve_to_show_df0 <- tibble::as_tibble(calcurve_se[i_samples_in_calcurv, ])
calcurve_to_show_df <- calcurve_to_show_df0 |> 
  tidyr::drop_na(loess_norm)
```

### Before normalization

```{r}
#| label: 'calcurve-per-chemical-linear-before-norm'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df0, aes(x = conc, y = raw)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Raw)"
  ) +  
  facet_wrap(~ chem_name, scales = "free_y", ncol = 3)
```

### After normalization

The peak areas were normalized by the internal standards applying the loess normalization.

#### Linear scale

```{r}
#| label: 'calcurve-per-chemical-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = loess_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calcurve-per-chemical-log'
#| fig-width: 12
#| fig-height: 11
ggplot(calcurve_to_show_df, aes(x = conc, y = loess_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

### After normalization by closest IS in RT

```{r}
#| label: 'calcurve-per-chemical-linear-closest'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
ggplot(calcurve_to_show_df, aes(x = conc, y = closest_norm)) +
  geom_point() +
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

## Calibration of the chemicals

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.


```{r}
#| label: 'calibration-data'
#| purl: true

quadratic_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc + I(conc^2), weights = weights)
  beta <- coef(lmfit)
  a <- beta[["I(conc^2)"]]
  b <- beta[["conc"]]
  cc <- beta[["(Intercept)"]]
  out <- function(x) {
    det <- b ^ 2 - 4 * a * (cc - x)
    conc <- suppressWarnings((-b + sqrt(det)) / (2 * a))      # det < 0
    conc[conc < 0 | conc < 0] <- 0
    conc
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}
linear_calcurve_model <- function(conc, value, weights) {
  lmfit <- lm(value ~ conc, weights = weights)
  beta <- coef(lmfit)
  b1 <- beta[["conc"]]
  b0 <- beta[["(Intercept)"]]
  out <- function(x) {
    (x - b0) / b1
  }
  attr(out, "R2") <- summary(lmfit)$r.squared
  out
}

build_calcurve_model <- function(conc, value) {
  if (all(is.na(value) | value == 0)) {
    return(list(
      "best_model" = function(x) x,
      "R2s" = NA,
      "best_model_name" = NA_character_
    ))
  }
  # Weight alternatives
  weights_alt <- rlang::list2(
    "1" = rep(1, length(conc)),
    "1/x" = 1 / conc,
    "1/x2" = 1 / (conc ^ 2),
  )
  # Linear and quadratic models
  lmodels <- lapply(weights_alt, \(w) linear_calcurve_model(conc, value, weights = w))
  names(lmodels) <- paste("linear", names(lmodels), sep = "-")
  qmodels <- lapply(weights_alt, \(w) quadratic_calcurve_model(conc, value, weights = w))
  names(qmodels) <- paste("quadratic", names(qmodels), sep = "-")
  models <- c(lmodels, qmodels)
  R2s <- sapply(models, \(f) attr(f, "R2"))
  list(
    "best_model" = models[[which.max(R2s)]],        # Best model by R2
    "R2s" = R2s,
    "best_model_name" = names(models)[which.max(R2s)]
  )
}

# Fit the calibration curve
conc <- calcurve_se$conc
rt_norm_mat <- assays(calcurve_se)$loess_norm
calcurve_models <- lapply(setNames(nm = rownames(calcurve_se)), function(ii) {
  build_calcurve_model(conc, rt_norm_mat[ii, ])
})
R2s <- lapply(calcurve_models, \(m) m$R2s)
R2s <- R2s[!is.na(R2s)] |> 
  dplyr::bind_rows(.id = "chem_id")

# Compute the concentration of the samples
classes_to_incl <- unique(quant_sumexp$Class) |> 
  setdiff(classes_not_for_measurement)      # Exclude the control samples
measurements_se <- quant_sumexp[, quant_sumexp$Class %in% classes_to_incl]
assays(measurements_se)$conc <- rowData(measurements_se)$chem_id |> 
  sapply(function(chem_id) {
    v <- assays(measurements_se)$loess_norm[chem_id, ]
    calcurve_models[[chem_id]]$best_model(v)
  }) |>
    t()          # Metabolites to rows
```

The distribution of $R^2$ values by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'

R2s |> 
  tidyr::pivot_longer(cols = -chem_id, names_to = "Model", values_to = "R2") |>
  ggplot() +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(
    x = "Model",
    y = expression(R^2)
  )
```

The selected best models were as below.

```{r}
#| label: 'calibration-best-models'
#| results: 'markup'
table(sapply(calcurve_models, \(m) m$best_model_name))
```

### All measurement samples

The concentration of the sample classes, `r knitr::combine_words(classes_to_incl)`, are computed using the calibration curve.
Any negative concentration is set to zero.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
measure_tbl <- tibble::as_tibble(measurements_se[i_samples_in_calcurv, ])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
# Retrieve the coefficients of the linear model for the calibration curve
chem_ids <- unique(measure_tbl$chem_id)
# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  # geom_line(data = calcurve_smooth, color = "cadetblue", alpha = 0.5) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

```{r}
#| purl: true

classes_excl_qc_blank <- setdiff(unique(measurements_se$Class), c("QC", "Blank"))
```


### `r knitr::combine_words(classes_excl_qc_blank, before = "\x60")` samples

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11
to_incl <- measurements_se$Class %in% classes_excl_qc_blank
measure_tbl <- tibble::as_tibble(measurements_se[i_samples_in_calcurv, to_incl])

# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
# Plot the calibration curve with the samples
ggplot(measure_tbl, aes(x = conc, y = loess_norm)) +
  geom_point(data = calcurve_to_show_df) +
  # geom_line(data = calcurve_smooth, color = "cadetblue", alpha = 0.5) +
  geom_point(aes(color = Class)) +
  geom_rug(color = rgb(0.5, 0, 0, alpha = 0.2)) +   # Density of points
  labs(
    x = "Concentration (ng/ml)",
    y = "Peak area (Normalized)"
  ) +  
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
```

# Concentration

The concentration of the samples has been computed using the calibration curve in the [section above](#calibration).
The distribution of the concentration of the samples is shown below.

```{r}
#| label: 'concentration-distribution-by-sample'
#| fig-width: 12
#| fig-height: 10
ggplot(measurements_se) +
  geom_boxplot(aes(x = given_sample_id, y = conc, fill = Class)) +
  labs(
    title = "Concentration distribution by class", 
    x = "Sample ID",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10()
```

```{r}
#| label: 'concentration-distribution-by-chemical'
#| fig-width: 12
#| fig-height: 20
ggplot(measurements_se) +
  geom_boxplot(aes(x = chem_name, y = conc)) +
  labs(
    title = "Concentration distribution by chemical", 
    x = "Metabolite",
    y = "Concentration (ng/ml)",
  ) +
  # Turn the x-axis title to vertical
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_log10() + 
  facet_grid(Class~.)
```

```{r}
#| label: 'write-concentration-table'
# Write the concentration table
write_concentration_tsv <- function(sumexp, file) {
  # Prepare the chemical table
  chem_columns <- SummarizedExperiment::rowData(sumexp) |> 
    tibble::as_tibble() |> 
    dplyr::select(
      `Alignment ID` = alignment_id,
      `Metabolite name` = chem_name,
      `Average Mz` = mz,
      `Average Rt(min)` = rt,
    )
  # Prepare the concentration table
  conc_df <- SummarizedExperiment::assays(sumexp)$conc |> 
    round(3) |>     # Reduce the number of decimals
    tibble::as_tibble()
  # Original sample ID
  colnames(conc_df) <- SummarizedExperiment::colData(sumexp)$given_sample_id
  readr::write_tsv(dplyr::bind_cols(chem_columns, conc_df), file)
}
write_concentration_tsv(measurements_se, FILE$o)
```


```{r}
#| echo: false
#| results: 'asis'
# Avoid duplication of the bibliography when this is called as a child.
if(! knitr::opts_knit$get("child")) cat('\n\n# References\n\n')
```

