---
params:
  input_file: "data/raw_internal/MMLC_ESIpos_for_Mun-Gwan.txt"
title: "Read MS-DIAL output file"
---

```{r}
#| label: 'setup-read-msdial'
#| purl: true

# Load packages and project local libraries
box::use(
  ggplot2[...],        # Too long to write and usually specific enough
  # Some methods were defined using S4 system, e.g. dim, `[`
  SummarizedExperiment[rowData, `rowData<-`, colData, `colData<-`, assays, `assays<-`],
  tidySummarizedExperiment,      # Defined using S3 but not exported, e.g. ggplot, as_tibble
)
# Input file
FILE <- list(i = params$input_file)
set.seed(7579)
```

```{r}
#| label: 'read-given-tsv-and-save-sumexp'
#| purl: true

#' Identify three column sections in the MS-DIAL output file
#' 
#' @param msdial_file Path to the MS-DIAL output file
#' @return A list of three column indices
get_three_section_indices <- function(msdial_file) {
  # Read the second (`File type`) line to identify three sections
  file_type_line <- read.delim(
    msdial_file, header = FALSE, nrow = 5L, na.strings = c("NA", "")
  )[2, ]
  i_2nd <- which(! is.na(file_type_line))    # Not NA is the 2nd column sections
  stopifnot(file_type_line[i_2nd[1L]] == "File type")    # As expected
  list(
    "1st" = 1L:i_2nd[1L],
    "2nd" = (i_2nd[1L] + 1L):max(i_2nd),         # "+ 1L" for the title column of the rows
    "3rd" = (max(i_2nd) + 1L):length(file_type_line)
  )
}

#' Fetch sample information from the first 5 lines
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the columns that contain sample information
#' @return A tibble with sample information
fetch_sample_info <- function(msdial_file, indices) {
  if (missing(indices)) indices <- get_three_section_indices(msdial_file)[[2]]
  # Read sample information lines (first 5 lines)
  sample_lines <- read.delim(
    msdial_file, 
    header = FALSE,
    nrow = 5L,
    na.strings = c("NA", "")
  )
  header <- sample_lines[, indices[1L] - 1L]
  
  # Extract sample information  
  sinfo <- t(sample_lines[, indices, drop = FALSE])
  colnames(sinfo) <- header          # There was no header in `sinfo` before
  stopifnot("Deviation from the expected names for samples" = identical(
    colnames(sinfo)[-5], c("Class", "File type", "Injection order", "Batch ID")
  ))
  colnames(sinfo)[5] <- "given_sample_id"
  # Create a tibble with the sample information
  sinfo <- tibble::as_tibble(sinfo) |> 
    dplyr::select(
      Class,
      sample_type = "File type",
      injection_order = "Injection order",
      given_sample_id,
    ) |> 
    # Syntactically valid for variable name in R
    dplyr::mutate(sample_id = make.names(given_sample_id, unique = TRUE), .before = 1L) |>
    # Control sample groups
    dplyr::mutate(
      proc_cat = dplyr::case_when(
        sample_type == "Standard" & Class == "CalCurve" ~ "CalCurve",
        sample_type == "QC"    ~ "QC",
        sample_type == "Blank" ~ "Blank",
        TRUE ~ ""         # # NA does not behave predictably with `==`
      )
    )
  stopifnot(
    "Calibration curve samples are required." = any(sinfo$proc_cat == "CalCurve"),
    "QC samples are required." = any(sinfo$proc_cat == "QC"),
    "Blank samples are required." = any(sinfo$proc_cat == "Blank")
  )
  
  # Function to find concentration values from the given IDs
  find_concentration <- function(sid) {
    sid |> 
      stringr::str_extract("Cal_([[:digit:]-]+)", group = 1) |> 
      stringr::str_replace("-", ".") |>     # Replace "-" with "."
      as.numeric()
  }
  sinfo <- sinfo |> 
    dplyr::mutate(
      c_conc = ifelse(proc_cat == "CalCurve", find_concentration(given_sample_id), NA_real_),
      injection_order = as.integer(injection_order),
    )
  stopifnot(
    "Error in Calibration sample IDs" = all(! is.na(sinfo$c_conc[sinfo$proc_cat == "CalCurve"]))
  )
  return(sinfo)
}

#' Fetch data of the selected columns of the MS-DIAL output file
#'
#' @param msdial_file Path to the MS-DIAL output file
#' @param indices The indices of the selected columns
#' @return A tibble 
fetch_data_of_columns <- function(msdial_file, indices) {
  stopifnot(!missing(indices))
  # Count total number of columns in the file
  l1 <- readr::read_tsv(msdial_file, skip = 4L, n_max = 1L, col_names = FALSE, col_types = "c")
  coltypes <- rep("-", ncol(l1))      # Skip the rest columns
  coltypes[indices] <- "c"
  # Read lines that contain chemical data
  readr::read_tsv(
    msdial_file, 
    skip = 4L,      # Data starts from the 5th line
    col_types = paste0(coltypes, collapse = ""),
    name_repair = "unique_quiet"
  )
}

# Parse the data in the file into SummarizedExperiment
parse_to_sumexp <- function(msdial_file) {
  i_three_sections <- get_three_section_indices(msdial_file)
  sample_info <- fetch_sample_info(msdial_file, i_three_sections[[2]])
  
  # Read chemical data in the first section
  chems <- fetch_data_of_columns(msdial_file, i_three_sections[[1]]) |>
    dplyr::select(        # SummarizedExperiment will `make.names` change anyhow
      alignment_id = "Alignment ID",
      chem_name = "Metabolite name",
      mz = "Average Mz",
      rt = "Average Rt(min)",
      sn_ratio = "S/N average",
      std_type = "Comment",            # Type of standard, e.g. "Quant", "IS", or NA 
    ) |> 
    # `chem_id` = Syntactically valid ID
    dplyr::mutate(chem_id = make.names(chem_name, unique = TRUE), .before = 1L) |> 
    dplyr::mutate(
      dplyr::across(c(mz, rt, sn_ratio), as.numeric),
      std_type = dplyr::case_when(
        std_type == "Quant" ~ "Quant",
        std_type == "IS"    ~ "IS",
        TRUE ~ ""              # Ignore other variants
      )
    )
  
  # Measured values of the chemicals into a matrix
  raw_df <- fetch_data_of_columns(msdial_file, i_three_sections[[2]])
  stopifnot(identical(colnames(raw_df), sample_info$given_sample_id))
  colnames(raw_df) <- sample_info$sample_id     # Update with syntactically valid names
  raw_mat <- lapply(raw_df, as.numeric) |> 
    as.data.frame(row.names = chems$chem_id) |> 
    as.matrix()
  
  # Into SummarizedExperiment
  SummarizedExperiment::SummarizedExperiment(
    assays = list(raw = raw_mat),
    colData = sample_info,
    rowData = chems,
  )
}
overall_sumexp <- parse_to_sumexp(FILE$i)  # To be updated throughout the pre-processing
```

Given file: `` `r FILE$i` `` 

The given file contains data of `r nrow(overall_sumexp)` chemicals and `r ncol(overall_sumexp)` samples.

## Samples

### By class

There are `r dplyr::n_distinct(overall_sumexp$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
table(overall_sumexp$Class) |> 
  knitr::kable(col.names = c("Class", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

### Special categories for preprocessing

There are `r length(table(overall_sumexp$proc_cat, exclude = ""))` special sample categories for in the data.
The data of those are important in the following processing steps.
The number of samples for each category is as follows:

```{r}
table(overall_sumexp$proc_cat, exclude = "") |> 
  knitr::kable(col.names = c("Category", "Number of samples")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

#### Calibration curve samples

The samples for the calibration curve are identified by the sample type `Standard` and class `CalCurve`.
The concentration is extracted from the provided sample ID.
The IDs are in the format `Cal_[[:digit:]-]+` with optional numbers and/or characters before and after of it. 
Concentration is extracted from the numeric part in the format after the replacement of `-` with `.`.

##### Sample IDs

```{r}
# Present the sample IDs for the calibration curve
df_show <- colData(overall_sumexp[, overall_sumexp$proc_cat == "CalCurve"]) |> 
  tibble::as_tibble() |> 
  dplyr::arrange(c_conc)
df_show |> 
  dplyr::select(given_sample_id, c_conc) |>
  knitr::kable(
    col.names = c("Sample ID", "Concentration (ng/ml)"),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

##### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
table(df_show$c_conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Quality control samples

```{r}
#| purl: true
# Extract QC samples into list
extract_qc_samples_to_list <- function(se) {
  se <- se[, se$proc_cat == "QC"]
  qc_id <- se$Class
  # Column-wise split
  lapply(setNames(nm = unique(qc_id)), \(ii) se[, qc_id == ii])
}
```

Aliquots of some samples were measured as technical replicates for quality control.
The number of the replicates were as below.

```{r}
extract_qc_samples_to_list(overall_sumexp) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Sample ID", "Number of replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Chemicals

The data contained `r nrow(overall_sumexp)` chemicals.
Among them, `r sum(rowData(overall_sumexp)$std_type != "")` chemicals were targeted. 
The rest were not targeted.

```{r}
#| purl: true
# Extract standard chemicals
is_targeted <- rowData(overall_sumexp)$std_type != ""
targeted_chems_se <- overall_sumexp[is_targeted, ]
```

For those targeted chemicals, the standard type was specified.
The number of chemicals for each type is as follows:

```{r}
table(rowData(targeted_chems_se)$std_type) |> 
  knitr::kable(col.names = c("Standard Type", "Number of chemicals")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Distribution of the values

* Overall samples and targeted chemicals

As shown below, the values are distributed in a skewed manner.

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 16 
#| fig-height: 6
#| out-width: '100%'

p <- ggplot(overall_sumexp) +
    geom_histogram(aes(x = raw), bins = 50)
gridExtra::grid.arrange(   # Two ggplots next to each other
  p + labs(title = "Linear scale", x = "Values"),
  p + scale_x_log10() + labs(title = "Logarithmic scale", x = "Values"),
  ncol = 2
)
```
