---
params:
  norm_method: "loess_norm"
  log_calibration: false
  interactive_plot: false
  project_title: "Test"        # No need here but needed by `report-external.qmd`
title: "Data quality evaluation"
format:
  html: 
    fig-width : 10
    fig-height: 6
    fig-align: "default"
knitr:
  opts_chunk:
    out.width: "100%"
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
# The code in this chunk is NOT evaluated here when knitting this file directly.
# It is included for DEBUGGING purposes only. 
params <- list(norm_method = "loess_norm", log_calibration = FALSE, project_title = "Test") 
source("code/reports/chapters/_internal-shared_setup_chunk.R")
```

```{r}
#| purl: true
set.seed(7579)
```

# Settings used for processing

```{r}
#| child: code/reports/chapters/process-internal-settings.qmd
```

--------------------------------------------

# Data input

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization by internal standards (IS)

## Internal standards (ISs) 

```{r}
#| child: code/reports/chapters/process-internal-stds.qmd
```

## Normalization by ISs

The dataset is normalized using the provided ISs. The user can opt to choose between 2 normalization methods: 
- "Closest RT" method, whereby the peak area intensity of each feature/chemical is corrected with that of the closest IS in RT. Target classes of compounds can be specified.
- "LOESS" method, whereby the peak area intensity of each feature/chemical is corrected by a LOESS model curve built using all the IS spanning across the entire chromatographic RT.

```{r}
#| purl: true
# Find the range of RT values
all_rts <- util$retention_time(pre_norm_se)
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[util$is_targeted_feature(pre_norm_se)])
rt <- range(all_rts[util$is_internal_std(pre_norm_se)])
```

The retention time (RT) range of the dataset is between `r overall_rt_range[1]` and `r overall_rt_range[2]` min.
The RT range of the targeted chemicals to be quantified is from `r targetted_rt_range[1]` to `r targetted_rt_range[2]` min.
The RT range of the internal standards is from `r range(rt)[1]` to `r range(rt)[2]` min.

```{r}
#| purl: true
norm_se <- to_report[["normalized"]]
```

### Closest RT (IS) normalization

```{r}
#| child: code/reports/chapters/process-norm-closest.qmd
```

### LOESS (IS) normalization

```{r}
#| child: code/reports/chapters/process-norm-loess.qmd
```

### Comparison of normalization methods

{{< include chapters/_process-norm-comparison.qmd >}}


### Blank subtraction

After data normalization, peak areas are blank corrected by subtracting the corresponding feature average peak area detected in the blanks (negative values set to zero).
Blank subtraction is applied to all samples in the dataset except for the calibration standards per batch.

```{r}
#| label: target_mode
if (IS_TARGET_MODE) {
  cat("\n\n# Calibration\n\n")
  cat("\nThe calibration is performed per batch.\n\n")
  qmd <- "process-calibration-llox_from_llox_signal.qmd"
  cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))
  # Stop here if the calibration is not performed
  if (all(sapply(lst_proc, is.null))) {
    cat("\n\n# No calibration\n\n")
    cat("Because it is not possible to perform calibration with the provided data.\n\n")
  } else {
    qmd <- "process-calibration-curves.qmd"
    cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))
    cat("\n\n# Concentration\n\n")
    cat(knitr::knit_child("code/reports/chapters/_process-concentration.qmd", quiet = TRUE))
  }
}
```

# References
