---
params:
  norm_method: loess_norm
  project_title: "Test"        # No need here but needed by `report-external.qmd`
title: "Data quality evaluation"
format:
  html: 
    fig-width : 10
    fig-height: 6
    out-width: "100%"
    fig-align: "default"
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
params <- list(norm_method = "loess_norm", project_title = "Test") 
source("code/reports/chapters/_internal-shared_setup_chunk.R")  # For "Run Cell" functionality
```

```{r}
#| purl: true
set.seed(7579)
```

# Data input

{{< include chapters/_read-msdial.qmd >}}

--------------------------------------------

# Normalization by internal standards (IS)

## Volumetric internal standard (vIS)

```{r}
#| label: 'vIS-variable-setup'
#| purl: true
is_vIS <- util$std_type(raw_se) == "vIS"
# Feature intensity to be used for normalization
mat_id_for_norm <- if (any(is_vIS)) "vol_norm" else "raw"
```

```{r}
if (any(is_vIS)) {
  cat(
    "The data contains volumetric internal standards (vIS).\n",
    "All values of all other features are normalized by the value of the vIS feature",
    "for each sample.\n\n",
    "(individual value of each feature and each sample) / (vIS value of each sample) *", 
    "(average of vIS values of all samples)\n"
  )
} else {
  cat("There is no volumetric internal standard (vIS) in the data.\n")
}
```

## Internal standards (ISs) 

{{< include chapters/_process-internal-stds.qmd >}}

## Normalization by ISs

The dataset is normalized using the provided ISs. The user can opt to choose between 2 normalization methods: 
- "Closest RT" method, whereby the peak area intensity of each feature/chemical is corrected with that of the closest IS in RT. Target classes of compounds can be specified.
- "LOESS" method, whereby the peak area intensity of each feature/chemical is corrected by a LOESS model curve built using all the IS spanning across the entire chromatographic RT.

```{r}
#| purl: true
# Find the range of RT values
wo_vIS <- raw_se[!is_vIS, ]
all_rts <- util$retention_time(wo_vIS)
overall_rt_range <- range(all_rts)
targetted_rt_range <- range(all_rts[util$is_targeted_feature(wo_vIS)])
rt <- range(all_rts[util$is_internal_std(wo_vIS)])
```

The retention time (RT) range of the dataset is between `r overall_rt_range[1]` and `r overall_rt_range[2]` min.
The RT range of the targeted chemicals to be quantified is from `r targetted_rt_range[1]` to `r targetted_rt_range[2]` min.
The RT range of the internal standards is from `r range(rt)[1]` to `r range(rt)[2]` min.

### Closest RT (IS) normalization

```{r}
#| purl: true
norm_se <- qc_steps[["normalized"]]
```

{{< include chapters/_process-norm-closest.qmd >}}

### LOESS (IS) normalization

```{r}
#| child: code/reports/chapters/process-norm-loess.qmd
```

### Comparison of normalization methods

{{< include chapters/_process-norm-comparison.qmd >}}


### Blank subtraction

After data normalization, peak areas are blank corrected by subtracting the corresponding feature average peak area detected in the blanks (negative values set to zero).
Blank subtraction is applied to all samples in the dataset except for the calibration standards.

```{r}
#| label: target_mode
if (!SumExp::metadata(raw_se)$is_non_target_mode) {
  cat("\n\n# Calibration\n\n")
  qmd <- "process-calibration-llox_from_llox_signal.qmd"
  cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))
  # Stop here if the calibration is not performed
  if (is.null(calib_interm_data[["calcurve_se all range"]])) {
    cat("\n\n# No calibration\n\n")
    cat("Because it is not possible to perform calibration with the provided data.\n\n")
  } else {
    qmd <- "process-calibration-curves.qmd"
    cat(knitr::knit_child(file.path("code/reports/chapters", qmd), quiet = TRUE))
    cat("\n\n# Concentration\n\n")
    cat(knitr::knit_child("code/reports/chapters/_process-concentration.qmd", quiet = TRUE))
  }
}
```

# References
