
Given file: `` `r SumExp::metadata(raw_se)$file_name` `` 

The given file contains data of `r nrow(raw_se)` features and `r ncol(raw_se)` samples.

## Samples

### By category/class

```{r}
#| purl: true
# Category and class of samples
catg <- util$ctrl_smpl_cat(raw_se)
catg[catg == ""] <- "Sample"      # Samples were not assigned to any special categories
```

There are `r length(table(catg))` categories and 
`r dplyr::n_distinct(SumExp::col_df(raw_se)$Class)` classes of samples in the data.
The number of samples for each class except `CalCurve` and `QC` is as follows:

```{r}
catg_cl <- tibble::tibble(
  Category = catg,
  Class = SumExp::col_df(raw_se)$Class,
) |>
  dplyr::mutate(Class = ifelse(Category %in% c("CalCurve", "QC"), "", Class)) |>
  tidyr::unite("catg_cl", c(Category, Class), sep = ":", remove = FALSE)
tb <- janitor::tabyl(catg_cl, catg_cl) |>
  janitor::adorn_pct_formatting() |>
  dplyr::right_join(dplyr::distinct(catg_cl), y = _, by = "catg_cl") |>
  dplyr::arrange(Category, Class)
stopifnot(anyDuplicated(tb$catg_cl) == 0)   # Ensure no duplicated category-class combinations

tb |>
  dplyr::select(-catg_cl) |>
  knitr::kable(
    row.names = FALSE,          # To avoid excluded row numbers
    col.names = c("Category", "Class", "Number of Samples", "Percent"),
    align = "llrr"
  ) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

### Calibration curve samples

The samples for the calibration curve are identified by the sample type `Standard`.
The concentration is extracted from the provided sample ID.
The IDs are in the format `Cal[[:digit:]-]+` with optional numbers and/or characters before and after of it. 
Concentration is extracted from the numeric part in the format after the replacement of `-` with `.`.

```{r}
#| purl: true
df_show <- if (IS_TARGET_MODE) {
  # Present the sample IDs for the calibration curve
  SumExp::col_df(raw_se) |> 
    dplyr::filter(util$ctrl_smpl_cat(raw_se) == "CalCurve") |> 
    dplyr::arrange(c_conc) |> 
    dplyr::select(sample_name, c_conc)
} else {
  tibble::tibble(
    sample_name = character(0),
    c_conc = numeric(0)
  )
}
```

#### Sample IDs

```{r}
df_show |> 
  knitr::kable(
    row.names = FALSE,
    col.names = labelled::get_variable_labels(df_show),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

#### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
v <- df_show$c_conc
table(v) |> 
  knitr::kable(
    col.names = c(labelled::get_label_attribute(v), "Number of Samples"), 
    align = "rr"
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Quality control samples

Aliquots of some samples were measured as technical replicates for quality control.
The number of the replicates were as below.

```{r}
show$extract_qc_samples_to_list(raw_se) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Class", "Number of Replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Features

The data contained `r nrow(raw_se)` features.
Among them, `r sum(util$std_type(raw_se) != "")` features were targeted. 
The rest were not targeted.

For those targeted features, the standard type was specified.
The number of features for each type is as follows:

```{r}
v <- util$std_type(raw_se)
replace(v, v == "", "Not targeted") |>
  show$kable_number_of(what = "Features")
```

The distribution of the features across m/z and retention time is shown below.

```{r}
#| label: 'distribution-features'
#| fig-width: 16
#| fig-height: 10
colors <- local({
  x <- unique(util$std_type(raw_se))
  n_non_target <- sum(x == "")
  col <- scales::hue_pal()(length(x) - n_non_target)
  col <- rlang::set_names(col, x[x != ""])
  if (n_non_target > 0) {
    c(col, "grey")     # grey for non-targeted
  } else {
    col
  }
})
df1 <- SumExp::row_df(raw_se)
ggplot(df1) +
  geom_point(aes(x = .rt, y = mz, color = .std_type), 
             alpha = ifelse(df1$.std_type == "", 0.1, 1)) +
  labs(
    x = show$label_if_has(df1$.rt), 
    y = show$label_if_has(df1$mz),
    color = show$label_if_has(df1$.std_type),
  ) +
  scale_color_manual(values = colors)
```

The distribution of m/z of the features is shown below.
The bin width is set to 0.5.

```{r}
#| label: 'distribution-mz'
#| fig-width: 16
#| fig-height: 8
df1 <- SumExp::row_df(raw_se)
ggplot(df1) +
  geom_histogram(aes(x = mz), binwidth = 0.5) +
  labs(
    x = show$label_if_has(df1$mz), 
    y = "Number of features"
  )
```

## Distribution of the values

* Overall samples and targeted features

As shown below, the values are distributed in a skewed manner.

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 16 
#| fig-height: 6

p <- SumExp::ggplot(raw_se) +
  geom_histogram(aes(x = raw), bins = 50)
gridExtra::grid.arrange(   # Two ggplots next to each other
  p + labs(title = "Linear scale", x = "Values"),
  p + scale_x_log10() + labs(title = "Logarithmic scale", x = "Values"),
  ncol = 2
)
```
