---
title: Calibration - curves
params:
  norm_method: loess_norm
  interactive_plot: false
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
# The code in this chunk is NOT evaluated here when knitting this file directly.
# It is included for DEBUGGING purposes only. 
params <- rlang::list2(
  norm_method = "loess_norm", 
  interactive_plot = FALSE,
  project_title = "Test",
) 
source("code/reports/chapters/_internal-shared_setup_chunk.R")
```

## Calibration curve per chemical

The calibration curve of each of the chemicals is presented below.

```{r}
#| purl: true
# Whether the calibration curve is fitted to the log-transformed data
log_scale <- to_report[["params"]][["log_calibration"]]

calcurve_lst <- lapply(lst_proc, \(ea) ea$calcurve)
# To keep the order
seq_feat_id <- rownames(calcurve_lst[[1]])  # All batches have the same features

calcurve_to_show_df <- lapply(calcurve_lst, SumExp::as_tibble) |> 
  dplyr::bind_rows(.id = "Batch") 
# Copy labels from the first batch
for (ii in names(calcurve_lst[[1]])) {
  labelled::label_attribute(calcurve_to_show_df[[ii]]) <- 
    labelled::label_attribute(calcurve_lst[[1]][[ii]])
}
for (ii in names(SumExp::col_df(calcurve_lst[[1]]))) {
  l <- labelled::label_attribute(SumExp::col_df(calcurve_lst[[1]])[[ii]])
  labelled::label_attribute(calcurve_to_show_df[[ii]]) <- l
}
calcurve_to_show_df <- calcurve_to_show_df |>
  dplyr::rename(.cc_pt = util$spiked_conc_pts_name) |>    # Rename for clarity
  dplyr::mutate(
    "c_type" = dplyr::case_when(
      .cc_pt >= min_c_conc & .cc_pt <= max_c_conc ~ "Within limits",
      .cc_pt == lod ~ "LOD",
      .cc_pt == 0 ~ "Cal0",
      TRUE ~ "Out-of-range"
    )
  )
# Split the data by feature
lst_calcurve_to_show_df <- split(calcurve_to_show_df, calcurve_to_show_df$.row_id)
# Keep the order of features as in the data
lst_calcurve_to_show_df <- local({
  nms <- names(lst_calcurve_to_show_df)
  lst_calcurve_to_show_df[match(seq_feat_id, nms, nomatch = 0)]
})
```

```{r}
# Adjust the normalization method text
txt_norm <- show$label_if_has(calcurve_to_show_df[[params$norm_method]], params$norm_method) |>
  sub("normalized", "normalization", x = _, fixed = TRUE)
```

The peak areas were normalized by the internal standards applying the `r txt_norm`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#calibration-curve-limits).

```{r}
#| label: 'calibration-curve-by-chemical-legend'
#| fig-width: 10
#| fig-height: 1
# Plot calibration curves
calcurve_plot <- function(dfm, y = params$norm_method) {
  dfm <- dfm |>
    dplyr::mutate(
      # Label for the points
      txt = paste0("inj: ", injection_order, ", conc: ", .cc_pt, 
                   "\nName: ", sample_name,
                   "\nint: ", signif(.data[[y]], 6))
    )
  facet_fill <- if (any(!dfm$has_proper_range)) "lightcoral" else "lightgray"
  ggplot(dfm, aes(x = .cc_pt, y = .data[[y]], shape = Batch)) +
    suppressWarnings(geom_point(aes(color = c_type, text = txt))) +
    labs(
      x = show$label_if_has(dfm$.cc_pt, "Calibrant Concentration"),
      y = show$label_if_has(dfm[[y]], "Signal"),
      color = "Concentration\ncategory"
    ) +
    scale_color_manual(
      values = c(
        "Within limits" = "green4",
        "Out-of-range" = "gray",
        "Cal0" = "red",
        "LOD" = "blue"
      )
    ) +
    facet_wrap(~ feature_name, scales = "free", ncol = 1) +
    theme(
      strip.background = element_rect(fill = facet_fill),
    )
}
# Show legend only at the top
p <- calcurve_plot(calcurve_to_show_df)
leg <- ggpubr::get_legend(p, position = "top")
ggpubr::as_ggplot(leg)
```

```{r}
#| results: 'asis'
p_by_feature <- lapply(lst_calcurve_to_show_df, \(dfm) {
  p <- calcurve_plot(dfm)
  if (log_scale) {
    p <- p + scale_x_log10() + scale_y_log10()
  }
  p <- p + theme(legend.position = "none")
  if (params$interactive_plot) {
    plotly::ggplotly(p, tooltip = c("text")) |> 
      plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 5))
  } else {
    p
  }
})

sapply(names(p_by_feature), \(ii) {
  rlang::chr(
    "```{r}",
    "#| label: 'calcurve-cal_pts-{{ii}}'",
    "#| fig-width: 7",
    "#| fig-height: 7",
    "#| out-width: '70%'",
    "p_by_feature${{ii}}",
    "```",
  ) |> 
    knitr::knit_expand(text = _)
}) |>
  knitr::knit_child(text = _, quiet = TRUE) |> 
  knitr::asis_output()    # print `asis`
```

## Calibration of the chemicals

```{r}
#| purl: true
concn_lst <- lapply(lst_proc, \(ea) {
  se <- ea$concn
  se[SumExp::row_df(se)$has_proper_range, ]
})
concn_row_df <- lapply(concn_lst, \(se) {
  SumExp::row_df(se) |>
    tibble::as_tibble(rownames = "feature_id")
}) |>
  purrr::list_rbind(names_to = "Batch")
concn_col_df <- lapply(concn_lst, \(se) {
  SumExp::col_df(se) |>
    tibble::as_tibble(rownames = "sample_id")
}) |> 
  purrr::list_rbind(names_to = "Batch")
calib_lst <- lapply(calcurve_lst, \(se) {
  se[SumExp::row_df(se)$has_proper_range, ]
})
```

### Calibration models

```{r}
weight_method <- to_report$params$weight
if (weight_method == "largestR2") {
  n_models <- 6
  weight_txt <- "three different weights, $1$, $\\frac{1}{x}$ and $\\frac{1}{x^2}$"
} else {
  n_models <- 2
  weight_txt <- switch(
    weight_method,
    "1" = "constant",
    "1_div_x" = "$\\frac{1}{x}$",
    "1_div_x2" = "$\\frac{1}{x^2}$"
  )
  weight_txt <- paste(weight_txt, "weight")
}
```

The calibration curve is fitted to the `r switch(params$norm_method, "loess_norm" = "LOESS", "closest_norm" = "Closest RT")` normalized data using `r n_models` different models.
Linear models and quadratic models are fitted to the data with `r weight_txt`.
The best model is selected based on the $R^2$ value.
Preferring simpler models, the measures of explained variation of quadratic models were penalized by 0.01.

The distribution of $R^2$ values after penalty by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s_adj <- concn_row_df |>
  dplyr::mutate(R2s_adj = lapply(calcurve_model, \(m) m$R2s_adj), .keep = "unused") |>
  tidyr::unnest_longer(R2s_adj, values_to = "R2", indices_to = "Model")

ggplot(R2s_adj) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(paste(R^2, " with penalty"))) +
  facet_wrap(~Batch)
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chemical'
#| results: 'markup'
concn_row_df |>
  dplyr::mutate(best_model = sapply(calcurve_model, \(m) m$best_model_name)) |>
  janitor::tabyl(Batch, best_model)
```

### Calibration curves

The concentration of the sample classes, `r knitr::combine_words(unique(concn_col_df$Class), before = "\x60")`,
are computed using the calibration curve.
After the computation, the following steps are applied to the data:

* Any negative concentration is adjusted to zero.
* Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
* Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
* Any concentration whose signal is below the average signal of LLOQ is adjusted to the half of LLOQ.
* Any concentration whose signal is above the average signal of LLOQ is adjusted to the LLOQ.
* Any chemical with values below LLOQ in all samples is removed from the data.

#### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

```{r}
#| results: 'asis'
qmd_out <- list()
for (batch_id in names(concn_lst)) {

  concn_se <- concn_lst[[batch_id]]
  calib_se <- calib_lst[[batch_id]]

  # Show legend only at the top
  qmd_out <- c(
    qmd_out,
    rlang::chr(
      "##### Batch: {{batch_id}}",
      "```{r}",
      "#| label: 'calibration-curve-with-all-samples-legend-{{batch_id}}'",
      "#| fig-width: 10",
      "#| fig-height: 1",
      "p <- show$ggplot_calcurve_samples(", 
      "  concn_se, calib_se, MAT_ID_IN_CALIB, COLORS_OF_CLASSES, log_scale", 
      ")",
      "leg <- ggpubr::get_legend(p, position = 'top')",
      "ggpubr::as_ggplot(leg)",
      "```",
    ) |> 
      knitr::knit_expand(text = _) |>
      knitr::knit_child(text = _, quiet = TRUE)
  )

  # Show calibration curves for each feature
  lst_concn_se <- SumExp::split_rows(concn_se, rownames(SumExp::row_df(concn_se)))
  # Keep the order of features as in the data
  lst_concn_se <- local({
    nms <- names(lst_concn_se)
    lst_concn_se[match(seq_feat_id, nms, nomatch = 0)]
  })

  p_by_feature <- lapply(lst_concn_se, \(se) {
    p <- show$ggplot_calcurve_samples_facet(se, calib_se, MAT_ID_IN_CALIB, COLORS_OF_CLASSES, 
                                            ncol = 1, log_scale = log_scale)
    p <- p + theme(legend.position = "none")
    if (params$interactive_plot) {
      plotly::ggplotly(p, tooltip = c("text")) |> 
        plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 5))
    } else {
      p
    }
  })

  qmd_out <- c(
    qmd_out,
    sapply(names(p_by_feature), \(ii) {
      rlang::chr(
        "```{r}",
        "#| label: 'calcurve-with-all-samples-{{batch_id}}-{{ii}}'",
        "#| fig-width: 7",
        "#| fig-height: 7",
        "#| out-width: '70%'",
        "p_by_feature${{ii}}",
        "```",
      ) |> 
        knitr::knit_expand(text = _)
    }) |>
      knitr::knit_child(text = _, quiet = TRUE)
  )
}
knitr::asis_output(paste(qmd_out, collapse = "\n"))    # print `asis`
```

#### All except `QC` samples

```{r}
#| purl: true
to_show_excl_qc_lst <- lapply(concn_lst, \(se) util$exclude_ctrl_smpl_cat(se, "QC"))
classes_excl_qc_blank <- lapply(to_show_excl_qc_lst, \(se) SumExp::col_df(se)[["Class"]]) |> 
  unlist() |> 
  unique()
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

```{r}
#| label: 'calibration-curve-with-all-samples-excl-qc'
#| results: 'asis'
qmd_out <- list()
for (batch_id in names(to_show_excl_qc_lst)) {

  to_show_excl_qc <- to_show_excl_qc_lst[[batch_id]]
  calib_se <- calib_lst[[batch_id]]

  qmd_out <- c(
    qmd_out,
    rlang::chr(
      "##### Batch: {{batch_id}}",
      "```{r}",
      "#| label: 'calibration-curve-with-all-samples-excl-qc-legend-{{batch_id}}'",
      "#| fig-width: 10",
      "#| fig-height: 1",
      "p <- show$ggplot_calcurve_samples(",
      "  to_show_excl_qc, calib_se, MAT_ID_IN_CALIB, COLORS_OF_CLASSES, log_scale", 
      ")",
      "leg <- ggpubr::get_legend(p, position = 'top')",
      "ggpubr::as_ggplot(leg)",
      "```",
    ) |> 
      knitr::knit_expand(text = _) |>
      knitr::knit_child(text = _, quiet = TRUE)
  )

  if (nrow(to_show_excl_qc) == 0) {
    qmd_out <- c(
      qmd_out,
      "No chemicals to show in the calibration curve.\n"
    )
  } else {
    # Keep the order of features as in the data
    to_show_excl_qc <- local({
      nms <- rownames(to_show_excl_qc)
      to_show_excl_qc[match(seq_feat_id, nms, nomatch = 0), ]
    })
    # Split by each row
    n_chem <- nrow(to_show_excl_qc)
    n_cols <- 2    # Number of columns in the plot (facets)
    n_rows <- ceiling(n_chem / n_cols)    # Number of rows in the plot
    i_grp <- rep(seq_len(n_rows - 1), each = n_cols)
    i_grp <- c(i_grp, rep(n_rows, (n_chem - 1) %% n_cols + 1))
    i_grp <- sprintf("g%04d", i_grp)
    lst_show_se <- SumExp::split_rows(to_show_excl_qc, i_grp)

    p_by_row <- lapply(lst_show_se, \(se) {
      p <- show$ggplot_calcurve_samples_facet(se, calib_se, MAT_ID_IN_CALIB, COLORS_OF_CLASSES, 
                                              ncol = n_cols, log_scale = log_scale)
      p <- p + show$geom_best_model_text(se)
      p + theme(legend.position = "none")
      # plotly::ggplotly(
      #   p + theme(legend.position = "none"),
      #   tooltip = c("text")
      # ) |> 
      #   plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 50))
    })

    qmd_out <- c(
      qmd_out,
      sapply(names(p_by_row), \(ii) {
        n_chem_in_row <- nrow(lst_show_se[[ii]])
        fig_width <- n_chem_in_row * 4.5    # Width 
        out_width <- paste0(fig_width * 10, "%")
        rlang::chr(
          "```{r}",
          "#| label: 'calcurve-with-all-samples-excl-qc-{{batch_id}}-{{ii}}'",
          "#| fig-width: {{fig_width}}",
          "#| fig-height: 4.8",
          "#| out-width: '{{out_width}}'",
          "p_by_row[['{{ii}}']]",
          "```",
        ) |> 
          knitr::knit_expand(text = _)
      }) |>
        knitr::knit_child(text = _, quiet = TRUE)
    )
  }
}
knitr::asis_output(paste(qmd_out, collapse = "\n"))
```

## Removal of the chemicals with no measurement samples

If all values of non-QC samples were below LLOQ, the chemical was removed from the final reports.
After the removal, the number of reported chemicals is,

```{r}
#| label: 'number-of-chemicals-after-removal'
#| results: 'markup'
lapply(concn_lst, \(se) {
  tibble::tibble(
    "Number of chemicals" = apply(se[["conc"]], 1, \(r) any(!is.na(r))) |> 
      sum()
  )
}) |>
  purrr::list_rbind(names_to = "Batch")
```
