---
title: LOESS fit along RT
---

```{r}
#| label: 'setup-norm-loess'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  io = projlib/check_io_exist,       # Check input/output files
  patchwork,           # Combine ggplot objects. `+`
  ggplot2[...],        # Too long to write and usually specific enough
  ggfortify,           # PCA plot (autoplot)
)
# Read `params.yml` if `params` not provided
if (! exists("params")) params <- yaml::read_yaml("params.yml")
# Input/Output files
FILE <- list(
  i = rlang::list2(
    qc = msdial$get_sumexp_file_name(params, "qc_steps"),   # Intermediate state of the data
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

qc_steps <- readRDS(FILE$i$qc)
overall_sumexp <- qc_steps[["Normalized"]]

sample_id_name_tbl <- SumExp::col_df(overall_sumexp) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)

# Find the range of RT values
all_rts <- SumExp::row_df(overall_sumexp)$rt
overall_rt_range <- range(all_rts)
```

```{r}
#| purl: true
excl_cat <- qc_steps[["excluded categories in normalization"]]
```

For each feature, the average of all samples excluding `r knitr::combine_words(excl_cat, before = "\x60")` samples was calculated after log-transformation.
The deviation of each sample from the average was calculated as normalization factor.
The data of the sample are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) features along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all features. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all features.
The normalized data are calculated by subtracting the normalization factor from the log-transformed data.

Randomly selected 3 samples are shown below.
```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 12

# Plot the normalized and original data for randomly selected 3 samples excluding those not
# used in normalization
to_excl <- SumExp::col_df(overall_sumexp)$contr_cat %in% excl_cat
i_sel <- sample(c(1:ncol(overall_sumexp))[!to_excl], 3)
i_sel <- setNames(i_sel, colnames(overall_sumexp)[i_sel])

# To include LOESS smooth line
loess_curve_df <- local({
  x <- seq(rg[1], rg[2], length.out = 100)   # X-axis (retention time)
  out <- lapply(i_sel, function(ii) predict(loess_fit[[ii]], newdata = x)) |> 
    as.data.frame() |> 
    dplyr::mutate(rt = x) |>
    tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")
    # Attach sample names
  dplyr::left_join(out, sample_id_name_tbl, by = "sample_id")
}, list(
  loess_fit = qc_steps[["LOESS fit"]],
  rg = overall_rt_range,
  i_sel = i_sel,
  sample_id_name_tbl = sample_id_name_tbl
))

# Deviation from the mean
dev_from_mean <- local({
  # Log-transform the data
  mat <- log(se[["raw"]][, !to_excl, drop = FALSE])
  mat[!is.finite(mat)] <- NA    # nan = -Inf - -Inf. Replace -Inf by `log` with NA
  # Overall mean of every feature
  mean_log <- rowMeans(mat, na.rm = TRUE)
  out_se <- se[, i_sel]
  for (ii in c("raw", "loess_norm")) {
    out_se[[ii]] <- sweep(log(out_se[[ii]]), 1, mean_log, FUN = "-")
  }
  out_se
}, list(se = overall_sumexp, i_sel = i_sel, to_excl = to_excl))

df_to_show <- SumExp::as_tibble(dev_from_mean) |>
  dplyr::rename(Raw = raw, Normalized = loess_norm) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(
    Type = ifelse(std_type == "", "non-target", std_type),
    Type = factor(Type, levels = c("IS", "Quant", "non-target"))
  ) |> 
  dplyr::arrange(desc(Type)) # To show IS on top

# Plot the normalized and original data
ggplot(df_to_show, aes(x = rt, y = Value)) + 
  geom_point(aes(color = Data, shape = Type, size = Type, alpha = Type)) +
  geom_line(data = loess_curve_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = labelled::get_label_attribute(rt),
    y = "log(Values) - log(avg)"
  ) +
  scale_shape_manual(values = c("IS" = 15, "Quant" = 21, "non-target" = 20)) +
  scale_size_manual(values = c("IS" = 3, "Quant" = 3, "non-target" = 1)) +
  scale_alpha_manual(values = c("IS" = 1, "Quant" = 1, "non-target" = 0.1)) +
  facet_wrap(~ sample_name, scales = "free_y", ncol = 3)
```

The line in the graph represents the LOESS fit.
The points at the bottom boundary are the **0**s, which were not normalized by this process.
Those zeros were kept as zeros in the normalized data.

#### Evaluation of normalization

The normalized data are evaluated by the changes of relative standard deviation (RSD) of the quantification standard features of the QC samples.
```{r}
#| label: 'rsd-eval'
#| out-width: '80%'

# Quantification standard features of QC samples
qc_se <- show$extract_quant_qc(overall_sumexp)
# Calculate RSD of the quantification standard samples
labs <- SumExp::name_labs(overall_sumexp)
labs <- labs[labs %in% c("raw", "loess_norm")]

rsd_qstd <- proc$calc_rsd_qstd(qc_se, labs) |> 
  dplyr::mutate(
    sign = ifelse(raw > loess_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
show$ggplot_rsdp_metab(rsd_qstd, labs) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'rsd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

Bar charts of the values of standard features are compared for raw and normalized data.

```{r}
#| fig-width: 14
#| fig-height: 6
#| out-width: '100%'
#| results: 'markup'

labs <- SumExp::name_labs(overall_sumexp)
df_show <- overall_sumexp[quote(std_type %in% c("Quant", "IS")), ] |> 
  SumExp::as_tibble() |> 
  tidyr::pivot_longer(cols = c(raw, loess_norm), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(Data = factor(Data, levels = labs, labels = names(labs)))

df_show |> 
  dplyr::filter(std_type == "Quant", contr_cat == "") |> 
  ggplot() +
  geom_col(aes(x = injection_order, y = Value, fill = Class)) +
  labs(title = "Quantification standards", x = "Injection order") +
  facet_wrap(~ Data, scales = "free_y") + 
  scale_y_log10()

df1 <- df_show |> 
  dplyr::filter(std_type == "IS") |> 
  dplyr::mutate(sum_value = sum(Value, na.rm = TRUE), .by = c(injection_order, Data))

# plotly::ggplotly(
ggplot(df1, aes(x = injection_order)) +
  geom_col(aes(y = Value, fill = Class)) +
  labs(title = "Internal standards", x = "Injection order") +
  geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
  facet_wrap(~ Data, scales = "free_y")
# )
# plotly::ggplotly(
ggplot(df1, aes(x = injection_order)) +
  geom_col(aes(y = Value, fill = feature_name)) +
  labs(
    title = "Internal standards", 
    x = labelled::get_label_attribute(df1$injection_order),
    fill = labelled::get_label_attribute(df1$feature_name),
  ) +
  geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
  facet_wrap(~ Data, scales = "free_y")
# )
```

Principal component analysis (PCA) is performed to evaluate the normalization.

```{r}
#| label: 'pca'
#| fig-width: 14
#| fig-height: 6
#| out-width: '100%'
# PCA

se <- overall_sumexp[quote(std_type == "IS"), ]
mat_ids <- setNames(nm = c("raw", DATA_ID))
pca_res <- lapply(mat_ids, \(.x) {
  mat <- se[[.x]]
  stats::prcomp(t(mat), scale. = TRUE)
})
se_df <- SumExp::col_df(se)
# Plot
p1 <- ggplot2::autoplot(pca_res$raw, data = se_df, colour = "Class") +
  labs(title = "Raw")
p2 <- ggplot2::autoplot(pca_res$loess_norm, data = se_df, colour = "Class") +
  labs(title = "LOESS Normalized")
# Requires patchwork package
p1 + p2 + patchwork::plot_layout(ncol = 2)
```

