---
title: Calibration
---

```{r}
#| label: 'setup-calibration'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'
DATA_ID <- setNames("loess_norm", nm = "LOESS normalization")

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/proc,       # Preprocessing functions
  projlib/show,       # Functions to present data
  ggplot2[...],        # Too long to write and usually specific enough
)
# Read `params.yml` if `params` not provided
if (! exists("params")) params <- yaml::read_yaml("params.yml")
# Input/Output files
FILE <- list(
  i = rlang::list2(
    qc = msdial$get_sumexp_file_name(params, "qc_steps"),   # Intermediate state of the data
  )
)
# Check if input files and output directories exist
box::use(projlib/io)
io$check_io_exist(FILE)

qc_steps <- readRDS(FILE$i$qc)
# Conversion tables
sample_id_name_tbl <- qc_steps[["sample_id_name_tbl"]]
chem_id_name_tbl <- qc_steps[["chem_id_name_tbl"]]
```

## Calibration curve per chemical

The calibration curve of each of randomly selected 9 chemicals is presented below.

```{r}
#| purl: true
calcurve_se <- qc_steps[[DATA_ID]][["calcurve_se within range"]]

set.seed(7579)
# Randomly selected 9 chemicals to present in calibration curve plots
i_9chem_to_show <- local({
  ii <- c(1:nrow(calcurve_se))
  ii <- ii[SumExp::row_df(calcurve_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve_to_show_df <- SumExp::as_tibble(calcurve_se[i_9chem_to_show, ]) |> 
  tidyr::drop_na(all_of(unname(DATA_ID)))        # Drop concentrations outside the range
```

The peak areas were normalized by the internal standards applying the `r names(DATA_ID)`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

### Linear scale

```{r}
#| label: 'calcurve-per-chemical-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
x <- "c_conc"
p <- ggplot(calcurve_to_show_df, aes(x = .data[[x]], y = .data[[DATA_ID]])) +
  geom_point() +
  labs(
    x = labelled::label_attribute(calcurve_to_show_df[[x]]), 
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ chem_name, scales = "free", ncol = 3)
print(p)
```

### Log scale

```{r}
#| label: 'calcurve-per-chemical-log'
#| fig-width: 12
#| fig-height: 11
p + scale_x_log10() + scale_y_log10()
```

## Calibration of the chemicals

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.

The distribution of $R^2$ values by different models for the calibration curves is compared below.

```{r}
#| purl: true
calcurve_models <- qc_steps[[DATA_ID]][["calcurve_models"]]
```

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s <- lapply(calcurve_models, \(m) m$R2s)
R2s <- R2s[!is.na(R2s)] |> 
  dplyr::bind_rows(.id = "chem_id") |> 
  tidyr::pivot_longer(cols = -chem_id, names_to = "Model", values_to = "R2")

ggplot(R2s) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(R^2))
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chem'
#| results: 'markup'
table(sapply(calcurve_models, \(m) m$best_model_name))
```

```{r}
#| purl: true
concn_se <- qc_steps[[DATA_ID]][["concn_se with conc"]]
# Table of the samples to show with the calibration curve
to_show_se <- concn_se[i_9chem_to_show, ]
```

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
Any negative concentration is adjusted to zero.

### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
show$ggplot_calcurve_samples_facet(to_show_se, calcurve_se, calcurve_models, DATA_ID) +
  labs(y = "Peak area (Normalized)")
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
show$ggplot_calcurve_samples_facet(to_show_se, calcurve_se, calcurve_models, DATA_ID) +
  scale_x_log10() +
  scale_y_log10()
```

### All except `QC` and `Blank` samples

```{r}
classes_excl_qc_blank <- local({
  unique(df$Class[!df$proc_cat %in% c("QC", "Blank")])
}, list(df = SumExp::col_df(concn_se)))
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11

# Table for the selected calibration curve models
models_to_show_tbl <- purrr::imap_dfr(calcurve_models[i_9chem_to_show], \(m, chem_id) {
  tibble::tibble(
    chem_id = chem_id,
    best_model_name = m$best_model_name,
    R2 = max(m$R2s),
    txt = paste0(best_model_name, "\n R2=", round(R2, 3)),
  )
}) |> 
  dplyr::left_join(chem_id_name_tbl, by = "chem_id")
geom_best_model_name <- geom_text(         # Display the best model name
  aes(x = Inf, y = 0, label = txt),
  data = models_to_show_tbl,
  hjust = 1,
  vjust = 0,
  size = 4,
  color = "red3"
)

to_show_se[, quote(Class %in% classes_excl_qc_blank)] |> 
  show$ggplot_calcurve_samples_facet(calcurve_se, calcurve_models, DATA_ID) +
  geom_best_model_name
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
to_show_se[, quote(Class %in% classes_excl_qc_blank)] |> 
  show$ggplot_calcurve_samples_facet(calcurve_se, calcurve_models, DATA_ID) +
  geom_best_model_name +
  scale_x_log10() +
  scale_y_log10()
```
