```{r}
#| purl: true
internal_std_se <- qc_steps[["internal std. raw"]]
```

There are `r nrow(internal_std_se)` internal standard (IS) features in the data.
Those features are listed below, in the order of average retention time.

```{r}
#| purl: true
SumExp::row_df(internal_std_se) <- SumExp::row_df(internal_std_se) |> 
  dplyr::mutate(
    num_zeros = proc$count_zeros_per_feature(internal_std_se[["raw"]]),
    rsd = proc$compute_rsd_per_feature(internal_std_se[["raw"]]),
  )
# Table of internal standard features
kable_internal_std <- function(se) {
  df <- SumExp::row_df(se) |> 
    dplyr::select(feature_name, rt, mz, sn_ratio, rsd, num_zeros) |> 
    dplyr::mutate(
      num_zeros = ifelse(num_zeros == 0, "", num_zeros) |> 
        labelled::copy_labels_from(num_zeros)
    )
  knitr::kable(
    df, 
    row.names = FALSE,
    col.names = labelled::get_variable_labels(df),
    align = "lrrrrr"
  ) |> 
    kableExtra::kable_styling(full_width = FALSE)
}
kable_internal_std(internal_std_se)
```

### Outliers

Samples deviating too much from others may be considered as outliers based on the values of internal standard (IS) features, which were supposed to be the same across all the samples. 
For each internal standard, the samples with values greater than 3 standard deviations from the mean of all samples after log-transformation ($log(1 + x)$) were marked as outlying samples for the feature.
In the other way, the feature was marked as an outlier for the sample.
When the number of outlying IS features of a sample exceeds 20% of the total number of IS, the sample was considered as outliers and eventually excluded from the data.

```{r}
#| label: 'outliers-table'
is_outlier <- qc_steps[["internal std. outlying samples"]]
n_out_istd <- qc_steps[["internal std. number of outliers per sample"]]

if (all(! is_outlier)) {
  cat("No outliers were identified.\n")
} else {
  cat("The following samples were identified as outliers.\n\n")
  stopifnot(identical(names(is_outlier), names(n_out_istd)))
  tb <- tibble::as_tibble(SumExp::col_df(internal_std_se), rownames = "sample_id")
  show_tb <- tibble::as_tibble(n_out_istd, rownames = "sample_id") |> 
    labelled::set_variable_labels(
      "value" = "Number of Outlying IS",
      "sample_id" = "Sample ID"
    ) |>
    # Retrieve the sample_name
    dplyr::left_join(tb, by = "sample_id") |> 
    dplyr::filter(value > 0) |> 
    dplyr::select(sample_name, Class, sample_type, value)
  knitr::kable(
    show_tb,
    col.names = labelled::get_variable_labels(show_tb),
    align = "lccr"
  ) |> 
    kableExtra::kable_classic(full_width = FALSE)
}
```

### Failed internal standards

Internal standard (IS) features that satisfy the following conditions were considered as failed IS:

* One or more samples have zero values.

```{r}
#| out-width: 100%

is_failed_istd <- qc_steps[["internal std. failed IS"]]

box_scatter_plot_of_one_feature <- function(se, ii, widths) {
  feature_name <- SumExp::row_df(se)$feature_name[ii]
  se_tbl <- SumExp::as_tibble(se[ii, ])
  # Box plot
  p1 <- ggplot(se_tbl) +
    geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
    labs(title = feature_name, y = "Peak area") +
    theme(legend.position = "none") +   # No legend for fill
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Rotate x-axis labels
    scale_y_log10()
  # Scatter plot across injection order
  p2 <- ggplot(se_tbl) +
    geom_point(aes(x = injection_order, y = raw, color = Class)) +
    labs(title = "",  y = "", x = "Injection order") +
    theme(legend.position = "none") +   # No legend for fill
    # No y-axis labels
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_log10()
  # Requires patchwork package
  p1 + p2 + patchwork::plot_layout(widths = widths)
}

if (any(is_failed_istd)) {
  cat("The following internal standard features were identified as failed IS,",
      "which were removed.\n\n")
  # Table of failed internal standard features
  print(kable_internal_std(failed_istd))
  # Box and scatter plots of the failed internal standard features 
  for(ii in 1:nrow(failed_istd)) {
    print(box_scatter_plot_of_one_feature(failed_istd, ii, widths = c(0.3, 0.7)))
  }
} else {
  cat("No failed internal standard features were identified.\n")
}
```

```{r}
if (any(is_outlier) || any(is_failed_istd)) {
  internal_std_se <- internal_std_se[!is_failed_istd, !is_outlier]
  SumExp::row_df(internal_std_se) <- SumExp::row_df(internal_std_se) |> 
    dplyr::mutate(
      num_zeros = proc$count_zeros_per_feature(internal_std_se[["raw"]]),
      rsd = proc$compute_rsd_per_feature(internal_std_se[["raw"]]),
    )
  # Update the table of internal standard features
  cat("\n\nAfter excluding the outliers and failed internal standards,", 
      "the table of internal standard features is as follows.\n\n")
  kable_internal_std(internal_std_se)
}
```

### Distribution of the values of individual IS features

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 4
#| out-width: 100%
# Internal standard features to show
se_show <- internal_std_se
for(ii in 1:nrow(se_show)) {
  print(box_scatter_plot_of_one_feature(se_show, ii, widths = c(0.3, 0.7)))
}
```

### Raw RSD%

The relative standard deviation (RSD%) of the raw values of the internal standard features are calculated.

#### `QC` samples

As these `QC` samples are technical replicates, the RSD% values of the samples are expected to be low.

```{r}
#| label: 'rsd-internal-standards-by-qc-samples'
# Divide the internal standard features by quality control sample
# Calculate RSD% per internal standard feature and per quality control sample
df1 <- show$extract_qc_samples_to_list(internal_std_se) |> 
  sapply(\(se) apply(se[["raw"]], 1, \(.x) round(proc$rsd_perc(.x), 2)))
stopifnot(identical(rownames(df1), rownames(internal_std_se)))
rownames(df1) <- SumExp::row_df(internal_std_se)$feature_name
knitr::kable(df1, caption = "RSD% of internal standard features by quality control sample") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### By sample classes

```{r}
se <- internal_std_se[, quote(contr_cat != "QC")]
df1 <- sapply(setNames(nm = unique(SumExp::col_df(se)$Class)), function(cls) {
  se <- se[, SumExp::col_df(se)$Class == cls]
  apply(se[["raw"]], 1, \(.x) round(proc$rsd_perc(.x), 2))
})
stopifnot(identical(rownames(df1), rownames(se)))
rownames(df1) <- SumExp::row_df(se)$feature_name
knitr::kable(df1, caption = "RSD% of internal standard features by sample classes") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
