---
title: Calibration
---

```{r}
#| label: 'setup-calibration'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'
# The preprocessed data used for calibration. Default is `loess_norm_blk`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm_blk"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/show,       # Functions to present data
  ggplot2[...],        # Too long to write and usually specific enough
  io = projlib/check_io_exist,       # Check input/output files
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
# For consistency throughout the report
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$contr_cat)

# Input/Output files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
feature_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(feature_id, feature_name)
```

## Calibration curve per feature

The calibration curve of each of randomly selected 9 features is presented below.

```{r}
#| purl: true
calcurve_se <- qc_steps$calibration[[DATA_ID]][["calcurve_se within range"]]

set.seed(7579)
# Randomly selected 9 features to present in calibration curve plots
i_9feature_to_show <- local({
  ii <- c(1:nrow(calcurve_se))
  ii <- ii[SumExp::row_df(calcurve_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve_to_show_df <- SumExp::as_tibble(calcurve_se[i_9feature_to_show, ]) |> 
  tidyr::drop_na(all_of(DATA_ID))        # Drop concentrations outside the range
```

The peak areas were normalized by the internal standards applying the `r show$label_if_has(calcurve_to_show_df[[DATA_ID]], DATA_ID)`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

### Linear scale

```{r}
#| label: 'calcurve-per-feature-linear'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
x <- "c_conc"
p <- ggplot(calcurve_to_show_df, aes(x = .data[[x]], y = .data[[DATA_ID]])) +
  geom_point() +
  labs(
    x = labelled::label_attribute(calcurve_to_show_df[[x]]), 
    y = "Peak area (Normalized)"
  ) +  
  facet_wrap(~ feature_name, scales = "free", ncol = 3)
print(p)
```

### Log scale

```{r}
#| label: 'calcurve-per-feature-log'
#| fig-width: 12
#| fig-height: 11
p + scale_x_log10() + scale_y_log10()
```

## Calibration of the features

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.

The distribution of $R^2$ values by different models for the calibration curves is compared below.

```{r}
#| purl: true
calcurve_models <- qc_steps$calibration[[DATA_ID]][["calcurve_models"]]
```

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s <- lapply(calcurve_models, \(m) m$R2s)
R2s <- R2s[!is.na(R2s)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(R^2))
```

The count of features favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-feature'
#| results: 'markup'
table(sapply(calcurve_models, \(m) m$best_model_name))
```

```{r}
#| purl: true
concn_se <- qc_steps$calibration[[DATA_ID]][["concn_se with conc"]]
# Table of the samples to show with the calibration curve
to_show_se <- concn_se[i_9feature_to_show, ]
# Models to be used for the calibration curve
cc_models <- lapply(calcurve_models, \(.x) .x$inv_model)
```

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
Any negative concentration is adjusted to zero.

### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each feature.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
show$ggplot_calcurve_samples_facet(to_show_se,
                                   calcurve_se,
                                   cc_models,
                                   DATA_ID,
                                   COLORS_OF_CLASSES) +
  labs(y = "Peak area (Normalized)")
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-samples-log'
#| fig-width: 12
#| fig-height: 11
show$ggplot_calcurve_samples_facet(to_show_se,
                                   calcurve_se,
                                   cc_models,
                                   DATA_ID,
                                   COLORS_OF_CLASSES) +
  scale_x_log10() +
  scale_y_log10()
```

### All except `QC` samples

```{r}
classes_excl_qc_blank <- local({
  unique(df$Class[!df$contr_cat %in% c("QC")])
}, list(df = SumExp::col_df(concn_se)))
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each feature.

#### Linear scale

```{r}
#| label: 'calibration-curve-with-normsample'
#| fig-width: 12
#| fig-height: 11

# Table for the selected calibration curve models
models_to_show_tbl <- purrr::imap_dfr(calcurve_models[i_9feature_to_show], \(m, feature_id) {
  tibble::tibble(
    feature_id = feature_id,
    best_model_name = m$best_model_name,
    R2 = max(m$R2s),
    txt = paste0(best_model_name, "\n R2=", round(R2, 3)),
  )
}) |> 
  dplyr::left_join(feature_id_name_tbl, by = "feature_id")
geom_best_model_name <- geom_text(         # Display the best model name
  aes(x = Inf, y = 0, label = txt),
  data = models_to_show_tbl,
  hjust = 1,
  vjust = 0,
  size = 4,
  color = "red3"
)

to_show_se[, quote(Class %in% classes_excl_qc_blank)] |> 
  show$ggplot_calcurve_samples_facet(calcurve_se, cc_models, DATA_ID, COLORS_OF_CLASSES) +
  geom_best_model_name
```

#### Log scale

```{r}
#| label: 'calibration-curve-with-normsample-log'
#| fig-width: 12
#| fig-height: 11
to_show_se[, quote(Class %in% classes_excl_qc_blank)] |> 
  show$ggplot_calcurve_samples_facet(calcurve_se, cc_models, DATA_ID, COLORS_OF_CLASSES) +
  geom_best_model_name +
  scale_x_log10() +
  scale_y_log10()
```

## Remove the features with no measurement samples

```{r}
concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
```

If all values of non-QC samples are below LLOQ, the feature is removed from the data.
After the removal, the number of remaining features is `r nrow(concn_se)`.

