
The closest internal standard features to each feature will be identified based on the retention time.
The measured peak area of each feature will be divided by the peak area of the closest internal standard feature.

When a compound target class is specified, the closest internal standard will be selected from the same compound target class.

#### Evaluation of normalization

```{r}
#| label: 'evaluation-normalization-closest-istd'
#| out-width: '80%'

# Quantification standard samples
qc_se <- show$extract_quant_qc(norm_se)

mat_ids <- c(mat_id_for_norm, "closest_norm")
rsd_qstd <- show$calc_rsd_qstd(qc_se, mat_ids)
rsd_qstd <- rsd_qstd |> 
  dplyr::mutate(
    sign = .data[[mat_id_for_norm]] > closest_norm,
    sign = ifelse(sign, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
show$ggplot_rsdp_metab(rsd_qstd, mat_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'evaluation-normalization-closest-istd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

#### Samples by run order

```{r}
#| purl: true
has_vol_norm <- !is.null(norm_se[["vol_norm"]])
mat_ids_to_compare <- c("raw", if (has_vol_norm) "vol_norm", "closest_norm")
```

Bar charts of the values of internal standard (`IS`) features are compared for raw and normalized data.
The size of the bars represent the peak area before and after normalization. 
Please note the injection orders may have gaps, as they are obtained information from the source file.

The bars are colored by the given `Class` of the samples

```{r}
#| fig-width: !expr if (has_vol_norm) 20 else 14
#| fig-height: 6
#| out-width: '100%'
df_show <- norm_se[util$is_internal_std(norm_se), ] |> 
  show$df_for_injection_order(mat_ids_to_compare)

plotly::ggplotly(
  show$ggplot_col_injection_order(df_show, fill = Class) +
    labs(title = "Internal standards") +
    scale_fill_manual(values = COLORS_OF_CLASSES)
)
```

The bars are colored by the internal standards and stacked for each sample.

```{r}
#| fig-width: !expr if (has_vol_norm) 20 else 14
#| fig-height: 6
#| out-width: '100%'
plotly::ggplotly(
  show$ggplot_col_injection_order(df_show, fill = .std_type) +
    labs(title = "Internal standards", fill = "Target class")
)
```

#### Principal component analysis

Principal component analysis (PCA) is performed to evaluate the normalization.

```{r}
#| label: 'pca-normalization-closest-istd'
#| fig-width: !expr if (has_vol_norm) 20 else 14
#| fig-height: 6
#| out-width: '100%'
# PCA

se <- norm_se[util$is_internal_std(norm_se), ]
# Plot
p_lst <- lapply(mat_ids_to_compare, \(.x) {
  out <- show$ggplot_pca(se, .x, colour = "Class")
  if (is.null(out)) return(NULL)
  out + scale_color_manual(values = COLORS_OF_CLASSES)
})
# Requires patchwork package
Reduce("+", p_lst) + patchwork::plot_layout(ncol = length(p_lst))
```

