## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The samples with a concentration of `0` are separated as the zero samples.
The data of the zeros is used to calculate lower-limits of quantification (LLOQ) and lower-limits of detection (LLOD).

The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
v <- SumExp::col_df(norm_se)$c_conc
v <- v[!is.na(v)] |> 
  labelled::copy_labels_from(v)
table(v) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c(labelled::get_label_attribute(v), "Number of Samples")
    # col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Concentration ranges

```{r}
#| purl: true
limit_df <- qc_steps$calibration[[DATA_ID]][["calcurve conc ranges"]]
```


The lower-limit of quantification (LLOQ) and lower-limit of detection (LLOD) are calculated using zero concentration samples.
The LLOD is defined as the lowest concentration at which the average signal of calibration samples exceeds three times the mean of the zero concentration samples. 
If the mean of the zeros is zero, LLOD is the lowest concentration having non-zero value. 
Similarly, the LLOQ is set at ten times the mean of the zero concentration samples.
If the calibration samples at the LLOQ concentration exhibit high variability (relative standard deviation, RSD, > 20%), the LLOQ is adjusted to the lowest higher concentration that satisfies RSD ≤ 20%.
Concentrations below this LLOQ are excluded from the calibration curve.

Quite a few concentrations in the calibration curve samples of a feature can be too far from the range observed in the samples for measurement. 
When there are sufficient data points (≥ 5 concentrations), the maximum concentration of the calibration curve is accordingly adjusted. 
Concentrations that produce peak areas after normalization exceeding ten times the **highest** peak area of non-`CalCurve` and non-`QC` sample categories are excluded.

If the LLOQ is set so high that fewer than three concentration points remain, that feature is excluded from the calibration curve.
As a result, `r english::Words(sum(limit_df$to_exclude))` features have been removed from the calibration curve.

The concentration range for each feature is detailed below.

```{r}
#| label: 'calcurve-conc-range-table'

show_df <- limit_df |> 
  tibble::as_tibble(rownames = "feature_id") |> 
  dplyr::mutate(
    Note = dplyr::case_when(
      to_exclude ~ "Excluded", 
      llod == lloq ~ "LLOD = LLOQ",
      TRUE ~ ""
    ),
    LLOD = scales::number(llod),
    LLOQ = ifelse(is.na(lloq), "All < LLOQ or RSD% > 20%", scales::number(lloq)),
    max_c_conc = ifelse(is.na(max_c_conc), "Not enough range", scales::number(max_c_conc)) |> 
      labelled::copy_labels_from(max_c_conc)
  ) |>
  dplyr::left_join(feature_id_name_tbl, by = "feature_id") |> 
  dplyr::select(feature_name, LLOD, LLOQ, max_c_conc, Note)

knitr::kable(
  show_df,
  col.names = labelled::get_variable_labels(show_df, null_action = "fill"),
  caption = "Concentration range of the calibration curve samples",
  align = "lrrrr",
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px") |> 
  # Color the rows that are marked in Note
  kableExtra::row_spec(which(show_df$Note == "LLOD = LLOQ"), background = "lightblue") |> 
  kableExtra::row_spec(which(show_df$Note == "Excluded"), background = "lightcoral")
```
