## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The samples with a concentration of `0` are separated as the zero samples.
The data of the zeros is used to calculate lower-limits of quantification (LLOQ) and lower-limits of detection (LLOD).

The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
table(calcurve_se$c_conc) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c("Concentration (ng/ml)", "Number of samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Concentration ranges

The lower-limit of quantification (LLOQ) and lower-limit of detection (LLOD) are calculated using the zero concentration samples.
The LLOD has been determined as the lowest concentration, for which the average signal of calibration samples were larger than 3 times of the mean of zero concentration samples. 
Likewise, the LLOQ has been determined with 10 times of the mean of zero concentration samples.
When the values of calibration samples for the LLOQ concentration had high variability, relative standard deviation (RSD) > 20%, the LLOQ has been chosen as the lowest higher concentration that satisfies RSD < 20%.
All concentrations below the LLOQ are excluded from the calibration curve.

Quite a few concentrations of calibration curve samples of a chemical may be too far from the range of the measured samples. 
When there are enough (≥ 5) number of concentration points, the maximum concentration of the calibration curve is adjusted. 
The concentrations were excluded for which the peak areas after normalization were more than 10 times higher than the **maximum** peak area of the samples excluding `CalCurve` and `QC` categories.

When the LLOQ is so high that the number of concentration points left is smaller than 3, the chemical is excluded from the calibration curve.
`r english::Words(sum(rowData(calcurve_se)$to_exclude))` chemicals are excluded from the calibration curve.

Concentration range of each chemical is presented below.

```{r}
#| label: 'calcurve-conc-range-table'

rowData(calcurve_se) |> 
  tibble::as_tibble() |> 
  dplyr::mutate(
    llod = scales::number(llod),
    lloq = ifelse(is.na(lloq), "All < LLOQ or RSD% > 20%", scales::number(lloq)),
    max_c_conc = ifelse(is.na(max_c_conc), "Not enough range", scales::number(max_c_conc)),
    to_exclude = ifelse(to_exclude, "Excluded", "")
  ) |>
  dplyr::select(chem_name, llod, lloq, max_c_conc, to_exclude) |>
  knitr::kable(
    caption = "Concentration range of the calibration curve samples",
    col.names = c("Chemical",
                  "LLOD (ng/ml)",
                  "LLOQ (ng/ml)",
                  "Maximum concentration (ng/ml)",
                  "Excluded?"), 
    align = "lrrrr",
  ) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px")
```
