## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
v <- SumExp::col_df(norm_se)$c_conc
v <- v[!is.na(v)] |> 
  labelled::copy_labels_from(v)
table(v) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c(labelled::get_label_attribute(v), "Number of Samples")
    # col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Calibration curve limits

```{r}
#| purl: true
limit_df <- calib_interm_data[["calcurve conc ranges"]]
```

* Concentration point ranges

In order to fit a calibration curve to a meaningful range of concentration, the range is adjusted.
The lower-limit of the range is chosen following the process described below. 
Concentration points below the limit are excluded from the calibration curve.

1. Begin with the lowest concentration point (e.g. `Cal_0`) in the calibration sample set with a detected signal (or peak area) greater than zero.
2. Examine the signal corresponding to the next highest concentration point to determine if it is statistically greater than the lowest non-zero signal, using a t-test for significance.
3. If no significant difference is found, continue to test the next higher concentration point sequentially until a significant increase in signal is identified.
4. If the calibration samples at the concentration with significantly higher signal exhibit high variability (relative standard deviation, RSD, > 20%), the lower-limit is adjusted to the lowest higher concentration point that satisfies RSD ≤ 20%.

As quite a few concentration points in the calibration curve samples of a chemical can be too far from the range observed in the samples for quantification (non-`CalCurve` and non-`QC` sample categories), the upper-limit of the calibration range is set. 
When there are sufficient number (≥ 5) of concentration points, that concentrations that produce signal after normalization exceeding ten times the **highest** signal of the quantification samples are excluded.

If the lower-limit is set so high that fewer than three concentration points remain, that chemical is excluded from the calibration curve, because the curve cannot be fitted properly.
As a result, `r english::Words(sum(!limit_df$has_proper_range))` chemicals have been excluded from the calibration.

The concentration range for each chemical is detailed below.

```{r}
#| label: 'calcurve-conc-range-table'

show_df <- limit_df |> 
  tibble::as_tibble(rownames = "feature_id") |>     # To attach `feature_name`
  dplyr::rename(min_c = min_c_conc, max_c = max_c_conc) |>
  dplyr::mutate(
    Note = dplyr::case_when(
      !has_proper_range ~ "Excluded", 
      TRUE ~ ""
    ),
    non_zero_conc = scales::number(non_zero_conc),
    min_c = ifelse(is.na(min_c), "No valid pts.", scales::number(min_c)),
    max_c = dplyr::case_when(
      max_c == 0 ~ "All zero",
      max_c == -9 ~ "No valid max",
      is.na(max_c) ~ "Not enough range",
      TRUE ~ scales::number(max_c)
    )
  ) |>
  dplyr::left_join(feature_id_name_tbl, by = "feature_id") |> 
  dplyr::select("Chemical Name" = feature_name, 
                "Non-zero conc." = non_zero_conc, 
                "Minimum conc." = min_c, 
                "Maximum conc." = max_c, 
                Note)

knitr::kable(
  show_df,
  caption = "Concentration range of the calibration curve samples",
  align = "lrrrl",
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px") |> 
  # Color the rows that are marked in Note
  kableExtra::row_spec(which(show_df$Note == "Excluded"), background = "lightcoral")
```
