

For each chemical, the average of all samples excluding 
`r knitr::combine_words(excluded_category_for_normalization, before = "\x60")`
samples was calculated after log-transformation.
The deviation of each sample from the average was calculated as normalization factor.
The data of the sample are normalized by a LOESS curve fitted to the deviation values of the internal standard (IS) chemicals along RT.
The curve was fitted over the smallest and largest RT values of IS to cover the entire range of RT values of all chemicals. 
The normalization factors of those at the boundaries of IS are used for the boundary RT values of all chemicals.
The normalized data are calculated by subtracting the normalization factor from the log-transformed data.

Randomly selected 3 samples are shown below.
```{r}
#| label: 'normalize-rt-plot'
#| fig-width: 14
#| fig-height: 12

to_excl <- overall_sumexp$proc_cat %in% excluded_category_for_normalization
# Plot the normalized and original data for randomly selected 3 samples
i_sel <- sample(c(1:ncol(overall_sumexp))[!to_excl], 3)

# To include LOESS smooth line
x <- seq(overall_rt_range[1], overall_rt_range[2], length.out = 100)
loess_curve_df <- lapply(setNames(i_sel, colnames(overall_sumexp)[i_sel]), function(ii) {
  predict(loess_fit[[ii]], newdata = x)
}) |> 
  as.data.frame() |> 
  dplyr::mutate(rt = x) |>
  tidyr::pivot_longer(-rt, names_to = "sample_id", values_to = "Value")
loess_curve_df$sample_name <-
  overall_sumexp$given_sample_id[match(loess_curve_df$sample_id, overall_sumexp$sample_id)]

# Log-transform the data
raw_log <- log(assays(overall_sumexp)$raw)
# Overall mean of every chemical
mat <- raw_log[, !to_excl, drop = FALSE]
mat[is.finite(mat) == FALSE] <- NA    # Visualization purpose only
mean_log <- rowMeans(mat, na.rm = TRUE)
# Deviation from the mean
selected_se <- overall_sumexp[, i_sel]
assays(selected_se) <- assays(selected_se) |> 
  lapply(\(x) sweep(log(x), 1, mean_log, FUN = "-"))

df_to_show <- tibble::as_tibble(selected_se) |>
  dplyr::rename(Raw = raw, Normalized = loess_norm, sample_name = given_sample_id) |>
  tidyr::pivot_longer(cols = c(Raw, Normalized), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(
    Type = ifelse(std_type == "", "non-target", std_type),
    Type = factor(Type, levels = c("IS", "Quant", "non-target"))
  ) |> 
  dplyr::arrange(desc(Type)) # To show IS on top

# Plot the normalized and original data
ggplot(df_to_show, aes(x = rt, y = Value)) + 
  geom_point(aes(color = Data, shape = Type, size = Type, alpha = Type)) +
  geom_line(data = loess_curve_df, alpha = 0.5) +
  labs(
    title = "Normalized and original data",
    x = "Average RT", 
    y = "log(Values) - log(avg)"
  ) +
  scale_shape_manual(values = c("IS" = 15, "Quant" = 21, "non-target" = 20)) +
  scale_size_manual(values = c("IS" = 3, "Quant" = 3, "non-target" = 1)) +
  scale_alpha_manual(values = c("IS" = 1, "Quant" = 1, "non-target" = 0.1)) +
  facet_wrap(~ sample_name, scales = "free_y", ncol = 3)
```

The line in the graph represents the LOESS fit.
The points at the bottom boundary are the **0**s, which were not normalized by this process.
Those zeros were kept as zeros in the normalized data.

#### Evaluation of normalization

The normalized data are evaluated by the changes of relative standard deviation (RSD) of the quantification standard chemicals of the QC samples.
```{r}
#| label: 'rsd-eval'
#| out-width: '80%'

# Quantification standard chemicals of QC samples
qc_se <- proc$extract_quant_qc(overall_sumexp)
# Calculate RSD of the quantification standard samples
assay_ids <- setNames(c("raw", "loess_norm"), nm = c("Raw", "LOESS Normalized"))
rsd_qstd <- proc$calc_rsd_qstd(qc_se, assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(raw > loess_norm, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
show$ggplot_rsdp_metab(rsd_qstd, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of chemicals with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'rsd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

Bar charts of the values of standard chemicals are compared for raw and normalized data.

```{r}
#| fig-width: 14
#| fig-height: 6
#| out-width: '100%'

se <- overall_sumexp
df_show <- se[rowData(se)$std_type %in% c("Quant", "IS"), ] |> 
  tibble::as_tibble() |> 
  tidyr::pivot_longer(cols = c(raw, loess_norm), names_to = "Data", values_to = "Value") |> 
  dplyr::mutate(
    Data = factor(Data, levels = c("raw", "loess_norm"), labels = c("Raw", "LOESS Normalized"))
  )
df_show |> 
  dplyr::filter(std_type == "Quant", proc_cat == "") |> 
  ggplot() +
  geom_col(aes(x = injection_order, y = Value, fill = Class)) +
  labs(title = "Quantification standards", x = "Injection order") +
  facet_wrap(~ Data, scales = "free_y") + 
  scale_y_log10()

df1 <- df_show |> 
  dplyr::filter(std_type == "IS") |> 
  dplyr::mutate(sum_value = sum(Value, na.rm = TRUE), .by = c(injection_order, Data))

plotly::ggplotly(
  ggplot(df1, aes(x = injection_order)) +
    geom_col(aes(y = Value, fill = Class)) +
    labs(title = "Internal standards", x = "Injection order") +
    geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
    facet_wrap(~ Data, scales = "free_y")
)
plotly::ggplotly(
  ggplot(df1, aes(x = injection_order)) +
    geom_col(aes(y = Value, fill = chem_name)) +
    labs(
      title = "Internal standards", 
      x = "Injection order",
      fill = "Chemical name"
    ) +
    geom_smooth(aes(y = sum_value), method = "loess", formula = 'y ~ x', se = FALSE) +
    facet_wrap(~ Data, scales = "free_y")
)
```

Principal component analysis (PCA) is performed to evaluate the normalization.
```{r}
#| label: 'pca'
#| fig-width: 14
#| fig-height: 6
#| out-width: '100%'

# PCA
se <- overall_sumexp
se <- se[SummarizedExperiment::rowData(se)$std_type == "IS", ]
assay_ids <- setNames(nm = c("raw", "loess_norm"))
pca_res <- lapply(assay_ids, \(.x) {
  mat <- SummarizedExperiment::assay(se, .x)
  stats::prcomp(t(mat), scale. = TRUE)
})
se_df <- SummarizedExperiment::colData(se)
# Plot
p1 <- ggplot2::autoplot(pca_res$raw, data = se_df, colour = "Class")
p2 <- ggplot2::autoplot(pca_res$loess_norm, data = se_df, colour = "Class")
# Requires patchwork package
p1 + p2 + patchwork::plot_layout(ncol = 2)
```

### Comparison of the normalization methods

The RSD% of the quantification standard samples is calculated for the raw data and the normalized data.

```{r}
#| label: 'norm-method-comparison'
#| out-width: '80%'

assay_ids <- setNames(c("raw", "closest_norm", "loess_norm"), nm = c("Raw", "Closest", "LOESS"))
df_to_show <- proc$extract_quant_qc(overall_sumexp) |> 
  proc$calc_rsd_qstd(assay_ids) |> 
  dplyr::mutate(
    sign = ifelse(loess_norm <= closest_norm, "↓ LOESS", "↓ Closest"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓ LOESS", "↓ Closest", "All 0s"))
  ) 
show$ggplot_rsdp_metab(df_to_show, assay_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of chemicals with lower RSD after each normalization is counted below.

```{r}
#| label: 'rsd-comparison-count'
#| results: 'markup'
# Count the number of chemicals with lower RSD
df_to_show |> 
  janitor::tabyl(QC, sign)
```

The number of compounds above or below 20 RSD% is counted below.

```{r}
#| label: 'rsd-comparison-count-20'
df_to_show |> 
  dplyr::summarize(
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x > 20, na.rm = TRUE), 
      .names = "{.col}_gt_20"
    ),
    dplyr::across(
      dplyr::all_of(unname(assay_ids)), 
      ~ sum(.x <= 20, na.rm = TRUE), 
      .names = "{.col}_eq_lt_20"
    ),
    .by = QC
  ) |>  
  knitr::kable(
    caption = "Number of chemicals with RSD% above or below 20", 
    col.names = c("QC", rep(names(assay_ids), 2))
  ) |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "RSD > 20%" = 3, "RSD ≤ 20%" = 3)) 
```

The RSD% values are presented in the table below.

```{r}
#| label: 'rsd-comparison-table'
df_to_show |> 
  dplyr::left_join(as.data.frame(rowData(overall_sumexp)), by = "chem_id") |>   # Add chemical names
  dplyr::select(QC, chem_name, dplyr::all_of(unname(assay_ids))) |> 
  dplyr::rename(all_of(assay_ids), `Chemical` = chem_name) |> 
  knitr::kable(caption = "RSD% of the quantification standard samples", digits = 2) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |>
  kableExtra::scroll_box(height = "300px")
```
