---
title: Calibration
params:
  data_id: loess_norm_blk
---

```{r}
#| label: 'setup-calibration'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'
# The preprocessed data used for calibration.
DATA_ID <- params$data_id

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/show,       # Functions to present data
  ggplot2[...],        # Too long to write and usually specific enough
  util = projlib/msdial_utils,        # Utility functions for MS-Dial data
  io = projlib/check_io_exist,       # Check input/output files
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
# For consistency throughout the report
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$.ctrl_cat)

# Input/Output files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])
calib_interm_data <- qc_steps[[paste0("calibration/", DATA_ID)]]

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
feature_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(feature_id, feature_name)
```

## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
v <- util$spiked_conc_pts(qc_steps[["normalized - blank"]])
v <- v[!is.na(v)] |> 
  labelled::copy_labels_from(v)
table(v) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c(labelled::get_label_attribute(v), "Number of Samples")
    # col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Calibration curve limits

```{r}
#| purl: true
limit_df <- calib_interm_data[["calcurve conc ranges"]]
```

* LOD and LLOQ

Limit of detection (LOD) and lower-limit of quantification (LLOQ) are identified as described below. 

1. Beginning with the lowest concentration point (e.g. `Cal0`), identify the lowest concentration point with a detected signal (or peak area) greater than zero.
```{r}
cat(switch(
  calib_interm_data[["llox method"]],
  "pt_signal_mean" = 
"2. Identify LOD as the lowest concentration point at which the average signal exceeds three times the mean of the non-zero concentration point.
3. Identify LLOQ as the lowest concentration point satisfying the following two conditions:
    1) The average signal exceeds ten times the mean of the non-zero concentration point.
    2) The relative standard deviation (RSD) of the signal is equal to or less than 20%.",
  "pt_signal_mean_plus_sd" =
"2. Identify LOD as the lowest concentration point at which the average signal exceeds three times the standard deviation of the non-zero concentration point plus the mean of the non-zero point.
3. Identify LLOQ as the lowest concentration point satisfying the following two conditions:
    1) The average signal exceeds ten times the standard deviation of the non-zero concentration point plus the mean of the non-zero point.
    2) The relative standard deviation (RSD) of the signal is equal to or less than 20%."
))
```

* Calibration working range

In order to fit a calibration curve to a meaningful range of concentration, the fitting range is adjusted.
The lower-limit of the range is chosen as the LLOQ.
Concentration points below the limit are excluded from the calibration curve.

As quite a few concentration points in the calibration curve samples of a chemical can be too far from the range observed in the samples for quantification (non-`CalCurve` and non-`QC` sample categories), the upper-limit of the calibration range is set. 
When there are sufficient number (≥ 5) of calibration points, that the points that produce signal after normalization exceeding ten times the **highest** signal of the quantification samples are excluded.

If the lower-limit is set so high that fewer than three concentration points remain, that chemical is excluded from the calibration curve, because the curve cannot be fitted properly.
As a result, `r english::Words(sum(!limit_df$has_proper_range))` chemicals have been excluded from the calibration.

The concentration range for each chemical is detailed below.

```{r}
#| label: 'calcurve-conc-range-table'

show_df <- limit_df |> 
  tibble::as_tibble(rownames = "feature_id") |>     # To attach `feature_name`
  dplyr::rename(min_c = min_c_conc, max_c = max_c_conc) |>
  dplyr::mutate(
    Note = dplyr::case_when(
      !has_proper_range ~ "Excluded", 
      TRUE ~ ""
    ),
    non_zero_conc = scales::number(non_zero_conc),
    lod = ifelse(is.na(lod), "N.A.", scales::number(lod)),
    min_c = ifelse(is.na(min_c), "No valid pts.", scales::number(min_c)),
    max_c = dplyr::case_when(
      max_c == 0 ~ "All zero",
      max_c == -9 ~ "No valid max",
      is.na(max_c) ~ "Not enough range",
      TRUE ~ scales::number(max_c)
    )
  ) |>
  dplyr::left_join(feature_id_name_tbl, by = "feature_id") |> 
  dplyr::select("Chemical Name" = feature_name, 
                "Non-zero conc." = non_zero_conc, 
                "LOD" = lod, 
                "Minimum conc. (LLOQ)" = min_c, 
                "Maximum conc." = max_c, 
                Note)

knitr::kable(
  show_df,
  caption = "Concentration range of the calibration curve samples",
  align = "lrrrrl",
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px") |> 
  # Color the rows that are marked in Note
  kableExtra::row_spec(which(show_df$Note == "Excluded"), background = "lightcoral")
```


## Calibration curve per chemical

The calibration curve of each of randomly selected 9 chemicals is presented below.

```{r}
#| purl: true
cc_se <- calib_interm_data[["calcurve_se within range"]]
set.seed(7579)
# Randomly selected 9 chemicals to present in calibration curve plots
i_9chemical_to_show <- local({
  ii <- c(1:nrow(cc_se))
  ii <- ii[SumExp::row_df(cc_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve9_se <- cc_se[i_9chemical_to_show, ]

calcurve_to_show_df <- local({
  cc9_se <- calib_interm_data[["calcurve_se all range"]][i_9chemical_to_show, ]
  SumExp::as_tibble(cc9_se) |> 
    tidyr::drop_na(all_of(DATA_ID)) |>         # Drop concentrations outside the range
    dplyr::rename(.cc_pt = util$spiked_conc_pts_name) |>    # Rename for clarity
    dplyr::mutate(
      "c_type" = dplyr::case_when(
        .cc_pt == non_zero_conc ~ "First non-zero",
        .cc_pt >= min_c_conc & .cc_pt <= max_c_conc ~ "Within limits",
        TRUE ~ "Out-of-range"
      )
    )
})
```

The peak areas were normalized by the internal standards applying the `r show$label_if_has(calcurve_to_show_df[[DATA_ID]], DATA_ID)`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

- **Full range**

```{r}
#| label: 'calcurve-all-data-pts'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
calcurve_plot <- function(df, y = DATA_ID) {
  ggplot(df, aes(x = .cc_pt, y = .data[[y]])) +
    geom_point(aes(color = c_type)) +
    labs(
      x = show$label_if_has(df$.cc_pt, "Calibrant Concentration"),
      y = show$label_if_has(df[[y]], "Signal"),
      color = "Concentration\ncategory"
    ) +
    scale_color_manual(
      values = c("First non-zero" = "red3", 
                 "Within limits" = "green4",
                 "Out-of-range" = "gray")
    ) +
    facet_wrap(~ feature_name, scales = "free", ncol = 3)
}
calcurve_plot(calcurve_to_show_df)
```

- **Zoom-in (upto minimum $\times$ 5)**

```{r}
#| label: 'calcurve-all-data-pts-zoom-in'
#| fig-width: 12
#| fig-height: 11
calcurve_to_show_df |> 
  dplyr::filter(.cc_pt <= min_c_conc * 5) |> 
  calcurve_plot()
```


## Calibration of the chemicals

```{r}
#| purl: true
concn_se <- calib_interm_data[["concn_se with conc"]]
```

### Calibration models

```{r}
weight_method <- calib_interm_data[["weight method"]]
if (weight_method == "lowestR2") {
  n_models <- 6
  weight_txt <- "three different weights, $1$, $\\frac{1}{x}$ and $\\frac{1}{x^2}$"
} else {
  n_models <- 2
  weight_txt <- switch(
    weight_method,
    "1" = "constant",
    "1/x" = "$\\frac{1}{x}$",
    "1/x2" = "$\\frac{1}{x^2}$"
  )
  weight_txt <- paste(weight_txt, "weight")
}
```

The calibration curve is fitted to the [normalized data](#normalization) using `r n_models` different models.
Linear models and quadratic models are fitted to the data with `r weight_txt`.
The best model is selected based on the $R^2$ value.
Preferring simpler models, the measures of explained variation of quadratic models were penalized by 0.01.

The distribution of $R^2$ values after penalty by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s_adj <- SumExp::row_df(concn_se)$calcurve_model |> 
  lapply(\(m) m$R2s_adj)
R2s_adj <- R2s_adj[!is.na(R2s_adj)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s_adj) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(paste(R^2, " with penalty")))
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chemical'
#| results: 'markup'
SumExp::row_df(concn_se)$calcurve_model |> 
  sapply(\(m) m$best_model_name) |>
  table()
```

### Calibration curves - examples

```{r}
#| purl: true
# Table of the samples to show with the calibration curve
to_show_se <- concn_se[i_9chemical_to_show, ]
```

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
After the computation, the following steps are applied to the data:

* Any negative concentration is adjusted to zero.
* Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
* Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
* Any chemical with values below LLOQ in all samples is removed from the data.

#### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
p <- show$ggplot_calcurve_samples_facet(to_show_se,
                                        calcurve9_se,
                                        DATA_ID,
                                        COLORS_OF_CLASSES) +
  labs(y = "Peak area (Normalized)")
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-samples-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

#### All except `QC` samples

```{r}
classes_excl_qc_blank <- local(envir = list(se = concn_se), {
  not_qc <- util$ctrl_smpl_cat(se) != "QC"
  SumExp::col_df(se)[["Class"]][not_qc] |> 
    unique()
})
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-quant-sample'
#| fig-width: 12
#| fig-height: 11

to_show_excl_qc <- to_show_se[, quote(Class %in% classes_excl_qc_blank)]
p <- show$ggplot_calcurve_samples_facet(to_show_excl_qc, 
                                        calcurve9_se, DATA_ID, COLORS_OF_CLASSES) +
  show$geom_best_model_text(to_show_excl_qc)
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-quant-sample-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

## Removal of the chemicals with no measurement samples

```{r}
concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
```

If all values of non-QC samples were below LLOQ, the chemical was removed from the final reports.
After the removal, the number of reported chemicals is `r nrow(concn_se)`.

