---
title: Calibration
---

```{r}
#| label: 'setup-calibration'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'
# The preprocessed data used for calibration. Default is `loess_norm_blk`.
if (!exists("DATA_ID")) DATA_ID <- "loess_norm_blk"

# Load packages and project local libraries
options(box.path = "code/")           # Path to project local libraries
box::use(
  SumExp,           # Light SummarizedExperiment, `[`
  projlib/msdial,     # Handle MS-Dial files
  projlib/show,       # Functions to present data
  ggplot2[...],        # Too long to write and usually specific enough
  io = projlib/check_io_exist,       # Check input/output files
)
# Get the input file name provided by the user
user_inputs <- msdial$get_user_input()
# To keep constants consistent across reports
constants_yml <- yaml::read_yaml("code/constants.yml")
COLORS <- lapply(constants_yml$COLORS, unlist)    # Lists to named vectors
# For consistency throughout the report
raw_se <- msdial$read_parsed_msdial_data(user_inputs)
COLORS_OF_CLASSES <- show$get_colors_of_classes(raw_se, COLORS$contr_cat)

# Input/Output files
FILE <- list(
  i = rlang::list2(
    # Processed data
    proc = msdial$get_raw_data_file_name(user_inputs, suffix = "proc"),
    # Intermediate status of the data
    qc = msdial$get_raw_data_file_name(user_inputs, suffix = "qc_steps"),
  )
)
# Check if input files and output directories exist
io$check_io_exist(FILE)

# Load the intermediate data during QC
qc_steps <- readRDS(FILE$i$qc)
stopifnot("Preprocessing NOT completed." = qc_steps[["Preprocessing Completed"]])
calib_interm_data <- qc_steps[[paste0("calibration/", DATA_ID)]]

# Conversion tables between IDs (syntactically acceptable) and names (human-readable)
# These are used for presenting tables and figures in reports
sample_id_name_tbl <- SumExp::col_df(raw_se) |> 
  tibble::rownames_to_column("sample_id") |> 
  dplyr::select(sample_id, sample_name)
feature_id_name_tbl <- SumExp::row_df(raw_se) |> 
  tibble::rownames_to_column("feature_id") |> 
  dplyr::select(feature_id, feature_name)
```

## Calibration curve per chemical

The calibration curve of each of randomly selected 9 chemicals is presented below.

```{r}
#| purl: true
cc_se <- calib_interm_data[["calcurve_se within range"]]
set.seed(7579)
# Randomly selected 9 chemicals to present in calibration curve plots
i_9chemical_to_show <- local({
  ii <- c(1:nrow(cc_se))
  ii <- ii[SumExp::row_df(cc_se)$std_type != "IS"]
  sample(ii, 9)
})
calcurve9_se <- cc_se[i_9chemical_to_show, ]

calcurve_to_show_df <- local({
  cc9_se <- calib_interm_data[["calcurve_se all range"]][i_9chemical_to_show, ]
  SumExp::as_tibble(cc9_se) |> 
    tidyr::drop_na(all_of(DATA_ID)) |>         # Drop concentrations outside the range
    dplyr::mutate(
      "c_type" = dplyr::case_when(
        c_conc == non_zero_conc ~ "First non-zero",
        c_conc >= min_c_conc & c_conc <= max_c_conc ~ "Within limits",
        TRUE ~ "Out-of-range"
      )
    )
})
```

The peak areas were normalized by the internal standards applying the `r show$label_if_has(calcurve_to_show_df[[DATA_ID]], DATA_ID)`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#concentration-ranges).

- **Full range**

```{r}
#| label: 'calcurve-all-data-pts'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
calcurve_plot <- function(df, x = "c_conc", y = DATA_ID) {
  ggplot(df, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(aes(color = c_type)) +
    labs(
      x = labelled::label_attribute(df[[x]]),
      y = labelled::label_attribute(df[[y]]),
      color = "Concentration\ncategory"
    ) +
    scale_color_manual(
      values = c("First non-zero" = "red3", 
                 "Within limits" = "green4",
                 "Out-of-range" = "gray")
    ) +
    facet_wrap(~ feature_name, scales = "free", ncol = 3)
}
calcurve_plot(calcurve_to_show_df)
```

- **Zoom-in (upto minimum $\times$ 5)**

```{r}
#| label: 'calcurve-all-data-pts-zoom-in'
#| fig-width: 12
#| fig-height: 11
calcurve_to_show_df |> 
  dplyr::filter(c_conc <= min_c_conc * 5) |> 
  calcurve_plot()
```


## Calibration of the chemicals

```{r}
#| purl: true
concn_se <- calib_interm_data[["concn_se with conc"]]
```

### Calibration models

The calibration curve is fitted to the [normalized data](#normalization) using 6 different models.
Linear models and quadratic models are fitted to the data with three different weights, $1$, $\frac{1}{x}$ and $\frac{1}{x^2}$.
The best model is selected based on the $R^2$ value.
Preferring simpler models, the measures of explained variation of quadratic models were penalized by 0.01.

The distribution of $R^2$ values after penalty by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s_adj <- SumExp::row_df(concn_se)$calcurve_model |> 
  lapply(\(m) m$R2s_adj)
R2s_adj <- R2s_adj[!is.na(R2s_adj)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s_adj) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(paste(R^2, " with penalty")))
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chemical'
#| results: 'markup'
SumExp::row_df(concn_se)$calcurve_model |> 
  sapply(\(m) m$best_model_name) |>
  table()
```

### LOD/LLOQ

The limit of detection (LOD) is first calculated from a signal perspective by multiplying the standard deviation of the signals at the lowest calibration concentration point that exhibits a non-zero response by three and then adding this product to the average signal of the concentration point. 
The LOD in terms of concentration is then derived using the chosen calibration curve model.

In a similar manner, the lower limit of quantification (LLOQ) is calculated by employing a factor of 10 instead of three, with all other steps remaining consistent with the LOD determination.

The LOD and LLOQ of each chemical is presented below.

```{r}
#| label: 'calcurve-llodq-table'

show_df <- SumExp::row_df(concn_se) |> 
  dplyr::mutate(dplyr::across(c(lod, lloq), ~ scales::number(.x))) |>
  dplyr::select(`Chemical name` = feature_name, LOD = lod, LLOQ = lloq)

knitr::kable(
  show_df,
  caption = "LOD and LLOQ of the chemicals",
  align = "lrr",
  row.names = FALSE
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px")
```

### Calibration curves - examples

```{r}
#| purl: true
# Table of the samples to show with the calibration curve
to_show_se <- concn_se[i_9chemical_to_show, ]
```

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
After the computation, the following steps are applied to the data:

* Any negative concentration is adjusted to zero.
* Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
* Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
* Any chemical with values below LLOQ in all samples is removed from the data.

#### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
p <- show$ggplot_calcurve_samples_facet(to_show_se,
                                        calcurve9_se,
                                        DATA_ID,
                                        COLORS_OF_CLASSES) +
  labs(y = "Peak area (Normalized)")
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-samples-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

#### All except `QC` samples

```{r}
classes_excl_qc_blank <- local({
  unique(df$Class[!df$contr_cat %in% c("QC")])
}, list(df = SumExp::col_df(concn_se)))
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-quant-sample'
#| fig-width: 12
#| fig-height: 11

to_show_excl_qc <- to_show_se[, quote(Class %in% classes_excl_qc_blank)]
p <- show$ggplot_calcurve_samples_facet(to_show_excl_qc, 
                                        calcurve9_se, DATA_ID, COLORS_OF_CLASSES) +
  show$geom_best_model_text(to_show_excl_qc)
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-quant-sample-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

## Removal of the chemicals with no measurement samples

```{r}
concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
```

If all values of non-QC samples were below LLOQ, the chemical was removed from the final reports.
After the removal, the number of reported chemicals is `r nrow(concn_se)`.

