
```{r}
#| purl: true
internal_std_se <- qc_steps[["internal std. before norm"]]
box::use(projlib/proc[count_zeros_per_feature])
```

There are `r nrow(internal_std_se)` internal standard (IS) features in the data.
Those features are listed below, in the order of average retention time.

```{r}
#| purl: true

SumExp::row_df(internal_std_se) <- SumExp::row_df(internal_std_se) |> 
  dplyr::mutate(
    num_zeros = count_zeros_per_feature(internal_std_se[[mat_id_for_norm]]),
    rsd = show$compute_rsd_per_feature(internal_std_se[[mat_id_for_norm]]),
  )
# Table of internal standard features
kable_internal_std <- function(se) {
  df <- SumExp::row_df(se) |> 
    dplyr::select(feature_name, .rt, mz, sn_ratio, rsd, num_zeros) |> 
    dplyr::mutate(
      num_zeros = ifelse(num_zeros == 0, "", num_zeros) |> 
        labelled::copy_labels_from(num_zeros)
    )
  knitr::kable(
    df, 
    row.names = FALSE,
    col.names = labelled::get_variable_labels(df),
    align = "lrrrrr"
  ) |> 
    kableExtra::kable_styling(full_width = FALSE)
}
```

```{r}
kable_internal_std(internal_std_se)
```

### Failed internal standards

Internal standard (IS) features that satisfy the following conditions were considered as failed IS:

* More than 10% samples have zero values.

```{r}
#| out-width: 100%
box::use(patchwork)           # Combine ggplot objects. `+`

is_failed_istd <- qc_steps[["is failed internal std."]]

box_scatter_plot_of_one_feature <- function(se, ii, widths) {
  feature_name <- SumExp::row_df(se)$feature_name[ii]
  se_tbl <- SumExp::as_tibble(se[ii, ])
  # Box plot
  p1 <- ggplot(se_tbl) +
    geom_boxplot(aes(x = Class, y = raw, fill = Class)) +
    scale_fill_manual(values = COLORS_OF_CLASSES) + 
    labs(title = feature_name, y = "Peak area") +
    theme(legend.position = "none") +   # No legend for fill
    # Rotate x-axis labels
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    scale_y_log10()
  # Scatter plot across injection order
  p2 <- ggplot(se_tbl) +
    geom_point(aes(x = injection_order, y = raw, color = Class)) +
    scale_color_manual(values = COLORS_OF_CLASSES) + 
    labs(title = "",  y = "", x = "Injection order") +
    theme(legend.position = "none") +   # No legend for fill
    # No y-axis labels
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_log10()
  # Requires patchwork package
  p1 + p2 + patchwork::plot_layout(widths = widths)
}

if (any(is_failed_istd)) {
  cat("The following internal standard features were identified as failed IS,",
      "which were removed.\n\n")
  failed_istd <- internal_std_se[is_failed_istd, ]
  # Table of failed internal standard features
  print(kable_internal_std(failed_istd))
  # Box and scatter plots of the failed internal standard features 
  for (ii in seq_len(nrow(failed_istd))) {
    print(box_scatter_plot_of_one_feature(failed_istd, ii, widths = c(0.3, 0.7)))
  }
  # Remove the failed internal standard features
  internal_std_se <- internal_std_se[!is_failed_istd, ]
} else {
  cat("No failed internal standard features were identified.\n")
}
```

Any remaining zero values in an internal standard feature are imputed with the average of the feature values across all samples of the same sample type (i.e. `Sample`, `Standard`, `QC` and `Blank`).

```{r}
#| label: 'outlier-removal'
is_outlier <- qc_steps[["is outlier sample"]]
# Only when outlier removal is performed
if (! is.null(is_outlier)) {
  cat(
    "### Outliers\n\n",
    "Each internal standard (IS) is expected to have consistent values across all samples.\n",
    "After applying a log-transformation ($log(1 + x)$),", 
    "the mean and standard deviation for each IS are calculated.\n",
    "A sample is flagged as an outlier for a given IS", 
    "if its value deviates by more than 3 standard deviations from the IS mean.\n",
    "If more than 20% of IS features in a sample are flagged this way,", 
    "the sample itself is classified as an outlier.\n",
    "The outlier samples are then removed from the data.\n\n"
  )
  # Number of outlying internal standard features per sample
  n_out_istd <- qc_steps[["number of outlier internal std. per sample"]]
  
  if (all(! is_outlier)) {
    cat("No outliers were identified.\n")
  } else {
    cat("The following samples were identified as outliers and excluded.\n\n")
    stopifnot(identical(names(is_outlier), names(n_out_istd)))
    # To retrieve the sample_name, Class, sample_type
    tb <- tibble::as_tibble(SumExp::col_df(internal_std_se), rownames = "sample_id")
    show_tb <- tibble::as_tibble(n_out_istd, rownames = "sample_id") |> 
      dplyr::mutate(is_outlier = is_outlier) |>
      labelled::set_variable_labels(
        "value" = "Number of Outlying IS",
        "sample_id" = "Sample ID"
      ) |>
      # Retrieve the sample_name, Class, sample_type
      dplyr::left_join(tb, by = "sample_id") |> 
      dplyr::filter(is_outlier) |> 
      dplyr::select(sample_name, Class, sample_type, value)
    knitr::kable(
      show_tb,
      col.names = labelled::get_variable_labels(show_tb),
      align = "lccr"
    ) |> 
      kableExtra::kable_classic(full_width = FALSE) |>
      print()
    
    # Remove the outlier samples  
    internal_std_se <- internal_std_se[, !is_outlier]
  }
}
```

```{r}
if (any(is_failed_istd) || any(is_outlier)) {      # any(NULL) == FALSE
  SumExp::row_df(internal_std_se) <- SumExp::row_df(internal_std_se) |> 
    dplyr::mutate(
      num_zeros = count_zeros_per_feature(internal_std_se[[mat_id_for_norm]]),
      rsd = show$compute_rsd_per_feature(internal_std_se[[mat_id_for_norm]]),
    )
  # Update the table of internal standard features
  cat("\n\nAfter excluding failed internal standards", 
      if (is.null(is_outlier)) "" else " and outliers samples",
      ", the table of internal standard features is as follows.\n\n",
      sep = "")
  kable_internal_std(internal_std_se)
}
```

### Distribution of the values of individual IS features

```{r}
#| label: 'distribution-values-internal-standards'
#| fig-width: 13
#| fig-height: 4
#| out-width: 100%
# Internal standard features to show
se_show <- internal_std_se
for (ii in seq_len(nrow(se_show))) {
  print(box_scatter_plot_of_one_feature(se_show, ii, widths = c(0.3, 0.7)))
}
```

### Raw RSD%

The relative standard deviation (RSD%) of the raw values of the internal standard features are calculated.

#### `QC` samples

As these `QC` samples are technical replicates, the RSD% values of the samples are expected to be low.

```{r}
#| label: 'rsd-internal-standards-by-qc-samples'
# Divide the internal standard features by quality control sample
# Calculate RSD% per internal standard feature and per quality control sample
m_rsd <- show$extract_qc_samples_to_list(internal_std_se) |> 
  lapply(function(se) {
    apply(se[[mat_id_for_norm]], 1, \(.x) round(show$rsd_perc(.x), 2))
  }) |>
  do.call(cbind, args = _)     # Make sure the result is a matrix even if there is only one value
# The `rownames` are syntactically valid names, not the original feature names
stopifnot(identical(rownames(m_rsd), rownames(internal_std_se)))
rownames(m_rsd) <- SumExp::row_df(internal_std_se)$feature_name
knitr::kable(m_rsd, caption = "RSD% of internal standard features by quality control sample") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### By sample classes

```{r}
se <- util$exclude_ctrl_smpl_cat(internal_std_se, "QC")
m_rsd <- setNames(nm = unique(SumExp::col_df(se)$Class)) |>
  lapply(function(cls) {
    se <- se[, SumExp::col_df(se)$Class == cls]
    apply(se[[mat_id_for_norm]], 1, \(.x) round(show$rsd_perc(.x), 2))
  }) |>
  do.call(cbind, args = _)     # Make sure the result is a matrix even if there is only a vector
stopifnot(identical(rownames(m_rsd), rownames(se)))
rownames(m_rsd) <- SumExp::row_df(se)$feature_name
knitr::kable(m_rsd, caption = "RSD% of internal standard features by sample classes") |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
