
Given file: `` `r SumExp::metadata(overall_sumexp)$file_name` `` 

The given file contains data of `r nrow(overall_sumexp)` chemicals and `r ncol(overall_sumexp)` samples.

## Samples

### By class

There are `r dplyr::n_distinct(SumExp::col_df(overall_sumexp)$Class)` classes of samples in the data.
The number of samples for each class is as follows:

```{r}
show$kable_number_of(SumExp::col_df(overall_sumexp)$Class)
```

### Special categories for preprocessing

There are `r length(table(SumExp::col_df(overall_sumexp)$proc_cat, exclude = ""))` special sample categories for in the data.
The data of those are important in the following processing steps.
The number of samples for each category is as follows:

```{r}
show$kable_number_of(SumExp::col_df(overall_sumexp)$proc_cat, exclude = "")
```

#### Calibration curve samples

The samples for the calibration curve are identified by the sample type `Standard` and class `CalCurve`.
The concentration is extracted from the provided sample ID.
The IDs are in the format `Cal_[[:digit:]-]+` with optional numbers and/or characters before and after of it. 
Concentration is extracted from the numeric part in the format after the replacement of `-` with `.`.

##### Sample IDs

```{r}
# Present the sample IDs for the calibration curve
df_show <- SumExp::col_df(overall_sumexp) |> 
  dplyr::filter(proc_cat == "CalCurve") |> 
  dplyr::arrange(c_conc) |> 
  dplyr::select(sample_name, c_conc)
df_show |> 
  knitr::kable(
    row.names = FALSE,
    col.names = labelled::get_variable_labels(df_show),
    format.args = list(scientific = FALSE)
  ) |> 
  kableExtra::kable_classic(full_width = FALSE) |> 
  kableExtra::scroll_box(height = "160px")
```

##### The number of samples per concentration

```{r}
#| label: 'calcurve-samples-summary'
# Present the number of samples in the calibration curve
v <- df_show$c_conc
table(v) |> 
  knitr::kable(
    col.names = c(labelled::get_label_attribute(v), "Number of Samples"), 
    align = "rr"
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

#### Quality control samples

Aliquots of some samples were measured as technical replicates for quality control.
The number of the replicates were as below.

```{r}
show$extract_qc_samples_to_list(overall_sumexp) |> 
  sapply(ncol) |> 
  knitr::kable(col.names = c("Class", "Number of Replicates")) |> 
  kableExtra::kable_styling(full_width = FALSE)
```

## Chemicals

The data contained `r nrow(overall_sumexp)` chemicals.
Among them, `r sum(SumExp::row_df(overall_sumexp)$std_type != "")` chemicals were targeted. 
The rest were not targeted.

For those targeted chemicals, the standard type was specified.
The number of chemicals for each type is as follows:

```{r}
show$kable_number_of(SumExp::row_df(overall_sumexp)$std_type, what = "Chemicals", exclude = "")
```

## Distribution of the values

* Overall samples and targeted chemicals

As shown below, the values are distributed in a skewed manner.

```{r}
#| label: 'distribution-values-overall-linear'
#| fig-width: 16 
#| fig-height: 6
#| out-width: '100%'

p <- SumExp::ggplot(overall_sumexp) +
    geom_histogram(aes(x = raw), bins = 50)
gridExtra::grid.arrange(   # Two ggplots next to each other
  p + labs(title = "Linear scale", x = "Values"),
  p + scale_x_log10() + labs(title = "Logarithmic scale", x = "Values"),
  ncol = 2
)
```
