---
title: Calibration
params:
  norm_method: loess_norm
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
```

## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
v <- util$spiked_conc_pts(qc_steps[["normalized - blank"]])
v <- v[!is.na(v)] |> 
  labelled::copy_labels_from(v)
table(v) |> 
  knitr::kable(
    align = c("r", "r"),
    col.names = c(labelled::get_label_attribute(v), "Number of Samples")
    # col.names = c("Concentration (ng/ml)", "Number of Samples")
  ) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Calibration curve limits

```{r}
#| purl: true
limit_df <- calib_interm_data[["calcurve conc ranges"]]
```

* LOD and LLOQ

Limit of detection (LOD) and lower-limit of quantification (LLOQ) are identified as described below. 

1. Beginning with the lowest concentration point (e.g. `Cal0`), identify the lowest concentration point with a detected signal (or peak area) greater than zero.
```{r}
cat(switch(
  calib_interm_data[["llox method"]],
  "pt_signal_mean" = 
"2. Identify LOD as the lowest concentration point at which the average signal exceeds three times the mean of the non-zero concentration point.
3. Identify LLOQ as the lowest concentration point satisfying the following two conditions:
    1) The average signal exceeds ten times the mean of the non-zero concentration point.
    2) The relative standard deviation (RSD) of the signal is equal to or less than 20%.",
  "pt_signal_mean_plus_sd" =
"2. Identify LOD as the lowest concentration point at which the average signal exceeds three times the standard deviation of the non-zero concentration point plus the mean of the non-zero point.
3. Identify LLOQ as the lowest concentration point satisfying the following two conditions:
    1) The average signal exceeds ten times the standard deviation of the non-zero concentration point plus the mean of the non-zero point.
    2) The relative standard deviation (RSD) of the signal is equal to or less than 20%."
))
```

* Calibration working range

In order to fit a calibration curve to a meaningful range of concentration, the fitting range is adjusted.
The lower-limit of the range is chosen as the LLOQ.
Concentration points below the limit are excluded from the calibration curve.

As quite a few concentration points in the calibration curve samples of a chemical can be too far from the range observed in the samples for quantification (non-`CalCurve` and non-`QC` sample categories), the upper-limit of the calibration range is set. 
When there are sufficient number (≥ 5) of calibration points, that the points that produce signal after normalization exceeding ten times the **highest** signal of the quantification samples are excluded.

If the lower-limit is set so high that fewer than three concentration points remain, that chemical is excluded from the calibration curve, because the curve cannot be fitted properly.
As a result, `r english::Words(sum(!limit_df$has_proper_range))` chemicals have been excluded from the calibration.

The concentration range for each chemical is detailed below.

```{r}
#| label: 'calcurve-conc-range-table'

show_df <- limit_df |> 
  tibble::as_tibble(rownames = "feature_id") |>     # To attach `feature_name`
  dplyr::rename(min_c = min_c_conc, max_c = max_c_conc) |>
  dplyr::mutate(
    Note = dplyr::case_when(
      !has_proper_range ~ "Excluded", 
      TRUE ~ ""
    ),
    non_zero_conc = scales::number(non_zero_conc),
    lod = ifelse(is.na(lod), "N.A.", scales::number(lod)),
    min_c = ifelse(is.na(min_c), "No valid pts.", scales::number(min_c)),
    max_c = dplyr::case_when(
      max_c == 0 ~ "All zero",
      max_c == -9 ~ "No valid max",
      is.na(max_c) ~ "Not enough range",
      TRUE ~ scales::number(max_c)
    )
  ) |>
  dplyr::left_join(feature_id_name_tbl, by = "feature_id") |> 
  dplyr::select("Chemical Name" = feature_name, 
                "First calibration pt with signal" = non_zero_conc, 
                "LOD" = lod, 
                "Minimum conc. (LLOQ)" = min_c, 
                "Maximum conc." = max_c, 
                Note)

knitr::kable(
  show_df,
  caption = "Concentration range of the calibration curve samples",
  align = "lrrrrl",
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px") |> 
  # Color the rows that are marked in Note
  kableExtra::row_spec(which(show_df$Note == "Excluded"), background = "lightcoral")
```


