---
title: Calibration
params:
  norm_method: loess_norm
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
params <- list(norm_method = "loess_norm", project_title = "Test") 
source("code/reports/chapters/_internal-shared_setup_chunk.R")  # For "Run Cell" functionality
```

```{r}
#| purl: true
calcurve_lst <- lst_proc |> 
  lapply(\(x) x$calcurve)
```

## Calibration curve samples

The data of the samples for the calibration curve is extracted from the quantification samples after the outlier removal and normalization.
The number of samples in the calibration curve after outlier removal is presented below.

```{r}
#| label: 'calcurve-samples-summary-calibration'
# Present the number of samples in the calibration curve
for (ii in names(calcurve_lst)) {
  se <- calcurve_lst[[ii]]
  v <- util$spiked_conc_pts(se)
  v <- v[!is.na(v)] |> 
    labelled::copy_labels_from(v)
  table(v) |> 
    knitr::kable(
      align = c("r", "r"),
      col.names = c(labelled::get_label_attribute(v), "Number of Samples"),
      caption = paste("Batch", ii),
      # col.names = c("Concentration (ng/ml)", "Number of Samples")
    ) |> 
    kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
}
```

## Calibration curve limits

```{r}
#| purl: true
limit_df <- lapply(calcurve_lst, \(se) {
  SumExp::row_df(se)[, c("min_c_conc", "max_c_conc", "lod", "lloq", "has_proper_range")] |>
    tibble::as_tibble(rownames = "feature_id")
}) |> 
  purrr::list_rbind(names_to = "batch")
```

* LOD and LLOQ

```{r}
cat(switch(
  to_report$params$llox_method,
  "pt_signal_mean" = paste(
    "LOD and LLOQ were defined as the lowest concentrations at which the mean signal exceeded 3×",
    "or 10× the mean signal of the blank (zero concentration point) respectively."
  ),
  "pt_signal_mean_plus_sd" = paste(
    "LOD and LLOQ were defined as the lowest concentrations at which the mean signal exceeded the",
    "mean signal of the blank (zero concentration point) plus 3× or 10× its standard deviation,",
    "respectively."
  )
))
if (to_report$params$use_rsd20) {
  cat(" Additionally, LLOQ required a signal RSD ≤ 20%.")
}

```

* Calibration working range

In order to fit a calibration curve to a meaningful range of concentration, the fitting range is adjusted.
The lower-limit of the range is chosen as the LLOQ.
Concentration points below the limit are excluded from the calibration curve.

As quite a few concentration points in the calibration curve samples of a chemical can be too far from the range observed in the samples for quantification (non-`CalCurve` and non-`QC` sample categories), the upper-limit of the calibration range is set. 
When there are sufficient number (≥ 3) of calibration points, that the points that produce signal after normalization exceeding the **highest** signal of the quantification samples are excluded except the lowest one.

If the lower-limit is set so high that fewer than three concentration points remain, that chemical is excluded from the calibration curve, because the curve cannot be fitted properly.
As a result, `r english::Words(sum(!limit_df$has_proper_range))` chemicals have been excluded from the calibration.

The concentration range for each chemical is detailed below.

```{r}
#| label: 'calcurve-conc-range-table'

show_df <- limit_df |> 
  dplyr::rename(min_c = min_c_conc, max_c = max_c_conc) |>
  dplyr::mutate(
    Note = dplyr::case_when(
      !has_proper_range ~ "Excluded", 
      TRUE ~ ""
    ),
    lod = ifelse(is.na(lod), "N.A.", as.character(lod)),
    min_c = ifelse(is.na(min_c), "No valid pts.", as.character(min_c)),
    max_c = dplyr::case_when(
      max_c == 0 ~ "All zero",
      max_c == -9 ~ "No valid max",
      is.na(max_c) ~ "Not enough range",
      TRUE ~ as.character(max_c)
    )
  ) |>
  dplyr::left_join(feature_id_name_tbl, by = "feature_id") |> 
  dplyr::select(
    "Batch" = batch,
    "Chemical Name" = feature_name, 
    "LOD" = lod, 
    "Minimum conc. (LLOQ)" = min_c, 
    "Maximum conc." = max_c, 
    Note,
  )

knitr::kable(
  show_df,
  caption = "Concentration range of the calibration curve samples",
  align = "lrrrrl",
) |> 
  kableExtra::kable_classic(lightable_options = "striped", full_width = FALSE) |> 
  kableExtra::scroll_box(height = "350px") |> 
  # Color the rows that are marked in Note
  kableExtra::row_spec(which(show_df$Note == "Excluded"), background = "lightcoral")
```
