---
title: Calibration - curves
params:
  norm_method: loess_norm
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
params <- list(norm_method = "loess_norm", project_title = "Test") 
source("code/reports/chapters/_internal-shared_setup_chunk.R")  # For "Run Cell" functionality
```

## Calibration curve per chemical

The calibration curve of each of the chemicals is presented below.

```{r}
#| purl: true
calib_se <- calib_interm_data[["calcurve_se within range"]]

# Figure dimensions
n_chem <- nrow(calib_se)
n_cols <- 2    # Number of columns in the plot (facets)
fig_dim <- c(
  # ## 3 columns
  # if (n_chem >= 3) 12 else n_chem * 4,  # Width
  # ((n_chem - 1) %/% 3 + 1) * 3.6 + 0.5       # Height
  ## 1 column
  9,  # Width
  n_chem * 2 + 1    # Height
)
out_width <- paste0(round(fig_dim[1] / 12 * 100), "%")

calcurve_to_show_df <- local({
  cc_se <- calib_interm_data[["calcurve_se all range"]]
  SumExp::as_tibble(cc_se) |> 
    tidyr::drop_na(all_of(DATA_ID)) |>         # Drop concentrations outside the range
    dplyr::rename(.cc_pt = util$spiked_conc_pts_name) |>    # Rename for clarity
    dplyr::mutate(
      "c_type" = dplyr::case_when(
        .cc_pt >= min_c_conc & .cc_pt <= max_c_conc ~ "Within limits",
        TRUE ~ "Out-of-range"
      )
    )
})
cal_data_id <- util$mat_id_before_blank_subtraction(DATA_ID)
txt <- show$label_if_has(calcurve_to_show_df[[cal_data_id]], cal_data_id) |>
  sub("normalized", "normalization", x = _, fixed = TRUE)
```

The peak areas were normalized by the internal standards applying the `r txt`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#calibration-curve-limits).

```{r}
#| fig-width: 10
#| fig-height: 1
#| out-width: 100%
# Plot calibration curves
calcurve_plot <- function(dfm, y = cal_data_id) {
  dfm <- dfm |>
    dplyr::mutate(txt = paste("inj:", injection_order))    # Label for the points
  ggplot(dfm, aes(x = .cc_pt, y = .data[[y]])) +
    suppressWarnings(geom_point(aes(color = c_type, text = txt))) +
    labs(
      x = show$label_if_has(dfm$.cc_pt, "Calibrant Concentration"),
      y = show$label_if_has(dfm[[y]], "Signal"),
      color = "Concentration\ncategory"
    ) +
    scale_color_manual(
      values = c(
        "Within limits" = "green4",
        "Out-of-range" = "gray"
      )
    ) +
    facet_wrap(~ feature_name, scales = "free", ncol = n_cols)
}
p <- calcurve_plot(calcurve_to_show_df)
leg <- ggpubr::get_legend(p, position = "top")
ggpubr::as_ggplot(leg)
```

```{r}
#| label: 'calcurve-all-data-pts'
#| fig-width: !expr fig_dim[1]
#| fig-height: !expr fig_dim[2]
#| out-width: !expr out_width
plotly::ggplotly(
  p + theme(legend.position = "none"),
  tooltip = c("text")
) |> 
  plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 50))
```

## Calibration of the chemicals

```{r}
#| purl: true
concn_se <- calib_interm_data[["concn_se with conc"]]
```

### Calibration models

```{r}
weight_method <- calib_interm_data[["weight method"]]
if (weight_method == "largestR2") {
  n_models <- 6
  weight_txt <- "three different weights, $1$, $\\frac{1}{x}$ and $\\frac{1}{x^2}$"
} else {
  n_models <- 2
  weight_txt <- switch(
    weight_method,
    "1" = "constant",
    "1/x" = "$\\frac{1}{x}$",
    "1/x2" = "$\\frac{1}{x^2}$"
  )
  weight_txt <- paste(weight_txt, "weight")
}
```

The calibration curve is fitted to the `r switch(params$norm_method, "loess_norm" = "LOESS", "closest_norm" = "Closest RT")` normalized data using `r n_models` different models.
Linear models and quadratic models are fitted to the data with `r weight_txt`.
The best model is selected based on the $R^2$ value.
Preferring simpler models, the measures of explained variation of quadratic models were penalized by 0.01.

The distribution of $R^2$ values after penalty by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s_adj <- SumExp::row_df(concn_se)$calcurve_model |> 
  lapply(\(m) m$R2s_adj)
R2s_adj <- R2s_adj[!is.na(R2s_adj)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s_adj) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(paste(R^2, " with penalty")))
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chemical'
#| results: 'markup'
SumExp::row_df(concn_se)$calcurve_model |> 
  sapply(\(m) m$best_model_name) |>
  table()
```

### Calibration curves

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
After the computation, the following steps are applied to the data:

* Any negative concentration is adjusted to zero.
* Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
* Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
* Any chemical with values below LLOQ in all samples is removed from the data.

#### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

```{r}
#| fig-width: 10
#| fig-height: 1
#| out-width: 100%
p <- show$ggplot_calcurve_samples_facet(concn_se, calib_se, DATA_ID, 
                                        COLORS_OF_CLASSES, ncol = n_cols)
leg <- ggpubr::get_legend(p, position = "top")
ggpubr::as_ggplot(leg)
```

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: !expr fig_dim[1]
#| fig-height: !expr fig_dim[2]
#| out-width: !expr out_width
plotly::ggplotly(
  p + theme(legend.position = "none"),
  tooltip = c("text")
) |> 
  plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 50))
```

#### All except `QC` samples

```{r}
classes_excl_qc_blank <- local(envir = list(se = concn_se), {
  not_qc <- util$ctrl_smpl_cat(se) != "QC"
  SumExp::col_df(se)[["Class"]][not_qc] |> 
    unique()
})
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

```{r}
#| fig-width: 10
#| fig-height: 1
#| out-width: 100%
to_show_excl_qc <- concn_se[, quote(Class %in% classes_excl_qc_blank)]
p <- show$ggplot_calcurve_samples_facet(to_show_excl_qc, calib_se, DATA_ID, 
                                        COLORS_OF_CLASSES, ncol = n_cols) +
  show$geom_best_model_text(to_show_excl_qc)
leg <- ggpubr::get_legend(p, position = "top")
ggpubr::as_ggplot(leg)
```

```{r}
#| label: 'calibration-curve-with-quant-sample'
#| fig-width: !expr fig_dim[1]
#| fig-height: !expr fig_dim[2]
#| out-width: !expr out_width
p + theme(legend.position = "none")
# plotly::ggplotly(
#   p + theme(legend.position = "none"),
#   tooltip = c("text")
# ) |> 
#   plotly::layout(margin = list(t = 50, b = 100, l = 50, r = 50))
```

## Removal of the chemicals with no measurement samples

```{r}
concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
```

If all values of non-QC samples were below LLOQ, the chemical was removed from the final reports.
After the removal, the number of reported chemicals is `r nrow(concn_se)`.
