---
title: Calibration - curves
params:
  norm_method: loess_norm
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
```

```{r}
#| purl: true
set.seed(7579)             # To get the same results in a separate child report run
```

## Calibration curve per chemical

The calibration curve of each of randomly selected 9 chemicals is presented below.

```{r}
#| purl: true
cc_se <- calib_interm_data[["calcurve_se within range"]]
# Randomly selected 9 chemicals to present in calibration curve plots
i_9chemical_to_show <- local({
  ii <- c(1:nrow(cc_se))
  ii <- ii[!util$is_internal_std(cc_se)]
  sample(ii, 9)
})
calcurve9_se <- cc_se[i_9chemical_to_show, ]

calcurve_to_show_df <- local({
  cc9_se <- calib_interm_data[["calcurve_se all range"]][i_9chemical_to_show, ]
  SumExp::as_tibble(cc9_se) |> 
    tidyr::drop_na(all_of(DATA_ID)) |>         # Drop concentrations outside the range
    dplyr::rename(.cc_pt = util$spiked_conc_pts_name) |>    # Rename for clarity
    dplyr::mutate(
      "c_type" = dplyr::case_when(
        .cc_pt == non_zero_conc ~ "First non-zero",
        .cc_pt >= min_c_conc & .cc_pt <= max_c_conc ~ "Within limits",
        TRUE ~ "Out-of-range"
      )
    )
})
```

The peak areas were normalized by the internal standards applying the `r show$label_if_has(calcurve_to_show_df[[DATA_ID]], DATA_ID)`.
The concentration range was narrowed down to the range after the trimming step described in the [previous section](#calibration-curve-limits).

- **Full range**

```{r}
#| label: 'calcurve-all-data-pts'
#| fig-width: 12
#| fig-height: 11
# Plot calibration curves
calcurve_plot <- function(df, y = DATA_ID) {
  ggplot(df, aes(x = .cc_pt, y = .data[[y]])) +
    geom_point(aes(color = c_type)) +
    labs(
      x = show$label_if_has(df$.cc_pt, "Calibrant Concentration"),
      y = show$label_if_has(df[[y]], "Signal"),
      color = "Concentration\ncategory"
    ) +
    scale_color_manual(
      values = c("First non-zero" = "red3", 
                 "Within limits" = "green4",
                 "Out-of-range" = "gray")
    ) +
    facet_wrap(~ feature_name, scales = "free", ncol = 3)
}
calcurve_plot(calcurve_to_show_df)
```

- **Zoom-in (upto minimum $\times$ 5)**

```{r}
#| label: 'calcurve-all-data-pts-zoom-in'
#| fig-width: 12
#| fig-height: 11
calcurve_to_show_df |> 
  dplyr::filter(.cc_pt <= min_c_conc * 5) |> 
  calcurve_plot()
```


## Calibration of the chemicals

```{r}
#| purl: true
concn_se <- calib_interm_data[["concn_se with conc"]]
```

### Calibration models

```{r}
weight_method <- calib_interm_data[["weight method"]]
if (weight_method == "largestR2") {
  n_models <- 6
  weight_txt <- "three different weights, $1$, $\\frac{1}{x}$ and $\\frac{1}{x^2}$"
} else {
  n_models <- 2
  weight_txt <- switch(
    weight_method,
    "1" = "constant",
    "1/x" = "$\\frac{1}{x}$",
    "1/x2" = "$\\frac{1}{x^2}$"
  )
  weight_txt <- paste(weight_txt, "weight")
}

```

The calibration curve is fitted to the `r switch(params$norm_method, "loess_norm" = "LOESS", "closest_norm" = "Closest RT")` normalized data using `r n_models` different models.
Linear models and quadratic models are fitted to the data with `r weight_txt`.
The best model is selected based on the $R^2$ value.
Preferring simpler models, the measures of explained variation of quadratic models were penalized by 0.01.

The distribution of $R^2$ values after penalty by different models for the calibration curves is compared below.

```{r}
#| label: 'calibration-R2-by-mmodels'
R2s_adj <- SumExp::row_df(concn_se)$calcurve_model |> 
  lapply(\(m) m$R2s_adj)
R2s_adj <- R2s_adj[!is.na(R2s_adj)] |> 
  dplyr::bind_rows(.id = "feature_id") |> 
  tidyr::pivot_longer(cols = -feature_id, names_to = "Model", values_to = "R2")

ggplot(R2s_adj) +
  geom_boxplot(aes(x = Model, y = R2)) +
  labs(y = expression(paste(R^2, " with penalty")))
```

The count of chemicals favoring each model is shown below.

```{r}
#| label: 'present-model-count-by-chemical'
#| results: 'markup'
SumExp::row_df(concn_se)$calcurve_model |> 
  sapply(\(m) m$best_model_name) |>
  table()
```

### Calibration curves - examples

```{r}
#| purl: true
# Table of the samples to show with the calibration curve
to_show_se <- concn_se[i_9chemical_to_show, ]
```

The concentration of the sample classes, `r knitr::combine_words(unique(SumExp::col_df(concn_se)$Class), before = "\x60")`,
are computed using the calibration curve.
After the computation, the following steps are applied to the data:

* Any negative concentration is adjusted to zero.
* Any concentration below the lower limit of quantification (LLOQ) is adjusted to the half of LLOQ.
* Any concentration below the limit of detection (LOD) is adjusted to the 1/4 of LLOQ.
* Any chemical with values below LLOQ in all samples is removed from the data.

#### All measurement samples

The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-samples'
#| fig-width: 12
#| fig-height: 11
p <- show$ggplot_calcurve_samples_facet(to_show_se,
                                        calcurve9_se,
                                        DATA_ID,
                                        COLORS_OF_CLASSES) +
  labs(y = "Peak area (Normalized)")
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-samples-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

#### All except `QC` samples

```{r}
classes_excl_qc_blank <- local(envir = list(se = concn_se), {
  not_qc <- util$ctrl_smpl_cat(se) != "QC"
  SumExp::col_df(se)[["Class"]][not_qc] |> 
    unique()
})
```

The data of those `r english::words(length(classes_excl_qc_blank))` classes are presented on the calibration curve.

The best model name is displayed at the bottom right corner of each plot.
The shaded gray area represents the region below the lower-limit of quantification (LLOQ) for each chemical.

- **Full range**

```{r}
#| label: 'calibration-curve-with-quant-sample'
#| fig-width: 12
#| fig-height: 11

to_show_excl_qc <- to_show_se[, quote(Class %in% classes_excl_qc_blank)]
p <- show$ggplot_calcurve_samples_facet(to_show_excl_qc, 
                                        calcurve9_se, DATA_ID, COLORS_OF_CLASSES) +
  show$geom_best_model_text(to_show_excl_qc)
print(p)
```

- **Zoom-in (3 calibration points)**

```{r}
#| label: 'calibration-curve-with-quant-sample-zoom-in'
#| fig-width: 12
#| fig-height: 11
p + show$coord_zoom_in_by_cc(n_cc = 3, mat_id = DATA_ID)
```

## Removal of the chemicals with no measurement samples

```{r}
concn_se <- readRDS(FILE$i$proc)[[DATA_ID]]
```

If all values of non-QC samples were below LLOQ, the chemical was removed from the final reports.
After the removal, the number of reported chemicals is `r nrow(concn_se)`.

