---
title: Normalization by closest internal standard in retention time
params:
  norm_method: "loess_norm"        # No need here but needed by `_internal-shared_setup_chunk.R`
---

```{r}
#| purl: true
#| file: "code/reports/chapters/_internal-shared_setup_chunk.R"
#| eval: !expr '!knitr::opts_knit$get("child")'
# The code in this chunk is NOT evaluated here when knitting this file directly.
# It is included for DEBUGGING purposes only. 
params <- list(norm_method = "loess_norm") 
source("code/reports/chapters/_internal-shared_setup_chunk.R")
```

```{r}
#| label: 'loaded-varibles-before-norm-closest-istd'
#| purl: true
#| eval: !expr '!knitr::opts_knit$get("child")'
# Feature intensity to be used for normalization
norm_se <- to_report[["normalized"]]
```

The closest internal standard features to each feature will be identified based on the retention time.
The measured peak area of each feature will be divided by the peak area of the closest internal standard feature.

When a compound target class is specified, the closest internal standard will be selected from the same compound target class.

#### Evaluation of normalization

Normalization is evaluated by comparing the relative standard deviation (RSD) of the peak areas of quantification standards before and after normalization.
Quality control (QC) samples are used for this evaluation because they are technical replicates and should have low RSD values.

```{r}
#| label: 'evaluation-normalization-closest-istd'
#| out.width: '80%'

# Quantification standard samples
qc_se <- show$extract_quant_qc(norm_se)

mat_ids <- c("before_norm", "closest_norm")
rsd_qstd <- show$calc_rsd_qstd(qc_se, mat_ids)
rsd_qstd <- rsd_qstd |> 
  dplyr::mutate(
    sign = .data[["before_norm"]] > closest_norm,
    sign = ifelse(sign, "↓", "↑"),
    sign = ifelse(is.na(sign), "All 0s", sign),
    sign = factor(sign, levels = c("↓", "↑", "All 0s"))
  ) 
show$ggplot_rsdp_metab(rsd_qstd, mat_ids) +
  aes(color = sign) +
  scale_color_manual(values = c("#00BFC4", "#F8766D", "#7f7f7f")) +
  facet_grid(~ QC)
```

The number of features with higher/lower RSD after normalization is counted below.
```{r}
#| label: 'evaluation-normalization-closest-istd-count'
#| results: 'markup'
rsd_qstd |> 
  janitor::tabyl(QC, sign)
```

#### Samples by injection order

```{r}
#| purl: true
is_vIS <- util$std_type(norm_se) == "vIS"
mat_ids_to_compare <- c("raw", "closest_norm")
```

Bar charts of the values of internal standard (`IS`) features are compared for raw and normalized data.
The size of the bars represent the peak area before and after normalization. 
Please note the injection orders may have gaps, as they are obtained information from the source file.

```{r}
#| label: 'injection-order-closest-istd'
#| fig-width: 20
#| fig-height: 6
if (any(is_vIS)) {
  cat("\n\nThe bars represent the peak areas of volumetric internal standards (vIS).\n\n")
  df_show <- norm_se[is_vIS, ] |> 
    show$df_for_injection_order(mat_ids_to_compare)
  plotly::ggplotly(
    show$ggplot_col_injection_order(df_show, fill = feature_name) +
      labs(title = "Volumetric internal standards") +
      # Skip the legend. Too long feature names make the plots too narrow.
      theme(legend.position = "none")
  )
}
```

The bars are colored by the given `Class` of the samples

```{r}
#| fig-width: 20
#| fig-height: 6
df_show <- norm_se[util$is_internal_std(norm_se), ] |> 
  show$df_for_injection_order(mat_ids_to_compare) |>
  dplyr::mutate(
    batch_id = factor(batch_id) |>         # Discrete color scale
      labelled::copy_labels_from(batch_id)
  )

plotly::ggplotly(
  show$ggplot_col_injection_order(df_show, fill = Class) +
    labs(title = "Internal standards") +
    scale_fill_manual(values = COLORS_OF_CLASSES)
)
```

The bars are colored by the given `Batch ID` of the samples

```{r}
#| fig-width: 20
#| fig-height: 6
plotly::ggplotly(
  show$ggplot_col_injection_order(df_show, fill = batch_id) +
    labs(title = "Internal standards") 
)
```

The bars are colored by the internal standards and stacked for each sample.

```{r}
#| fig-width: 20
#| fig-height: 6
plotly::ggplotly(
  show$ggplot_col_injection_order(df_show, fill = feature_name) +
    labs(title = "Internal standards") +
    # Skip the legend. Too long feature names make the plots too narrow.
    theme(legend.position = "none")
)
```

#### Principal component analysis

Principal component analysis (PCA) is performed to evaluate the normalization.
Here the PCA is performed on the internal standard features only.

```{r}
#| label: 'pca-normalization-closest-istd'
#| fig-width: 14
#| fig-height: 6
# PCA
box::use(patchwork)           # Combine ggplot objects. `+`

se <- norm_se[util$is_internal_std(norm_se), ]
# Plot
p_lst <- lapply(mat_ids_to_compare, \(.x) {
  out <- show$ggplot_pca(se, .x, colour = "Class")
  if (is.null(out)) return(NULL)
  out + scale_color_manual(values = COLORS_OF_CLASSES)
})
# Requires patchwork package
Reduce("+", p_lst) + patchwork::plot_layout(ncol = length(p_lst))
```

